# Annotating figures and formatting axes

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

Every hvtiPlotR [@R-hvtiPlotR] plot renders as a bare `ggplot` object: no labels, annotations,
or cropping are applied by the constructor. You add those by chaining `+`
layers. This chapter covers axis/title text (`labs()`), in-panel annotations
(`annotate()`), viewport cropping (`coord_cartesian()`), and draft footnotes
(`make_footnote()`).

## When to use it

A bare plot shows the data correctly but says nothing about what the data mean.
The decorations in this chapter are how you turn a correct plot into one a
reader can interpret without your narration. You add labels so the axes name
real quantities, an annotation so a sample size or a landmark is on the panel
rather than in the caption, a crop so the eye lands on the region that carries
the story, and, while the work is still in progress, a draft footnote so nobody
mistakes a working figure for a final one.

Think of it as the last pass before a figure leaves your hands. The plot is
right; now make it legible. Each tool below is small, and you reach for them in
combination: a labelled, cropped curve with an `n =` callout is the everyday
case. We build one base Kaplan-Meier plot and decorate it throughout, so you can
see each layer added to the same starting point.

```{r}
#| label: base
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)
```

## Labels with `labs()`

`labs()` sets axis labels, title, subtitle, legend title, and caption. Set them
on the rendered plot so they can be overridden per project.

```{r}
#| label: fig-annotation
#| fig-cap: "Overall survival curve with axis labels, percent-formatted y-axis, and a statistical caption applied via labs()"
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title   = "Overall Survival",
    x       = "Years after Operation",
    y       = "Freedom from Death (%)",
    caption = "Logit CI, α = 0.6827 (1 SD)"
  ) +
  theme_hv_manuscript()
```

## Annotations with `annotate()`

`annotate()` places text, segments, or arrows at fixed data coordinates —
a sample-size callout in a corner, or an arrow pointing to an event.

```{r}
#| label: fig-annotation-callout
#| fig-cap: "Survival curve with an in-panel sample-size callout and an arrow annotation marking median survival"
plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  annotate("text",    x = 1,  y = 5,
           label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  annotate("segment", x = 10, xend = 10, y = 30, yend = 50,
           arrow = arrow(length = unit(0.2, "cm")), colour = "grey40") +
  annotate("text",    x = 10.3, y = 40,
           label = "Median survival", hjust = 0, size = 3, colour = "grey40") +
  theme_hv_manuscript()
```

## Naming series without a legend {#sec-annotation-series}

A CORR figure carries no legend. Not in a manuscript, not on a slide.
`theme_hv_manuscript()`, `theme_hv_poster()`, `theme_hv_ppt_light()` and
`theme_hv_ppt_dark()` all set `legend.position = "none"`, so the groups get
named on the panel instead, with `annotate()`, where the reader's eye already
is. A key sends that eye out to the margin and back once per curve. On a slide
that is most of the fifteen seconds you had.

Two things make a direct label read as part of the figure rather than as one
more object sitting on it.

The first is placement. Put the label beside the series it names, at a point
where that series is clearly on its own. The four groups below fan out at the
left edge and converge later, so the left edge is where the labels go.

The second is colour, and it is the one people get wrong. The label takes **the
theme's ink**, the same colour the theme already gives the axis text and the
axis line. White on a dark slide. Black on a light slide and in a manuscript.
Not the colour of the series it names.

```{r}
#| label: fig-annotation-series
#| fig-cap: "Four series named on the panel in the theme's white ink, no legend, previewed against a dark slide background"
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_trends   <- plot(hv_trends(dta_trends))

p_trends +
  hv_ppt_series(mode = "dark", base_size = 18) +
  labs(x = "Surgery Year", y = "Outcome") +
  annotate("text",
           x     = 1990,
           y     = c(64, 51.5, 41.5, 30.5),
           label = c("Group I", "Group II", "Group III", "Group IV"),
           colour = "white", size = 4.5, hjust = 0) +
  theme(plot.background = element_rect(fill = "grey15", colour = NA))
```

`annotate()` takes vectors, so the four labels are one call rather than four.
The single `colour = "white"` recycles across all of them, which is the whole
point: one ink for every label on the panel.

Now the same figure for a journal page. The labels do not move and their text
does not change. Only the ink does, following the theme from white to black.

```{r}
#| label: fig-annotation-series-manuscript
#| fig-cap: "The same four labels for a journal page, where the theme's ink is black and the annotation follows it"
p_trends +
  scale_colour_manual(values = hv_ppt_palette("light")) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Surgery Year", y = "Outcome") +
  annotate("text",
           x     = 1990,
           y     = c(64, 51.5, 41.5, 30.5),
           label = c("Group I", "Group II", "Group III", "Group IV"),
           colour = "black", size = 3.5, hjust = 0) +
  theme_hv_manuscript()
```

Neither scale needs `guide = "none"` here. The theme has already switched the
legend off, and that is the house default rather than something you opt into per
figure.

Colouring each label to match its curve looks like the helpful thing to do, so
it is worth saying plainly why it is not. A coloured label reads as data. It
becomes one more series in the panel, and it inherits every weakness the colour
already has: it flattens on a projector, and it collapses to a mid-grey in a
black-and-white printout, which is exactly the case where the label was doing
the most work. Annotation is frame, not data, so it takes the frame's colour.

Which also means `hv_ppt_palette()` is not where you get an annotation colour,
even though it is close at hand and holds the series colours. The
[colour chapter](colors.qmd) covers what it is for, and
[slides and presentations](presentations.qmd) has the deck-wide decorator that
assigns those colours in the first place.

## Cropping with `coord_cartesian()`

`coord_cartesian()` crops the viewport without dropping data, so any fit
computed on the full range is preserved. This is the distinction that trips up
nearly everyone at some point. `coord_cartesian(xlim = ...)` is a camera: it
zooms in on a region while every point and every fitted line still knows about
the data outside the frame. `scale_x_continuous(limits = ...)` is a filter: it
removes the out-of-range data before anything is drawn, so a LOESS smooth or a
regression line is now fitted to a *subset* and can shift, sometimes
dramatically. When you only want to crop the view, reach for `coord_cartesian()`
every time.

```{r}
#| label: fig-annotation-crop
#| fig-cap: "Trends plot cropped to a sub-range with coord_cartesian(), which zooms the viewport without dropping data"
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))

p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  coord_cartesian(xlim = c(1995, 2020), ylim = c(20, 70)) +
  theme_hv_manuscript()
```

## Draft footnotes with `make_footnote()`

`make_footnote()` stamps the **bottom-right corner** of the current graphics
device using grid. Call it *after* `print()`ing a plot to flag a
work-in-progress figure. The ggplot object itself is untouched, so the final
`ggsave()` output is clean.

```
# During analysis                   # For publication
print(p)                             ggsave("fig1.pdf", p, ...)
make_footnote("R/analysis.R")        # <- no footnote call
```

Build any figure, then `print()` it and call `make_footnote()` with the source
path. The default stamp appends a timestamp.

```{r}
#| label: fig-annotation-footnote
#| fig-cap: "Survival curve stamped with a draft footnote naming the source file and a timestamp"
#| fig-width: 7
#| fig-height: 5
p_draft <- plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Survival (%)") +
  theme_hv_manuscript()

print(p_draft)
make_footnote("annotation.qmd")
```

Pass any string as `text` and set `timestamp = FALSE` for a stable analyst tag.

```{r}
#| label: footnote_analyst
#| fig-width: 7
#| fig-height: 5
print(p_draft)
make_footnote(
  text      = paste("J. Ehrlinger |", Sys.Date()),
  timestamp = FALSE,
  prefix    = ""
)
```

For publication, write the ggplot object directly with `ggsave()` and simply
omit the `make_footnote()` call. The saved file carries no stamp.

## Pitfalls

- **`coord_cartesian()` zooms, `scale_*(limits =)` drops.** This is the one to
  memorise. If you set limits on the scale to "zoom in," you have silently
  thrown away the data outside the window, and any smoother, ribbon, or summary
  refits to what is left. The curve you see is then a fit to a subset, not the
  region of a fit to everything. Use `coord_cartesian()` when you want the same
  fit, viewed closer. Use a scale limit only when you genuinely mean to exclude
  data, and say so.
- **An annotation coloured to match the series it names.** The label then
  reads as one more piece of data rather than as part of the figure's frame, and
  it fails in every place the series colour fails: flattened by a projector,
  collapsed to grey by a black-and-white printer. Take the theme's ink instead,
  white on a dark slide and black everywhere else, so every label on the panel
  matches the axis and each other.
- **A draft footnote left on a final figure.** `make_footnote()` stamps the
  device, not the ggplot object, which is the safe design: the stamp cannot
  survive into a `ggsave()` file unless you deliberately keep printing and
  stamping. The failure mode is human, not technical. You screenshot a working
  figure mid-analysis, the file path and timestamp ride along, and it ends up in
  a slide deck. Keep draft stamps to the analysis session and let the publication
  path be the plain `ggsave()`.
