# Bar chart plots

How many observations fall in each category, and which denominator makes the
comparison honest? Bar charts answer the count question directly and switch to
proportions when cohort size would otherwise dominate the story.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
library(patchwork)
```

## When to use it

A bar chart answers a counting question. How many patients had a concomitant
CABG this year? What fraction arrived in NYHA class III or IV? How did the size
of each annual cohort change over the study period? Whenever the thing you want
to show is a count or a proportion of a category, and you want it broken out
across a second axis (usually time), the bar chart is the plain, honest display
to reach for. The geometry is familiar, but the denominator still needs saying.

There are two related jobs in this chapter, and the package splits them across
two functions. `hv_eda()` [@R-hvtiPlotR] is the exploratory workhorse you point
at a single variable to see how its categories fall out across a reference axis;
it is the same function that draws scatter plots for continuous columns, but
here we stay with the categorical case where it returns bars. `hv_longitudinal()`
is the reporting display: a grouped bar chart of how many patients and how many
measurements you have at each follow-up window, the figure you put at the front
of a longitudinal results section so the reader knows how much data backs the
later curves.

Both hand back a bare ggplot (or, for the longitudinal table, a panel you compose
with `patchwork`). You dress them with colour scales, labels, and a house theme
using the usual `+`.

## The data it needs

`hv_eda()` expects one row per observation with a reference axis in `x_col`
(here the surgery `year`) and the variable of interest in `y_col`. It does not
ask you to tell it whether that variable is categorical: `eda_classify_var()`
inspects the column and decides. Numeric 0/1 and small-integer columns come back
as `"Cat_Num"`, character columns as `"Cat_Char"`, and continuous columns as
`"Cont"`. The classifier is worth running on the whole frame before you build
anything, so you know which columns will draw as bars and which as a scatter.

```{r}
#| label: eda_data
dta_eda <- sample_eda_data(n = 300, seed = 42)
head(dta_eda)

# Auto-detected types for each column
sapply(dta_eda, eda_classify_var)
```

`hv_longitudinal()` is fussier: it expects **pre-aggregated** long-format data,
one row per time-window by series combination, with a count column. You do not
hand it raw patient records. `sample_longitudinal_counts_data()` derives that
shape from a simulated 300-patient registry so the recipe runs end to end.

## Inspect it

Build the `hv_longitudinal` S3 object once and inspect its class and metadata
before plotting. The metadata records the count, interval, and series columns
that the plot method will use.

```{r}
#| label: lc_data
lc_dat <- sample_longitudinal_counts_data(n_patients = 300, seed = 42L)
lc_dat

# Build the S3 object once; use for both panels
lc <- hv_longitudinal(lc_dat)
class(lc)
lc$meta
```

## Build it

Start from the bare panel for `hv_longitudinal()` so you can see what the
constructor produced before any styling. Reuse the inspected object across both
panels below.

```{r}
#| label: lc_bare
plot(lc)
```

That is the raw grouped bar chart: paired bars at each follow-up window, no fill
colours, no axis range, no theme. Now layer on the house style. We give the two
series distinct fills, expand the y-axis with `coord_cartesian()` so the tallest
bar has headroom, and let `hv_legend_inside()` drop the legend into the empty
upper-left corner where the short early-window bars leave room.

```{r}
#| label: fig-bar
#| fig-cap: "Grouped bar chart of patient and measurement counts at each follow-up window"
p_lc_bar <- plot(lc) +
  ggplot2::scale_fill_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    name   = NULL
  ) +
  ggplot2::scale_y_continuous(
    breaks = seq(0, 700, 100),
    expand = c(0, 0)
  ) +
  ggplot2::coord_cartesian(ylim = c(0, 700)) +
  ggplot2::labs(x = NULL, y = "Count (n)") +
  theme_hv_manuscript() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

hv_legend_inside(p_lc_bar)
```

## Read it

A grouped count bar chart is read one window at a time. Follow-up window is on
the x-axis and raw count is on the y-axis. Within each window, compare unique
patients with all measurements contributed by those patients; these are two
different denominators, not percentages of a shared total. The bars describe
data availability and cannot establish why follow-up thinned or whether an
outcome changed. A few things to look for:

- **Measurements should always be at least Patients.** Each patient contributes
  one or more measurements, so the Measurements bar should never sit below the
  Patients bar in the same window. If it does, either the series are swapped or
  the count construction needs checking.
- **The shape across windows.** These are disjoint follow-up intervals, not a
  cumulative at-risk count, so the bars do not have to decline from left to
  right. A thin interval identifies a part of follow-up with little information;
  check its counts before trusting a curve there.
- **The gap between the two bars.** A wide gap means some patients contributed
  more than one measurement within that interval. A narrow gap means most
  represented patients contributed one measurement.

## Adapt it

The remaining variants come from `hv_eda()`, which draws the single-variable
exploratory bars.

### Binary categorical: count bars

Numeric 0/1 columns are classified as `"Cat_Num"`, and `NA` values appear as an
explicit `"(Missing)"` fill level rather than vanishing, so you can colour and
count them. The `y_label` argument sets the title and fill-legend name in place
of the raw column name. Here, sex by surgery year.

```{r}
#| label: fig-bar-binary-count
#| fig-cap: "Count bars of a binary categorical variable by surgery year, with missing values kept as an explicit level"
eda_male <- hv_eda(dta_eda, x_col = "year", y_col = "male",
                   y_label = "Sex")
class(eda_male)
eda_male$meta

plot(eda_male) +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick", "(Missing)" = "grey80"),
    labels = c("0" = "Female", "1" = "Male", "(Missing)" = "Missing"),
    name   = NULL
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

The x-axis is surgery year and the y-axis is the number of records in that
year. Each bar is partitioned into female, male, and missing records, so the
annual record count is the denominator behind the comparison. Differences in
the bars describe case mix and volume; they do not show that calendar time
caused the mix to change.

### Binary categorical: percentage bars

When you care about the *mix* rather than the absolute volume, set
`show_percent = TRUE`. That switches `geom_bar()` to `position = "fill"`, so
every year's bar runs the full height and the fill shows the proportion in each
category. This is the version to use when annual cohort sizes vary a lot and a
raw count would tell you more about volume than about case mix.

```{r}
#| label: fig-bar-binary-percent
#| fig-cap: "Percentage bars of a binary categorical variable by year, with each bar filling the full height to show case mix"
plot(hv_eda(dta_eda, x_col = "year", y_col = "cabg",
            y_label = "Concomitant CABG", show_percent = TRUE)) +
  scale_fill_manual(
    values = c("0" = "grey70", "1" = "steelblue", "(Missing)" = "grey90"),
    labels = c("0" = "No CABG", "1" = "CABG", "(Missing)" = "Missing"),
    name   = NULL
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Surgery Year", y = "Proportion") +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

Here each year's full bar is the denominator and the segments sum to 100%.
Compare shares rather than heights, and remember that equal-height bars can sit
on very different numbers of patients. The display does not estimate a CABG
effect or explain why practice changed.

### Ordinal and multi-level categorical

Columns with more than two numeric levels render as stacked count bars, one
level per fill colour. For an ordered grade such as NYHA class, a reversed
diverging palette (`"RdYlGn"`) carries the severity ordering visually, from
green at the mild end through yellow to red at the severe end, so the reader
sees worsening case mix without reading the legend.

```{r}
#| label: fig-bar-ordinal
#| fig-cap: "Stacked count bars of an ordinal NYHA class by year, with a diverging palette carrying the severity ordering"
plot(hv_eda(dta_eda, x_col = "year", y_col = "nyha",
            y_label = "Preoperative NYHA Class")) +
  scale_fill_brewer(
    palette = "RdYlGn", direction = -1,
    labels  = c("1" = "I", "2" = "II", "3" = "III", "4" = "IV",
                "(Missing)" = "Missing"),
    name    = "NYHA"
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

### Character categorical

String columns are classified as `"Cat_Char"` and also produce stacked count
bars. The difference is that character levels are ordered alphabetically by
default rather than by an implied scale, so use `scale_fill_manual()` to assign
colours that carry clinical meaning, here one hue per valve morphology type.

```{r}
#| label: fig-bar-char-cat
#| fig-cap: "Stacked count bars of a character categorical variable by year, with one hue per valve morphology type"
plot(hv_eda(dta_eda, x_col = "year", y_col = "valve_morph",
            y_label = "Valve Morphology")) +
  scale_fill_manual(
    values = c(Bicuspid   = "steelblue",
               Tricuspid  = "firebrick",
               Unicuspid  = "goldenrod3",
               "(Missing)" = "grey80"),
    name = "Morphology"
  ) +
  scale_x_discrete(breaks = seq(2005, 2020, 5)) +
  labs(x = "Surgery Year", y = "Count") +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

### Longitudinal: the numeric table panel

`hv_longitudinal()` carries a second panel. `plot(lc, type = "table")` renders
the same counts as coloured text below the x-axis labels, the numeric companion
to the bars. On its own it is sparse; it is meant to sit under the bar chart.

```{r}
#| label: lc_table
p_lc_tbl <- plot(lc, type = "table") +
  ggplot2::scale_colour_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    guide  = "none"
  ) +
  theme_hv_manuscript() +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title   = element_blank()
  )

p_lc_tbl
```

### Longitudinal: the two-panel composite

Stack the bar chart above the table with `patchwork`'s `/` operator, then use
`plot_layout(heights = c(3, 1))` to give the bars three times the vertical space.
This is the figure you actually publish: the bars carry the visual story and the
table gives the reader the exact counts.

```{r}
#| label: fig-bar-longitudinal-composite
#| fig-cap: "Two-panel longitudinal display with the grouped bar chart above and the matching count table below"
#| fig-height: 5
p_lc_bar / p_lc_tbl +
  patchwork::plot_layout(heights = c(3, 1))
```

## Deliver it

Use [Finish and deliver the result](publications.qmd) once the caption names
whether the bars show counts or within-year percentages. For the longitudinal
display, keep the count table attached to the bars so the clinical reader can
see the exact patient and measurement totals.

## Pitfalls

- **Count versus proportion.** A raw count bar and a `show_percent = TRUE` bar
  answer different questions. If annual cohort sizes swing widely, a count chart
  can make a stable case mix look like a trend, and a percentage chart can hide a
  collapse in volume. Show whichever matches the claim, and say which it is.
- **Dropping the missing level.** `hv_eda()` keeps `NA` as an explicit
  `"(Missing)"` level on purpose. Recolouring it to white or styling it away
  hides how much of each bar is unknown. Leave it visible, or at least account
  for it in the caption.
- **Feeding `hv_longitudinal()` raw records.** It expects pre-aggregated
  one-row-per-window-per-series data. Hand it patient-level rows and the counts
  will be wrong without an error. Aggregate first, or use
  `sample_longitudinal_counts_data()` as the template for the shape.
- **Reading a stacked bar's middle segments.** Only the bottom segment of a
  stacked bar starts from a common baseline; the middle segments float, so their
  heights are hard to compare across years. If a middle category is the story,
  pull it out into its own binary count or percentage bar.
