# Box plots

How do the centre and spread of a continuous measurement compare across
clinical groups? A box plot makes that comparison compact, provided we remember
what its summary leaves out.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

When you want to compare a continuous measurement across a handful of groups,
the box plot is the compact summary that fits in a single panel. It draws the
median, the two quartiles, and the whiskers for each
group side by side, so the reader can scan across categories and see at once
which group sits higher, which is more variable, and which has stragglers. Reach
for it when the question is "does this number differ by group?" and the groups
number in the low single digits. With twenty groups the boxes get too thin to
read; with two groups you might prefer to overlay densities. In the sweet spot of
three to six categories, the box plot is hard to beat.

`hvtiPlotR` does not provide a box-plot helper, so this chapter builds the plot
directly with ggplot2's [@R-ggplot2] `geom_boxplot()` and styles it with the
house theme `theme_hv_manuscript()` so it stays visually consistent with the
rest of the book. As with the other bare-ggplot recipes, you write the
`ggplot()` call yourself and finish with `+ theme_hv_manuscript()`.

## The data it needs

A box plot needs a categorical column for the x-axis and a continuous column for
the y-axis. We use `hvtiPlotR::sample_eda_data()` with `valve_morph` (valve
morphology: Bicuspid, Tricuspid, or Unicuspid) as the group and `ef` (ejection
fraction) as the measurement. Naming both columns directly keeps the example
tied to one clinical question.

```{r}
#| label: data
dat <- sample_eda_data()
num <- "ef"
grp <- "valve_morph"
c(numeric = num, group = grp)
```

The `.data[[grp]]` idiom is how ggplot2 reads a column whose name lives in a
variable, so the same recipe runs after you replace `grp` or `num` with another
column name.

## Inspect it

Because this recipe creates a ggplot directly, inspect the group counts and
numeric summary before collapsing the observations into boxes.

```{r}
#| label: boxplots_inspect
table(dat[[grp]], useNA = "ifany")
summary(dat[[num]])
```

## Build it

The grouped box plot is the base case: one box per level of the categorical
variable, all on a shared y-axis.

```{r}
#| label: fig-boxplots
#| fig-cap: "Box plot of a continuous measurement by group, showing the median, quartiles, whiskers, and outliers for each category"
ggplot(dat, aes(x = .data[[grp]], y = .data[[num]])) +
  geom_boxplot() +
  labs(
    title = "Numeric distribution by group",
    x = "Valve morphology",
    y = "Ejection fraction (%)"
  ) +
  theme_hv_manuscript()
```

Each box packs a lot in. The line inside the box is the median; the box edges
are the first and third quartiles (the 25th and 75th percentiles), so the box
spans the middle half of the data, the interquartile range or IQR. The whiskers
reach out to the most extreme points within 1.5 times the IQR of the box, and
anything past the whiskers is drawn as an individual outlier point.

## Read it

Read a box plot one box at a time, then across. Valve morphology is on the
x-axis and ejection fraction is on the y-axis. Each box summarizes the
nonmissing values in its group, with no weighting or model adjustment. Compare
medians and IQRs, but do not read their separation as a causal morphology effect
or as a formal test of a group difference.

- **The median line.** This is the typical value for the group, and comparing
  median lines across boxes is the first thing your eye should do. A box that
  sits higher has a higher typical value.
- **The box height (IQR).** A tall box means the middle half of patients are
  spread out; a short box means they cluster. This is the spread you compare
  across groups, and it is more stable than the whiskers.
- **The whiskers and outliers.** The whiskers show the reach of the bulk of the
  data, and they extend to 1.5 times the IQR, not to the minimum and maximum.
  Points drawn beyond them are flagged as outliers by that same rule. They are
  candidates for a second look, not automatically errors.
- **Asymmetry.** If the median sits low in the box and the upper whisker is long,
  the group is right-skewed; the box plot shows skew even though it hides the
  finer shape.

What a box plot does not show is the shape inside the box. Two groups can have
identical boxes while one is a single clump and the other is two separated
clusters. That blind spot is the reason for the next variation.

## Adapt it

### Box plot with jittered points

Overlaying the raw observations with `geom_jitter()` shows the underlying sample
size and spread that a box plot alone hides. We turn off the box's own outlier
points (`outlier.shape = NA`) so they are not drawn twice, then scatter the
points with a little horizontal jitter and low alpha so dense regions read as
darker.

```{r}
#| label: fig-boxplots-jitter
#| fig-cap: "Box plot with the raw observations jittered over each box, exposing the sample size and spread the summary hides"
ggplot(dat, aes(x = .data[[grp]], y = .data[[num]])) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 0.7) +
  labs(
    title = "Box plot with jittered observations",
    x = "Valve morphology",
    y = "Ejection fraction (%)"
  ) +
  theme_hv_manuscript()
```

Now you can see how many patients sit behind each box, and whether the points
under a box form one cloud or two. For the small group in this sample, the
handful of points makes plain how little the box is summarising.

### Letting the box width show the group size

A box plot draws every box the same width regardless of how many patients it
summarises, which can mislead. Setting `varwidth = TRUE` makes each box's width
proportional to the square root of its group size, so a thin box is a visual
warning that it rests on few patients.

```{r}
#| label: fig-boxplots-varwidth
#| fig-cap: "Box plot with box width scaled to the square root of group size, flagging boxes that rest on few patients"
ggplot(dat, aes(x = .data[[grp]], y = .data[[num]])) +
  geom_boxplot(varwidth = TRUE) +
  labs(
    title = "Box width scaled to group size",
    x = "Valve morphology",
    y = "Ejection fraction (%)"
  ) +
  theme_hv_manuscript()
```

### Filling the boxes by group

A grey box plot is honest but plain. Mapping `fill` to the group colours each box
from a palette, which makes a busy panel easier to scan and ties the boxes to the
same colours the group uses elsewhere in a figure. Because the x-axis already
names the groups, the fill only reinforces them, so we drop the redundant legend
with `guide = "none"` rather than spend panel space on it.

```{r}
#| label: fig-boxplots-filled
#| fig-cap: "Box plot with each box filled by group from a palette, with the redundant legend dropped"
ggplot(dat, aes(x = .data[[grp]], y = .data[[num]], fill = .data[[grp]])) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  labs(
    title = "Numeric distribution by group, filled",
    x = "Valve morphology",
    y = "Ejection fraction (%)"
  ) +
  theme_hv_manuscript()
```

When the fill encodes a *second* variable the boxes are dodged by, not the
x-axis group, the legend earns its place; keep it then, placed inside the panel
so it does not crowd the boxes.

### The house percentile box

Everything above is a Tukey box, which is what `geom_boxplot()` draws. Our
house box is a different one. Its hinges sit at the 15th and 85th percentiles,
its whiskers run to the 5th and 95th, and it draws no outlier points. The
percentiles match the median (15th, 85th) summary our
[manuscript tables](hv_tables.qmd) report, so a figure and its table describe
a group with the same numbers. It is also the box `gg_varpro()` draws in the
[variable priority](varpro.qmd) chapter, whose plot carries the caption "Not a
Tukey boxplot".

`geom_boxplot()` has no argument for other percentiles, so we compute the five
numbers per group and hand them over with `stat = "identity"`.

```{r}
#| label: fig-boxplots-percentile
#| fig-cap: "House percentile box: the median line, hinges at the 15th and 85th percentiles, and whiskers to the 5th and 95th, with no outlier points"
pct <- do.call(rbind, lapply(split(dat[[num]], dat[[grp]]), function(y) {
  q <- quantile(y, c(0.05, 0.15, 0.50, 0.85, 0.95), na.rm = TRUE)
  data.frame(ymin = q[1], lower = q[2], middle = q[3], upper = q[4], ymax = q[5])
}))
pct$group <- rownames(pct)

ggplot(pct, aes(x = group, ymin = ymin, lower = lower, middle = middle,
                upper = upper, ymax = ymax)) +
  geom_boxplot(stat = "identity") +
  labs(
    title = "Percentile box by group",
    x = "Valve morphology",
    y = "Ejection fraction (%)",
    caption = "Hinges: 15th/85th percentiles; whiskers: 5th/95th. Not a Tukey boxplot."
  ) +
  theme_hv_manuscript()
```

Two things change in how you read it. The box now holds the middle 70% of the
group rather than the middle half, so it is taller than a Tukey box drawn from
the same data. The whiskers are fixed percentiles with no outlier rule behind
them, so 10% of the group lies beyond them by construction, and nothing past
the whiskers is singled out. Say so in the caption every time, as the
`caption` here does, because a reader will assume Tukey otherwise.

## Deliver it

When the group definition and box rule are clear in the caption, continue to
[Finish and deliver the result](publications.qmd) for the manuscript, poster, or
slide export path. If sample size affects the interpretation, deliver the
jittered or variable-width version rather than the bare box alone.

## Pitfalls

- **Hiding the shape.** A box plot collapses the distribution to five numbers,
  so a bimodal group (two clusters with a gap between them) looks the same as a
  single clump centred at the same median. When the shape might matter, overlay
  the points as in the jitter variation, or draw a density alongside.
- **Small-n boxes.** A box computed from a dozen patients has quartiles and
  whiskers that can swing on one or two values, yet it is drawn with the same
  authority as a box from hundreds. Treat a box over a small group with caution:
  scale the width to n, overlay the points, or just report the numbers.
- **Misreading the whiskers.** The whiskers are not the minimum and maximum. They
  reach to the most extreme point within 1.5 times the IQR, and points past them
  are flagged as outliers by that rule. A long whisker does not mean the data run
  exactly that far, and an "outlier" point is a statistical flag, not a verdict
  that the value is wrong.
- **Mixing box rules without saying so.** This book draws two kinds of box: the
  Tukey box from `geom_boxplot()` and the house percentile box. They look the
  same and summarise different things. State the rule in every caption, and
  keep to one rule within a figure.
- **Too many or unordered groups.** Past five or six categories the boxes get too
  narrow to compare; and leaving them in alphabetical order when the variable has
  a natural order (severity, era) makes the trend harder to see than reordering
  the factor would.
