# Color templates for consistency

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

hvtiPlotR plots carry no colour scale by default. ggplot2's [@R-ggplot2]
`scale_colour_*` controls
line and point colours; `scale_fill_*` controls filled areas (ribbons, bars).
Both take the same `name` (legend title) and `guide` (legend display)
arguments. Set them together so the two scales stay in sync.

## When to use it

Colour is the fastest way to separate groups on a figure, and the easiest to
overuse. Reach for a manual colour scale when the groups have a convention you
want to honour (a brand colour, red for the high-risk arm) or when you have one
series and simply want it off the default grey. Reach for a ColorBrewer palette
when the groups are arbitrary and you want a set of hues that stay distinct in
print and for colour-blind readers. And reach for legend suppression whenever
the figure already names its groups some other way, since a colour that needs no
key earns back panel space.

The guiding question is what the colour is *for*. If it carries information, the
distinct group it marks, it has to survive a black-and-white printer and a
colour-blind reader, which is where palette choice and a second visual channel
come in. If it is decoration, keep it to one or two restrained hues. We use a
grouped trends plot below so the colour mapping is visible across levels.

```{r}
#| label: base
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))
```

## Manual colours

Use `scale_colour_manual()` / `scale_fill_manual()` to assign specific brand or
convention colours to known levels. Here a single-stratum KM curve is set to one
colour.

```{r}
#| label: fig-colors-manual
#| fig-cap: "Single-stratum survival curve with a manual colour scale assigning one named colour to the lone level"
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)

plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_manuscript()
```

## ColorBrewer palettes

`scale_colour_brewer()` applies a perceptually uniform, print-safe palette —
which matters when figures go to black-and-white PDF. Use `palette = "Set1"`
for categorical data, `"RdYlGn"` for diverging, `"Blues"` for sequential.

```{r}
#| label: fig-colors
#| fig-cap: "Grouped trends plot coloured by a print-safe ColorBrewer Set1 palette, with a matched point shape per group"
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_manuscript()
```

## The house slide palette

`hv_ppt_palette()` hands back the six colours `hv_ppt_series()` assigns across a
deck, as a plain character vector. Reach for it when a figure needs those
colours outside the decorator, so the script uses the one definition instead of
hex codes pasted out of an older script, which go stale the first time the
palette changes.

```{r}
#| label: ppt-palette
hv_ppt_palette("dark")
hv_ppt_palette("light", n = 4)
```

These are the Okabe-Ito colourblind-safe hues, reordered for the background the
figure will sit on. High-luminance colours come first on a dark slide and the
darker ones first on a light slide, so the leading series always has the
strongest contrast against the surface behind it. Black appears only in the
light ordering, since on a dark panel there is nothing there to see. They are
not CORR brand colours, and nothing stops a project from handing its own set to
`scale_colour_manual()` instead.

Six colours are the entire supply, which covers a grouping variable up to six
levels. Ask for a seventh and you get an error rather than a quietly recycled
palette, on the reasoning that two series sharing a colour is worse than a
stopped script.

Pair the `"light"` ordering with `theme_hv_manuscript()` and a journal figure
takes the same hues the deck does, so one analysis looks like itself in both
places.

```{r}
#| label: fig-colors-ppt-palette
#| fig-cap: "Grouped trends plot taking the light-background ordering of the house palette, with the matching point shapes"
p_base +
  scale_colour_manual(values = hv_ppt_palette("light")) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_manuscript()
```

The shape scale is not decoration here. Okabe-Ito is built to survive colour
blindness, which is not the same thing as surviving a greyscale printer, so the
second channel still has work to do. And the colours are only half the recipe.
The deck-wide decorator that assigns them, and the theme that rides along with
them, are in [slides and presentations](presentations.qmd).

## Suppressing legends

Pass `guide = "none"` to any scale and that aesthetic drops out of the legend.
Do this when axis labels or an annotation already name the group, so a redundant
legend does not eat panel space.

```{r}
#| label: fig-colors-no-legend
#| fig-cap: "The grouped plot with both colour and shape legends suppressed via guide = \"none\", relying on redundant colour and shape channels"
p_base +
  scale_colour_brewer(palette = "Dark2", guide = "none") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    guide = "none"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_manuscript()
```

Notice that @fig-colors-no-legend pairs colour with a distinct point shape per
group. That redundancy is deliberate, and it is the heart of the first pitfall
below.

## Pitfalls

- **Colour as the only channel.** If the only thing separating two groups is
  their hue, the figure fails for a sizeable fraction of readers (red-green
  colour blindness alone affects roughly one man in twelve) and for anyone who
  prints it in greyscale, where two mid-tone colours collapse to the same grey.
  The fix is cheap: give each group a second channel as well, a point shape or a
  line type, so the distinction survives even when the colour does not. The
  recipes here map `scale_shape_manual()` alongside the colour scale for exactly
  this reason. A quick check is to convert the figure to greyscale and confirm
  you can still tell the groups apart.
- **Too many hues.** A palette with eight saturated colours does not help a
  reader; it asks them to memorise a legend and hunt for matches. Past about
  four or five groups, distinct colours stop being distinct and the figure turns
  to confetti. When you have many series, consider showing fewer at once,
  faceting into small panels, or muting the background series in grey and
  colouring only the one or two you are making a point about. Restraint reads as
  confidence.
