# Combination figures

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
library(patchwork)
```

Every recipe so far has produced one finished figure. This chapter is about the
last step many manuscripts need: taking two or three of those figures and
setting them side by side as a single labelled exhibit. Journals ask for it
constantly. "Figure 1. Panel A shows survival; Panel B shows covariate balance"
is one figure with one caption and one number, even though you built each panel
separately. The patchwork package [@R-patchwork] is how we assemble them, and
because the panels stay independent until the final line, you can re-style any
one of them without disturbing the rest.

## When to use it

Compose a combination figure when the panels are stronger together than apart.
That is not the same as "whenever you have more than one plot." A good reason to
combine: two views of the same cohort that the reader should compare at a glance
(the outcome and the balance check that justifies it), or a main plot paired
with a companion strip (a survival curve above its numbers-at-risk table). A bad
reason: cramming unrelated figures together to save a figure number. If a reader
would not look back and forth between the panels, they belong in separate
figures.

The mechanics are deliberately small. patchwork gives you two operators (`|`
puts plots in a row, `/` stacks them in a column), `plot_layout()` to set the
relative sizes, and `plot_annotation()` to add the shared title and the A, B, C
panel tags. Everything else is the ggplot you already know.

## The plots you already built

Where a single-figure recipe needs a data frame, a composition recipe needs the
finished plots. So the "data" here is two hvtiPlotR [@R-hvtiPlotR] plots built
the way the earlier chapters built them: a survival curve and a covariate
balance panel. We assign each to a name (`p_km`, `p_cb`) and do nothing else to
them yet. That naming is the whole trick. Each panel is a self-contained object
we compose later.

```{r}
#| label: combo_panels
km <- hv_survival(sample_survival_data(n = 300, seed = 42))

p_km <- plot(km) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_manuscript()

cb <- hv_balance(sample_covariate_balance_data(n_vars = 12))

p_cb <- plot(cb, alpha = 0.8) +
  scale_color_manual(
    values = c("Before match" = "red4", "After match" = "blue3"),
    name   = NULL
  ) +
  scale_shape_manual(
    values = c("Before match" = 17L, "After match" = 15L),
    name   = NULL
  ) +
  labs(x = "Standardized difference (%)", y = "") +
  theme_hv_manuscript()
```

## Build it

Start with the simplest composition: two panels in a row. The `|` operator
places `p_cb` and `p_km` next to each other, sharing the figure height. We widen
the chunk (`fig-width: 10`) so neither panel is squeezed.

```{r}
#| label: combo_side_by_side
#| fig-width: 10
p_cb | p_km
```

That is the bare composite. It reads, but the two panels split the width evenly
even though the balance plot has long covariate labels that want more room. Most
real combination figures need three more things layered on: a deliberate size
split, panel tags, and a shared title. We add them in the sections below, the
same way you decorate a single plot, one layer at a time.

## Read it

A combination figure asks the reader to do something a single panel does not:
move between panels and connect them. Whether that lands is about the
composition, not the data. Look for:

- **Panel tags that match the caption.** The A, B, C labels are how the caption
  points at a panel ("Panel B shows..."). If the tags are missing or out of
  order, the caption has nothing to anchor to. Set them once and check they read
  left-to-right, top-to-bottom.
- **A size split that follows the content.** When one panel carries long axis
  labels or a tall legend, an even split crushes it. The relative widths should
  give each panel the room its content needs, not a reflexive fifty-fifty.
- **Shared structure doing shared work.** Stacking a plot above its own
  numbers-at-risk strip works because they share an x-axis; the reader's eye
  drops straight down from a curve to its at-risk count. If two stacked panels
  do not share an axis, stacking implies a relationship that is not there.

## Variations

### Controlling relative widths and heights

`plot_layout(widths = ...)` takes a numeric vector of relative widths. `c(2, 1)`
makes the left panel twice as wide as the right, which suits the balance plot's
long labels. `heights` does the same for stacked layouts.

```{r}
#| label: fig-combination-layout
#| fig-cap: "Two-panel composite with the balance panel set twice as wide as the survival panel to fit its long covariate labels"
#| fig-width: 10
(p_cb | p_km) +
  plot_layout(widths = c(2, 1))   # left panel twice as wide as right
```

### Stacking a plot above a risk table

A survival curve reads best with its numbers-at-risk strip directly below, the
two sharing an x-axis. `hv_survival()` stores that table as a data frame at
`km$tables$risk` (columns `strata`, `report_time`, `n.risk`), so we build a thin
ggplot text panel from it and stack with `/`. The `plot_layout(heights = ...)`
keeps the table strip short.

```{r}
#| label: fig-combination-risk-table
#| fig-cap: "Survival curve stacked above its numbers-at-risk strip, the two panels sharing a common x-axis"
#| fig-height: 5
risk_df <- km$tables$risk

rt_panel <- ggplot(risk_df,
                   aes(x = report_time, y = factor(strata),
                       label = n.risk)) +
  geom_text(size = 3) +
  labs(x = "Years after Operation", y = NULL) +
  theme_hv_manuscript() +
  theme(
    axis.line  = element_blank(),
    axis.ticks = element_blank()
  )

p_km / rt_panel +
  plot_layout(heights = c(4, 1))
```

### Shared panel tags and a figure title

`plot_annotation()` adds the shared title and the panel tags most journals
require. `tag_levels = "A"` numbers the panels A, B, C... automatically. The `&`
operator applies a `theme()` to every panel at once, so you format the tags in
one place rather than per panel.

```{r}
#| label: fig-combination
#| fig-cap: "Finished combination figure pairing the covariate balance and survival panels with a shared title and A, B panel tags"
#| fig-width: 10
(p_cb | p_km) +
  plot_annotation(
    title = "Figure 1. Outcomes after cardiac surgery",
    tag_levels = "A"
  ) &
  theme(plot.tag = element_text(size = 12, face = "bold"))
```

## Pitfalls

- **Compose last, style first.** Finish each panel as its own figure before you
  combine. Trying to fix a panel's axes or theme after composition means
  reaching back through the patchwork, which is harder than re-styling the named
  object and re-composing.
- **Even splits crush uneven panels.** The default 50/50 width ignores that one
  panel may carry long labels or a legend. Reach for `plot_layout(widths = ...)`
  whenever a panel looks pinched.
- **Tag once, at the composition.** Set `tag_levels` in `plot_annotation()`, not
  by hand inside each panel. Hand-set tags drift out of order the moment you
  reorder the panels.
- **Stacking implies a shared axis.** Putting one panel directly above another
  tells the reader the x-axes line up. If they do not, either give the panels a
  genuinely shared axis or set them side by side instead.
- **One caption, one story.** A combination figure gets one caption. If you
  cannot write a single sentence that ties the panels together, that is the
  signal they should be separate figures.
