# CONSORT patient-flow diagrams

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

Every outcomes paper has to answer one question before any analysis: how did you
get from the patients you started with to the patients you actually analysed? The
CONSORT diagram is that answer drawn as a flow chart. It is the box-and-arrow
figure near the front of the methods section: a top box of everyone screened, a
side box at each step listing how many were dropped and why, and a bottom box of
the final analysed cohort. A reviewer reads it to check that your exclusions are
defensible and that the numbers add up.

Reach for it whenever you have built a study cohort by applying a sequence of
exclusion rules and you need to account for every patient between the registry
pull and the analysis set. `hv_consort()` [@R-hvtiPlotR] builds the diagram
through a multi-step API: `hv_consort_start()` initializes a patient-level
tracker, `hv_consort_exclude()` adds an exclusion stage, and `hv_consort()`
renders the result.

One thing sets this figure apart from the rest of the book. The diagram is drawn
by the `consort` package [@R-consort] using **grid graphics**, not ggplot. So
unlike every other plot here, you do **not** finish it with `+ theme_*()` or any
other ggplot layer. There is nothing to add a `+` to; the object renders itself.
If you want to change its look, you do it through the constructor's own
arguments, not through ggplot. The tracking helpers below give you the
exclusion-count audit you would otherwise build by hand.

## The data it needs

The tracker starts from a data frame with one row per patient and whatever
columns your exclusion rules will test. `sample_consort_data()` simulates a
cardiac-surgery cohort and returns a finished three-stage `hv_consort_tracker`,
which is the quickest way to see the shape of the object.

```{r}
#| label: consort_sample
tracker <- sample_consort_data(n = 300, seed = 42)
tracker
```

To build a tracker from your own data, start with `hv_consort_start()`, which
records the patient identifier and marks every patient as screened. Each
`hv_consort_exclude()` call then adds one exclusion stage. The rules are
two-sided formulas, `<condition> ~ "<reason>"`, evaluated against the data. The
first matching rule wins, and patients dropped in an earlier stage are skipped in
later ones, so each stage operates only on the survivors of the last. That last
point is what keeps the counts honest: a patient is removed once, for the first
reason that applies, and is not counted again.

The identifiers below are generated labels, not medical-record numbers. Keep
that same boundary in your own teaching and review code: this book uses
synthetic or public data only, never PHI.

```{r}
#| label: consort_build
set.seed(42)
cohort <- data.frame(
  patient_id  = sprintf("SYN%04d", 1:300),
  age         = sample(12:85, 300, replace = TRUE),
  had_surgery = sample(c(TRUE, FALSE), 300, replace = TRUE, prob = c(0.9, 0.1)),
  echo        = sample(c(TRUE, FALSE), 300, replace = TRUE, prob = c(0.85, 0.15))
)

tracker2 <- hv_consort_start(cohort, patient_id = patient_id, label = "Screened") |>
  hv_consort_exclude(
    label        = "Eligible",
    col          = "excl_screen",
    age < 18     ~ "Age < 18",
    !had_surgery ~ "No qualifying surgery"
  ) |>
  hv_consort_exclude(
    label = "Analyzed",
    col   = "excl_eligible",
    !echo ~ "Missing echocardiogram"
  )
tracker2
```

## Inspect it

Inspect the tracker and its stage summary before drawing the boxes. The printed
tracker confirms the identifier column and stage order; the summary gives the
arithmetic the diagram must reproduce. In this synthetic cohort, 300 screened
minus 73 screening exclusions leaves 227 eligible, and 227 minus 41 later
exclusions leaves 186 analysed.

```{r}
#| label: consort_inspect
class(tracker)
tracker
hv_consort_summary(tracker)
```

## Build it

`hv_consort()` derives the box layout from the tracker's stage metadata, and
`plot()` draws it. There is no bare-then-decorated pair here, because there is
nothing to decorate: the diagram comes out finished. By default every exclusion
column becomes a side box; pass a character vector to `side_box` if you want to
show only some of them.

```{r}
#| label: fig-consort
#| fig-cap: "CONSORT patient-flow diagram tracing a cardiac-surgery cohort from screening through exclusions to the analysed set"
fig <- hv_consort(tracker)
plot(fig)
```

## Read it

A CONSORT diagram is read top to bottom, with a glance sideways at each step.
Look for:

- **The arithmetic.** Every patient in the top box ends up either in a side box
  (excluded) or in the box below. The number entering a stage minus the side-box
  count should equal the number in the next box down. If it does not, an exclusion
  rule is double-counting or a patient is falling through; the tracker's own
  audit (below) is where you check.
- **The biggest side box.** The exclusion that drops the most patients is the one
  a reviewer scrutinises hardest. If most of your cohort leaves at a single step,
  be ready to defend that rule, because it is doing most of the work of defining
  who you studied.
- **The final n.** The bottom box is the number every later figure is built on.
  It is worth carrying that number forward into the captions of the survival and
  trend chapters so the reader never loses track of the denominator.

## Variations

### Auditing the cohort

The tracker is auditable, which is the point of building the diagram through it
rather than drawing boxes by hand. The `hv_consort_summary()` output inspected
above is a per-stage count table you can carry into a methods section.

`hv_consort_patients()` returns the patient IDs active at a stage, or the subset
excluded for a specific reason. This is how you go from a box on the diagram back
to the records behind it, to spot-check an exclusion or pull a list for a
sensitivity analysis. The output shown here contains only the synthetic
`PT####` labels returned by `sample_consort_data()`.

```{r}
#| label: consort_patients
# IDs still in the analysed cohort
head(hv_consort_patients(tracker, "Analyzed"))

# IDs dropped at screening for being under 18
head(hv_consort_patients(tracker, "Screened", reason = "Age < 18"))
```

## Pitfalls

- **Trying to theme it like a ggplot.** This is the mistake to avoid. The diagram
  is grid graphics, so `plot(fig) + theme_hv_manuscript()` (or any `+` layer) will
  error or be ignored. Adjust it through `hv_consort()`'s own arguments, such as
  `side_box`, not through ggplot.
- **Overlapping exclusion rules.** Because the first matching rule wins, the order
  of the formulas inside an `hv_consort_exclude()` call decides which reason a
  patient is logged under when more than one applies. Put the rule you want to be
  the recorded reason first, and read the summary table to confirm the counts land
  where you expect.
- **Numbers that do not reconcile.** If the boxes do not add up, do not patch the
  figure; fix the rules. Use `hv_consort_summary()` and `hv_consort_patients()` to
  find the patients in question rather than trusting the visual.
- **A stale final n in captions.** The bottom-box count is the denominator for
  everything downstream. Re-run the tracker whenever the cohort or the exclusion
  rules change, and update the figure captions that quote it, so the diagram and
  the rest of the manuscript never disagree.
