# Data governance and reproducibility

Most chapters in this book end in a figure, and the four in the *Tables* part
end in a table. This one comes first in practice, because reproducible analysis
starts before either. A documented dataset, typed and labelled variables, and a
manifest pinning the exact extract are the difference between a figure you can
defend in review and one you cannot. The `hvtiRutilities` [@R-hvtiRutilities] package supplies the
primitives for that work: synthetic data for examples, a data dictionary, label
and type maps, dataset manifests, and version diffs.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(hvtiRutilities)
library(gt)
```

## When to use it

Reach for this workflow at the start of any analysis that will end up in front
of a reviewer, a co-author, or your future self. The moment to run it is the
moment a data extract lands: before you fit a model, before you draw a curve,
while you still remember which warehouse pull this was and what date it covered.
The payoff comes later, when someone asks "which version of the data is this
figure built on" and you can answer with a hash instead of a shrug, or when a
re-delivery arrives and you need to know in seconds whether anything moved.

None of this requires touching protected health information. Everything below
runs on synthetic data, and the same helpers run unchanged against your real
extract once the pipeline is wired up. Think of it as the paperwork that makes
the science reproducible, kept light enough that you will actually do it.
Real patient extracts never belong in this recipe repository or its freeze
cache.

## The data it needs

The workflow needs a clinical data frame, one row per patient, with whatever
columns the study carries. For the examples we generate a synthetic survival
cohort so the chapter runs end to end without any real patient data.
`generate_survival_data()` is deterministic: fix the seed and every table and
figure downstream is reproducible.

```{r}
#| label: sample-data
dat <- generate_survival_data(n = 200, seed = 42)
str(dat)
```

Notice that each column carries a `label` attribute. That label is not
decoration. The governance helpers below read it and carry it forward into
dictionaries, tables, and figure axes, so a variable described once stays
described everywhere.

## Build it

The governance workflow is a short sequence of steps. Each one produces an
artifact you can save next to the analysis: a dictionary, a label map, a typing
check, a manifest, and a comparison report. We walk them in order.

### Step 1: a data dictionary

A data dictionary is the human-readable index of an extract, and it is the first
artifact worth saving next to the analysis. `data_dictionary()` returns a
per-column summary as a data frame, one row per variable, which renders straight
through `gt()` into a table you can drop into a supplement. The
[describing a dataset](data_tables.qmd) chapter takes the six columns apart and
says how to read them; here we want the artifact itself, dated and filed with
everything else that pins this analysis down.

```{r}
#| label: dict
dd <- data_dictionary(dat)
gt(dd) |>
  tab_header(title = "Data dictionary")
```

### Step 2: variable labels in one place

`label_map()` returns the current variable-to-label mapping as a data frame:
`key` and `label`, plus `label_full` and `truncated`. Holding labels in one
place is what keeps an axis title in a figure and a column header in a table
reading the same way. Change a label here and it propagates everywhere the map
is read, so you never hand-edit "Ejection Fraction" in five places and miss the
sixth.

The `label` column is trimmed to 40 characters, because a long descriptive
label has nowhere to go on an axis or in a column header. Nothing is lost: the
source text stays in `label_full`, and `truncated` marks the rows that were
cut. So `subset(lm, truncated)` tells you which labels a figure is going to
abbreviate, before a reviewer asks you about one. Pass `label_max` if you want
a different width, or `Inf` to leave every label whole.

```{r}
#| label: labels
lm <- label_map(dat)
gt(head(lm, 10))
```

### Step 3: infer and check R data types

Raw extracts arrive with everything as character or numeric, and the wrong type
quietly breaks an analysis: an ID read as a number gets averaged, a 0/1 flag
read as a count gets summed. `r_data_types()` classifies each column by role and
returns the dataset with columns coerced to the inferred type, so character IDs
become factors and 0/1 indicators become logical. Read that sentence twice: it
returns the *coerced data*, not a report about it. To see what changed, compare
the source class against the inferred class yourself.

```{r}
#| label: dtypes
typed <- r_data_types(dat)
data.frame(
  variable = names(dat),
  source   = vapply(dat,   \(x) class(x)[1], character(1)),
  inferred = vapply(typed, \(x) class(x)[1], character(1)),
  row.names = NULL
) |>
  subset(source != inferred) |>
  gt() |>
  tab_header(title = "Columns retyped by r_data_types()")
```

### Step 4: pin the extract with a manifest

A dictionary documents *what* the data is; a manifest documents *which* data it
is. A manifest records provenance for every extract: the source system, the
extract date, the row count, and a SHA-256 hash of the file. `update_manifest()`
appends an entry, hashing the file on disk, and `verify_manifest()` re-hashes the
data directory against the manifest to confirm nothing has drifted since you
wrote it down. The chunk below is not evaluated because it writes files to disk;
run it interactively, then commit the manifest (never the data) to the repo.

```{r}
#| label: manifest
#| eval: false
saveRDS(dat, "cohort.rds")
mp <- tempfile(fileext = ".yaml")

update_manifest(
  file         = "cohort.rds",
  manifest_path = mp,
  extract_date = Sys.Date(),
  n_rows       = nrow(dat),
  source       = "CORR DWH"
)

verify_manifest(manifest_path = mp, data_dir = ".", stop_on_error = FALSE)
```

A representative manifest entry looks like this. The hash is the load-bearing
field: it is what lets `verify_manifest()` catch a file that changed under you.

```yaml
datasets:
- file: cohort.rds
  extract_date: '2026-05-28'
  n_rows: 200
  sha256: 8b3e0188eae294e61a16f0e09e3989e8b5249601209bde987fb5153c9ab1bd76
  source: CORR DWH
```

### Step 5: compare versions on re-delivery

Data warehouses re-deliver extracts, and a re-delivery is rarely a no-op: rows
get added, a column is renamed, a coding changes. `compare_datasets()` diffs the
old and new versions and reports row-count and structural changes through its
print method. It is the fast first check that a fresh pull is what you asked for
before you rerun anything on it.

```{r}
#| label: compare
old <- generate_survival_data(n = 150, seed = 42)
new <- generate_survival_data(n = 200, seed = 99)
compare_datasets(old, new)
```

## Read it

One of these artifacts is a report you read rather than a file you store. The
data dictionary is a report too, but its reading guide lives with the function,
over in [describing a dataset](data_tables.qmd).

A **dataset-comparison report** is read for the diff, not the totals. A change in
row count is expected on re-delivery and is fine if it matches what the warehouse
told you to expect. The lines that should make you stop are structural ones: a
column that appeared, disappeared, or changed type. Those silently break joins
and recodes downstream, so confirm each one is intentional before you build
anything on the new extract.

## Pitfalls

- **Unpinned extracts.** A figure built on "the data I pulled sometime last
  month" is not reproducible. Write a manifest the day the extract lands, commit
  it, and reference its hash from the analysis. The hash is cheap insurance
  against a silently swapped file.
- **Silent type coercion.** `r_data_types()` returns the *coerced data frame*,
  not a summary of what it did. If you treat its output as a report you will miss
  that it changed your data. Diff source class against inferred class, as in
  step 3, and keep the typed object as the one you analyse.
- **Manifests not re-verified after a re-delivery.** Writing a manifest once is
  not enough. When a new extract arrives, run `verify_manifest()` and
  `compare_datasets()` again. A manifest that nobody re-checks is documentation,
  not a guard rail.
- **PHI in the repo.** Commit the manifest, the dictionary, and the code; never
  commit the data file itself. The whole point of the hash is to pin the extract
  without carrying protected health information into version control. Examples in
  a book stay on synthetic data for the same reason.
