# Describing a dataset

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(hvtiRutilities)
```

## When to use it

The first thing most of us do with a new SAS dataset is run `PROC CONTENTS`,
then `PROC MEANS`. Neither one produces a table anybody will publish. You run
them because you cannot decide what belongs in Table 1 until you know what you
are holding: which variables are there, what type each one came in as, and how
much of each is actually populated.

That is this recipe. It is the *pre-analysis* table, the one that runs before
the Table 1 in the [publication tables](qt_tables.qmd) chapter has a shape.
`hvtiRutilities` gives you both procedures under the names you already type,
`proc_contents()` and `proc_means()`, plus `data_dictionary()`, which flattens
the first into the one-page artifact you hand a collaborator.

Reach for it whenever a dataset is new to you, has just been re-extracted, or
arrives with someone telling you it is "the same as last time." Ten seconds of
`proc_contents()` catches the variable that came back as character this quarter,
or the one that is four-fifths missing, while it is still cheap to ask about.

## The data it needs

Any data frame. Labels, when the columns carry them, are picked up and
reported; unlabelled columns fall back to the variable name. We use
`generate_survival_data()`, the same simulated cohort the
[publication tables](qt_tables.qmd) chapter builds its Table 1 from, so you can
read the two chapters against one dataset.

```{r}
#| label: dd-data
dta <- hvtiRutilities::generate_survival_data(n = 200, seed = 42)
dim(dta)
```

## Build it

### What is in this dataset — `proc_contents()`

`proc_contents()` prints what SAS prints: a header carrying the observation and
variable counts, then one row per variable with its creation position, name, SAS
type, format, and label. Next to those it adds three columns you would otherwise
go looking for separately: `class`, the R type sitting behind the SAS one;
`n_unique`, how many distinct values the column holds; and `pct_missing`, how
much of it is empty.

```{r}
#| label: dd-contents
proc_contents(dta)
```

Variables come back sorted alphabetically, matching SAS's *Alphabetic List of
Variables and Attributes*. Pass `order = "varnum"` for creation order instead.
The `num` column reports creation position either way, so the original order is
always recoverable by eye.

Two columns repay a slow read. `type` is deliberately two-valued, `Char` or
`Num`, because that is all SAS storage offers: `sex` is a factor in R and
reports `Char` here, while `class` keeps the R truth alongside it. Then
`pct_missing`, which is where the surprises live. Here 23 of the 24 variables
are complete and `iv_reop` is 83.5% missing, which is exactly right for a
reoperation follow-up time. It exists only for the 33 patients who had a
reoperation. Missingness with a reason is structure; missingness you cannot
account for is a question for whoever built the extract.

### What the numbers look like — `proc_means()`

`proc_means()` summarises the numeric columns you name, defaulting to SAS's own
default five statistics.

```{r}
#| label: dd-means
proc_means(dta, vars = c("age", "bmi", "gfr_bs"))
```

Omit `vars` and every numeric column is analysed, the way `PROC MEANS` behaves
with no `VAR` statement. `stats =` takes the SAS keywords you already know,
including `"median"`, `"q1"`, `"q3"`, `"nmiss"`, `"stderr"`, and any `"pNN"`
percentile.

Quantiles are worth one note. They use `stats::quantile(type = 2)`, which is the
R equivalent of SAS's `QNTLDEF=5`, not R's own `type = 7` default. On a large
sample the two agree. On the small, even-numbered subgroups a clinical
stratification hands you, they do not: for `c(1, 2, 3, 4)` the first quartile is
1.5 in SAS and 1.75 under R's default. The median agrees, which is how the
disagreement stays hidden.

### Stratified by a class variable

Give it `class =` and you get the same table once per group, the `CLASS`
statement by another name.

```{r}
#| label: dd-means-class
proc_means(dta, vars = c("age", "bmi"), class = "sex")
```

Rows sort by analysis variable first, then by class level, so you read a
variable's groups down the page rather than hunting across. A factor class
variable orders by its declared levels rather than alphabetically, matching
SAS's `ORDER=INTERNAL`, which is what keeps an ordered clinical scale in
clinical sequence.

Check the `n` column against your cohort size. Rows with a missing value in any
class variable are dropped, again matching SAS, so a total that comes up short
tells you how many patients had no group to sit in. Here 77 and 123 add to 200,
and nothing was dropped.

### The whole dictionary — `data_dictionary()`

`data_dictionary()` is `proc_contents()` in creation order with the SAS-storage
columns dropped and a summary put in their place. Out go `num`, `type`, and
`format`, which describe how a value is stored rather than what it is. In comes
a compact `summary` giving min, median, and max for a numeric variable, the
levels for a factor, the TRUE percentage for a logical. That leaves six columns:
`variable`, `label`, `class`, `n_unique`, `pct_missing`, and `summary`.

```{r}
#| label: dd-dictionary
data_dictionary(dta)
```

That summary column is what makes the table hand-off-able, and it is also where
degenerate variables announce themselves. `lvmsi_b` reads `40 / 40 / 40` with an
`n_unique` of 1, so it is a constant. A constant costs a row in Table 1 and
tells the reader nothing. Send this frame through `write.csv()` and you have the
data-dictionary appendix a protocol or a data-sharing agreement asks for.

## Read it

Three outputs, three questions.

`proc_contents()` answers *what is here*: names, types, labels, and how complete
each column is. Read it top to bottom once, then read `pct_missing` on its own,
column-first rather than row-first. That single pass is most of the value.

`proc_means()` answers *what do the numbers look like*: centre, spread, and
range per variable, and per group once you add `class =`. Read `min` and `max`
before you read the mean. Impossible values show up at the ends, and a mean is
good at hiding one. The `age` minimum of 1 in this simulated cohort is the kind
of thing you would query on real data.

`data_dictionary()` answers *what do I hand someone else*. It is the one of the
three you save to a file.

None of the three is a publication table. They are the reconnaissance you do
before deciding what the publication table should contain, and
[Table 1](qt_tables.qmd) comes after.

## Pitfalls

- **`pct_missing` is `NaN`, not `0`, when the frame has no rows.** Filter a
  cohort down to nothing and every `pct_missing` comes back `NaN`, because
  `mean(is.na(x))` over an empty vector is 0/0. The table looks broken and is
  in fact reporting honestly: the proportion of missing values among no values
  is undefined, and a `0` there would assert that nothing is missing. If you
  format the column for display, handle `NaN` explicitly rather than letting it
  print raw next to real percentages, and treat it as the signal it is that your
  filter removed everything.

- **`class =` groups on the stored value, not the printed one.** Two values that
  render identically but differ underneath stay separate groups, and you get
  back two rows that look like duplicates of each other. A trailing space in a
  character code does it, and so does a numeric group variable whose values
  differ below print precision. The behaviour is correct, since grouping on the
  rendered string would quietly merge values that are genuinely distinct, but it
  will mislead you if you are diagnosing from printed output alone. When a group
  count looks doubled, run `unique()` or `table()` on the class variable and
  compare what you get to what the table showed you.
