# Density plots

Is a continuous measurement symmetric, skewed, or hiding more than one cluster?
A density curve answers that shape question without tying the answer to one set
of histogram bins.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

A histogram answers "how many patients fall in each bin," and that is often
enough. But the moment you want to compare the *shape* of two distributions, or
you want a smooth curve to put in a figure rather than a staircase of bars, the
kernel density estimate is the cleaner tool. A density plot draws a continuous
curve whose height at any value reflects how concentrated the data are there,
and the area under the whole curve is one. Reach for it when the question is
about the form of a single continuous variable: is it roughly symmetric, is it
skewed toward high values, does it have one peak or two?

`hvtiPlotR` does not provide a density helper, so this chapter builds the plot
directly with ggplot2's [@R-ggplot2] `geom_density()`. We still style it with the
house theme `theme_hv_manuscript()` so a density panel sits comfortably next to
the survival curves and scatter plots in the same paper. The pattern is the one
you will use for any bare-ggplot recipe in the book: write the `ggplot()` call
yourself, then finish with `+ theme_hv_manuscript()`.

## The data it needs

A density plot needs one continuous column. To compare groups you add a
categorical column to map to colour. We use `hvtiPlotR::sample_eda_data()` with
`ef` (ejection fraction) as the measurement and `valve_morph` (valve
morphology: Bicuspid, Tricuspid, or Unicuspid) as the comparison group. Naming
the columns directly makes the clinical question clear when you copy the recipe.

```{r}
#| label: data
dat <- sample_eda_data()
num <- "ef"
grp <- "valve_morph"
c(numeric = num, group = grp)
```

The `.data[[num]]` idiom you will see below is how ggplot2 reads a column whose
name lives in a variable rather than typed out as a bare symbol. It lets the
same recipe run after you replace `num` with another continuous column name.

## Inspect it

There is no constructor object to inspect in this direct ggplot2 recipe, so
inspect the measurement and the rows each curve will actually use before
smoothing. `geom_density()` drops a row with a missing ejection fraction; the
grouped version also needs a nonmissing group. These nonmissing counts are the
information a normalized curve will otherwise hide.

```{r}
#| label: density_inspect
single_curve_rows  <- !is.na(dat[[num]])
grouped_curve_rows <- single_curve_rows & !is.na(dat[[grp]])

summary(dat[[num]][single_curve_rows])
c(single_curve_n = sum(single_curve_rows),
  grouped_curve_n = sum(grouped_curve_rows))
table(dat[[grp]][grouped_curve_rows], useNA = "ifany")
```

## Build it

Start with a single curve so you can see the estimate before any grouping. The
`fill` gives the curve a soft body and `alpha` keeps it from reading as a solid
block.

```{r}
#| label: fig-density
#| fig-cap: "Kernel density estimate of a single continuous variable, scaled so the area under the curve is one"
ggplot(dat, aes(x = .data[[num]])) +
  geom_density(fill = "grey70", alpha = 0.7) +
  labs(
    title = "Distribution of a single numeric variable",
    x = "Ejection fraction (%)",
    y = "Density"
  ) +
  theme_hv_manuscript()
```

That curve is a smoothed version of the histogram you would otherwise draw. The
y-axis is density, not count: it is scaled so the total area is one, which is
what lets you put two of these on the same panel and compare them fairly even
when the groups have different sizes.

## Read it

A density curve gives you the shape of the distribution at a glance. The x-axis
is ejection fraction and the y-axis is normalized density, so the area under the
curve is one. Compare concentration across the x-axis, not curve height as a
patient count. The smoother describes this sample; it does not establish a
clinical effect or explain what caused the shape. Look for:

- **Peaks (modes).** One hump means the data cluster around a single typical
  value. Two humps (bimodal) can mean you have mixed two populations that
  belong in separate groups, an early-era and a late-era cohort, say. Check the
  source groups before deciding what the second peak means.
- **Skew.** A long tail to the right (a few large values pulling the curve out)
  is right skew; a long left tail is left skew. Skew tells you whether the mean
  and the median will agree, and whether a log transform might help downstream.
- **Width.** A tall, narrow curve means the values are tightly concentrated; a
  low, broad curve means they are spread out. The height is not a count, so do
  not read a taller curve as "more patients."

One thing to keep in the back of your mind: the curve depends on a smoothing
choice called the bandwidth, and the same data can look smoother or bumpier
depending on it. We come back to that in the pitfalls.

## Adapt it

### Grouped density

Mapping the grouping variable to `fill` overlays one density curve per group,
with transparency so the overlapping regions stay readable. This is the
comparison you actually want most of the time: not "what does this variable look
like" but "does it look different across valve morphologies?"

```{r}
#| label: fig-density-grouped
#| fig-cap: "Overlaid density curves by valve morphology, each normalised to area one so groups of different sizes compare on shape"
ggplot(dat, aes(x = .data[[num]], fill = .data[[grp]])) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Density by group",
    x = "Ejection fraction (%)",
    y = "Density",
    fill = "Valve morphology"
  ) +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

Because each curve integrates to one, a small group and a large group are drawn
on the same footing. That is the right default for comparing shape, but it also
hides how many patients sit under each curve, which is the trap we flag below.
The comparison is descriptive and unadjusted; separation between curves does
not show that morphology caused a difference in ejection fraction.

### Tuning the bandwidth

The single most consequential knob is the bandwidth, passed through `adjust`
(a multiplier on the automatic bandwidth). Drawing two settings side by side is
the honest way to show how much of a bump is signal and how much is the smoother.

```{r}
#| label: fig-density-bandwidth
#| fig-cap: "The same data drawn at a narrow and a wide bandwidth, showing how the smoothing choice shapes apparent structure"
ggplot(dat, aes(x = .data[[num]])) +
  geom_density(aes(colour = "wiggly (adjust = 0.4)"),
               adjust = 0.4, fill = NA, linewidth = 0.8) +
  geom_density(aes(colour = "smooth (adjust = 1.5)"),
               adjust = 1.5, fill = NA, linewidth = 0.8) +
  scale_colour_manual(values = c("wiggly (adjust = 0.4)" = "firebrick",
                                 "smooth (adjust = 1.5)" = "steelblue"),
                      name = "Bandwidth") +
  labs(
    title = "Same data, two bandwidths",
    x = "Ejection fraction (%)",
    y = "Density"
  ) +
  theme_hv_manuscript()
```

If a feature survives both settings it is less sensitive to the smoothing
choice. A feature that appears only at the small bandwidth needs more checking;
the two curves are a sensitivity display, not a test of whether a peak is real.

## Deliver it

Carry the finished panel to [Finish and deliver the result](publications.qmd)
for manuscript, poster, or slide output. Put the group sizes and the bandwidth
choice in the caption whenever the grouped shape is the result.

## Pitfalls

- **Over-smoothing and under-smoothing.** Too large a bandwidth erases real
  structure, a second mode flattens into a shoulder. Too small a bandwidth
  invents structure, every cluster of three points becomes its own little peak.
  When a bump matters to your story, redraw it at a wider and a narrower
  bandwidth before you commit, the way @fig-density-bandwidth does.
- **Treating a density like a histogram.** The y-axis is a density, not a count.
  A taller curve does not mean more patients; it means the data are more
  concentrated there. If the reader needs counts, draw a histogram or label the
  group sizes.
- **Comparing groups with very different n.** Each curve is normalised to area
  one, so a 19-patient group and a 167-patient group look equally trustworthy
  even though one rests on a tenth of the data. The small group's curve is far
  more sensitive to a handful of values. Note the group sizes in the caption so
  the reader weights the comparison correctly.
- **Densities that spill past the possible range.** A kernel density will happily
  put mass below zero for a strictly positive measurement, or above 100 for a
  percent. The smoother does not know the bounds. Constrain the axis or
  transform the variable rather than letting the curve imply impossible values.
