# Tables that accompany figures

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(hvtiPlotR)
library(gt)
library(ggplot2)
library(patchwork)
```

Many manuscript figures are paired with a small numeric table: a
numbers-at-risk strip below a Kaplan-Meier curve, or the count panel
beneath a longitudinal participation bar chart. The `hvtiPlotR` plot
objects carry these companion tables alongside the plot data, so we can
pull the numbers straight out and render them with `gt` [@R-gt], the
right tool, as in the [publication tables](qt_tables.qmd) chapter, when a
table's destination is this book or another HTML document.
[Manuscript tables](hv_tables.qmd) headed for a Word file go through
`hvtiRtables` instead, which turns a `gtsummary` summary into a CORR-compliant
flat-header table.

## When to use it

A figure shows the shape of a result; the companion table gives the reader the
numbers the shape rests on. The two go together when a curve hides how much data
backs each region of it. The clearest case is the numbers-at-risk strip beneath
a Kaplan-Meier curve: the curve might look confident out at ten years, but if
only eight patients remain at risk there, the at-risk row is what tells the
reader to read that tail with caution. Reach for a companion table whenever the
figure invites a question the plot cannot answer on its own ("based on how
many?") and you want the answer on the same page rather than buried in the text.

The at-risk count is a denominator, not a second estimate. At time $t$ it counts
patients who are still observed and event-free immediately before $t$. An event
changes the survival curve and removes that patient from later risk sets;
censoring removes a patient from later risk sets without creating a survival
step. Thus the count tells you how much information supports $S(t)$ and its
confidence interval, but it does not tell you the survival probability by
itself.

We work through two cases the package already supports: the at-risk strip that a
survival curve needs beside it, and the participation counts that sit beneath a
longitudinal bar chart. In both, the numbers already live on the plot object, so
we never recompute them and risk a table that disagrees with its own figure.

## Numbers at risk

`hv_survival()` stores its companion tables in the object's `$tables`
slot. The numbers-at-risk strip lives in `$tables$risk` (one row per
strata × report-time) and the per-time-point summary in `$tables$report`.

```{r}
#| label: km_build
km <- hv_survival(
  sample_survival_data(n = 500, seed = 42),
  report_times = seq(0, 20, 5)
)
names(km$tables)
km$tables$risk
```

We render the at-risk table directly with `gt`, relabelling the columns
for presentation.

```{r}
#| label: km_risk_gt
km$tables$risk |>
  gt() |>
  tab_header(
    title = "Numbers at Risk",
    subtitle = "Patients remaining under follow-up at each report time"
  ) |>
  cols_label(
    strata      = "Stratum",
    report_time = "Time (years)",
    n.risk      = "At risk (n)"
  )
```

The companion report table carries the survival estimates and confidence
limits at the same report times. It lets you quote a 5- or 10-year estimate
together with the denominator that supports it.

```{r}
#| label: km_report_gt
km$tables$report |>
  gt() |>
  tab_header(title = "Survival Estimates at Report Times") |>
  cols_label(
    strata      = "Stratum",
    report_time = "Time (years)",
    surv        = "S(t)",
    lower       = "95% LCL",
    upper       = "95% UCL",
    n.risk      = "At risk",
    n.event     = "Events"
  ) |>
  fmt_number(columns = c(surv, lower, upper), decimals = 3)
```

For the figure itself, build the curve and the risk panel from that same `km`
object. `hv_atrisk_compose()` reads the finished curve's x-range and applies it
to the table, so the 0-, 5-, 10-, 15-, and 20-year counts sit directly beneath
their curve positions. This is why the report times and axis breaks are set to
the same sequence.

```{r}
#| label: fig-tables-atrisk
#| fig-cap: "Kaplan-Meier survival with the supporting at-risk denominator aligned at 0, 5, 10, 15, and 20 years"
#| fig-height: 5
km_curve <- plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values = c(All = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after operation", y = "Freedom from death (%)") +
  theme_hv_manuscript()

km_risk <- hv_atrisk(km)
hv_atrisk_compose(km_curve, km_risk)
```

## Longitudinal count table

`hv_longitudinal()` produces a two-panel figure: a grouped bar chart of
participation counts and a numeric table panel below it. The underlying
counts live in the object's `$data` slot in long format (one row per
time-window × series), with the column roles recorded in `$meta`.

```{r}
#| label: lc_build
lc <- hv_longitudinal(sample_longitudinal_counts_data(n_patients = 300, seed = 42L))
names(lc)
lc$meta[c("x_col", "group_col", "count_col")]
str(lc$data)
```

We reshape the long counts into a wide table (one row per follow-up
window, one column per series) and render it with `gt`. This is the
tabular companion to the bar chart's table panel.

```{r}
#| label: lc_table_gt
lc_wide <- lc$data |>
  tidyr::pivot_wider(
    names_from  = series,
    values_from = count
  )
lc_wide

lc_wide |>
  gt(rowname_col = "time_label") |>
  tab_header(
    title    = "Longitudinal Participation Counts",
    subtitle = "Patients and measurements at each follow-up window"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 0) |>
  tab_stubhead(label = "Follow-up window")
```

The `gt` version above is for a standalone table, say in the manuscript body or
a supplement. When the numbers belong *inside* the figure, the package draws
them as a table panel you stack under the chart. For reference, the same counts
rendered as the in-figure table panel appear below the bar chart when the two
are composed with `patchwork`. Building both from the same `lc$data` keeps the
standalone table and the in-figure panel telling one story.

```{r}
#| label: lc_figure
#| fig-height: 5
p_bar <- plot(lc) +
  scale_fill_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    name   = NULL
  ) +
  labs(x = NULL, y = "Count (n)")

p_tbl <- plot(lc, type = "table") +
  scale_colour_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    guide  = "none"
  )

p_bar / p_tbl + plot_layout(heights = c(3, 1))
```

## Pitfalls

- **At-risk counts must match the curve's time axis.** The risk strip is only
  useful if its columns sit under the same time points the curve uses. If the
  plot runs 0 to 20 years on a five-year grid, the at-risk table needs the same
  grid, or the reader cannot line a count up with the part of the curve it
  describes. Set `report_times` explicitly when you build `km`, use those times
  as the curve's breaks, and let `hv_atrisk_compose()` align the finished
  panels. If you later crop the curve, inspect the table again.
- **Keeping the table and figure in sync.** The reason we pull numbers off the
  plot object rather than recomputing them is to keep a single source of truth.
  The moment you summarise the cohort separately for the table, the two can
  drift: a filter applied to one and not the other, a different handling of
  missing follow-up, a stale data load. If you must build a companion table by
  hand, derive it from the exact frame the figure was built from, and re-derive
  both whenever the data change. A figure and a table that disagree are worse
  than either alone, because a reader cannot tell which to believe.
