# Tables that accompany figures

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(hvtiPlotR)
library(gt)
library(ggplot2)
library(patchwork)
```

Many manuscript figures are paired with a small numeric table: a
numbers-at-risk strip below a Kaplan-Meier curve, or the count panel
beneath a longitudinal participation bar chart. The `hvtiPlotR` plot
objects carry these companion tables alongside the plot data, so we can
pull the numbers straight out and render them with `gt` [@R-gt] — the
right tool, as in the [publication tables](qt_tables.qmd) chapter, when a
table's destination is this book or another HTML document. Manuscript
tables headed for a Word file go through `hvtiRtables` instead, which
turns a `gtsummary` summary into a CORR-compliant flat-header table.

## When to use it

A figure shows the shape of a result; the companion table gives the reader the
numbers the shape rests on. The two go together when a curve hides how much data
backs each region of it. The clearest case is the numbers-at-risk strip beneath
a Kaplan-Meier curve: the curve might look confident out at ten years, but if
only eight patients remain at risk there, the at-risk row is what tells the
reader to read that tail with caution. Reach for a companion table whenever the
figure invites a question the plot cannot answer on its own ("based on how
many?") and you want the answer on the same page rather than buried in the text.

We work through two cases the package already supports: the at-risk strip that a
survival curve needs beside it, and the participation counts that sit beneath a
longitudinal bar chart. In both, the numbers already live on the plot object, so
we never recompute them and risk a table that disagrees with its own figure.

## Numbers at risk

`hv_survival()` stores its companion tables in the object's `$tables`
slot. The numbers-at-risk strip lives in `$tables$risk` (one row per
strata × report-time) and the per-time-point summary in `$tables$report`.

```{r}
#| label: km_build
km <- hv_survival(sample_survival_data(n = 500, seed = 42))
names(km$tables)
str(km$tables$risk)
```

We render the at-risk table directly with `gt`, relabelling the columns
for presentation.

```{r}
#| label: km_risk_gt
km$tables$risk |>
  gt() |>
  tab_header(
    title = "Numbers at Risk",
    subtitle = "Patients remaining under follow-up at each report time"
  ) |>
  cols_label(
    strata      = "Stratum",
    report_time = "Time (years)",
    n.risk      = "At risk (n)"
  )
```

The companion report table carries the survival estimates and confidence
limits at the same report times — useful as the figure caption's numeric
backing.

```{r}
#| label: km_report_gt
km$tables$report |>
  gt() |>
  tab_header(title = "Survival Estimates at Report Times") |>
  cols_label(
    strata      = "Stratum",
    report_time = "Time (years)",
    surv        = "S(t)",
    lower       = "95% LCL",
    upper       = "95% UCL",
    n.risk      = "At risk",
    n.event     = "Events"
  ) |>
  fmt_number(columns = c(surv, lower, upper), decimals = 3)
```

## Longitudinal count table

`hv_longitudinal()` produces a two-panel figure: a grouped bar chart of
participation counts and a numeric table panel below it. The underlying
counts live in the object's `$data` slot in long format (one row per
time-window × series), with the column roles recorded in `$meta`.

```{r}
#| label: lc_build
lc <- hv_longitudinal(sample_longitudinal_counts_data(n_patients = 300, seed = 42L))
names(lc)
lc$meta[c("x_col", "group_col", "count_col")]
str(lc$data)
```

We reshape the long counts into a wide table (one row per follow-up
window, one column per series) and render it with `gt`. This is the
tabular companion to the bar chart's table panel.

```{r}
#| label: lc_table_gt
lc_wide <- lc$data |>
  tidyr::pivot_wider(
    names_from  = series,
    values_from = count
  )
lc_wide

lc_wide |>
  gt(rowname_col = "time_label") |>
  tab_header(
    title    = "Longitudinal Participation Counts",
    subtitle = "Patients and measurements at each follow-up window"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 0) |>
  tab_stubhead(label = "Follow-up window")
```

The `gt` version above is for a standalone table, say in the manuscript body or
a supplement. When the numbers belong *inside* the figure, the package draws
them as a table panel you stack under the chart. For reference, the same counts
rendered as the in-figure table panel appear below the bar chart when the two
are composed with `patchwork`. Building both from the same `lc$data` keeps the
standalone table and the in-figure panel telling one story.

```{r}
#| label: lc_figure
#| fig-height: 5
p_bar <- plot(lc) +
  scale_fill_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    name   = NULL
  ) +
  labs(x = NULL, y = "Count (n)")

p_tbl <- plot(lc, type = "table") +
  scale_colour_manual(
    values = c(Patients = "steelblue", Measurements = "firebrick"),
    guide  = "none"
  )

p_bar / p_tbl + plot_layout(heights = c(3, 1))
```

## Pitfalls

- **At-risk counts must match the curve's time axis.** The risk strip is only
  useful if its columns sit under the same time points the curve uses. If the
  plot runs 0 to 20 years on a five-year grid, the at-risk table needs the same
  grid, or the reader cannot line a count up with the part of the curve it
  describes. Because `$tables$risk` is computed by the same constructor that
  drew the curve, the report times already agree; the trap appears when you
  hand-edit one axis or crop the plot with `coord_cartesian()` and forget the
  table still shows the full range. Change the axis in one place and check the
  other followed.
- **Keeping the table and figure in sync.** The reason we pull numbers off the
  plot object rather than recomputing them is to keep a single source of truth.
  The moment you summarise the cohort separately for the table, the two can
  drift: a filter applied to one and not the other, a different handling of
  missing follow-up, a stale data load. If you must build a companion table by
  hand, derive it from the exact frame the figure was built from, and re-derive
  both whenever the data change. A figure and a table that disagree are worse
  than either alone, because a reader cannot tell which to believe.
