# Function-to-recipe map

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

Use this map when you know the function name but not the recipe. It answers the
question that arrives when you are reading someone else's script: *what is
`hv_atrisk_compose()`, and where in this book is it worked through?*

The map is a shortcut back to a reproducible example, not a substitute for one.
It keeps a function lookup from becoming an invitation to reconstruct a workflow
from its argument list.

## How to read it

**Role** says what the function is for in one line. **Recipe** links to the
chapter that works it through. **Coverage** distinguishes a function this book
demonstrates from one it names but leaves to the package reference.

The table is generated rather than typed. The function lists below are an
explicit contract, and the chapter column is computed by scanning the book's own
sources at render time, so the two can disagree and say so. If a function is
added to the contract and no chapter picks it up, its coverage flips to *Package
reference* on the next render without anyone remembering to update a list.

```{r}
#| label: contract
expected <- list(
  hvtiR = c("install", "status", "update", "doctor"),
  hvtiPlotR = c(
    "hv_alluvial", "hv_atrisk", "hv_atrisk_compose", "hv_balance",
    "hv_consort", "hv_consort_exclude", "hv_consort_patients",
    "hv_consort_start", "hv_consort_summary", "hv_eda", "hv_followup",
    "hv_ggsave_dims", "hv_hazard", "hv_legend_inside", "hv_longitudinal",
    "hv_mirror_hist", "hv_nnt", "hv_nonparametric", "hv_ordinal",
    "hv_ph_location", "hv_ppt_palette", "hv_ppt_series", "hv_sankey",
    "hv_spaghetti", "hv_stacked", "hv_survival", "hv_survival_difference",
    "hv_trends", "hv_upset", "hv_venn", "save_manuscript", "save_ppt",
    "theme_hv_manuscript", "theme_hv_poster", "theme_hv_ppt_dark",
    "theme_hv_ppt_light"),
  hvtiRtables = c(
    "hv_tbl_summary", "hv_man_table", "hv_man_footnotes", "hv_man_table_save",
    "hv_check_docx", "hv_man_table_jtcvs", "hv_test_footnotes_jtcvs",
    "hv_man_table_save_jtcvs"),
  ggRandomForests = c(
    "gg_error", "gg_rfsrc", "gg_survival", "gg_vimp", "gg_shap", "gg_variable",
    "gg_partial_rfsrc", "gg_roc", "gg_brier", "gg_varpro", "gg_beta_varpro",
    "gg_ivarpro", "gg_isopro", "gg_beta_uvarpro", "gg_udependent",
    "gg_sdependent", "gg_partial_varpro", "gg_rhf", "gg_auct",
    "gg_rhf_importance", "gg_tune_rhf"),
  TemporalHazard = c(
    "hazard", "hzr_argument_mapping", "hzr_bootstrap", "hzr_calibrate",
    "hzr_competing_risks", "hzr_deciles", "hzr_gof", "hzr_kaplan",
    "hzr_nelson", "hzr_phase", "hzr_phase_cumhaz", "hzr_phase_hazard",
    "hzr_read_outhaz", "hzr_stepwise", "hzr_translate_sas", "stepwise_trace")
)

## Every name in the contract must still be exported. A rename upstream fails
## the build here rather than leaving a dead entry in the map.
missing <- unlist(lapply(names(expected), function(p)
  setdiff(expected[[p]], getNamespaceExports(p))))
stopifnot(length(missing) == 0)
```

```{r}
#| label: roles
roles <- c(
  # hvtiR: family lifecycle
  install = "Install the whole hvtiR family on a fresh machine",
  status = "Compare installed family versions against GitHub main",
  update = "Install only the missing or stale family members",
  doctor = "Diagnose R, pak, GitHub access, and member status",

  # hvtiPlotR: constructors
  hv_survival = "Kaplan-Meier survival estimate and its risk table",
  hv_survival_difference = "Difference between two survival curves",
  hv_nnt = "Number needed to treat over follow-up",
  hv_hazard = "Parametric or life-table hazard estimate",
  hv_nonparametric = "Nonparametric hazard and cumulative hazard",
  hv_ordinal = "Ordinal outcome over time",
  hv_atrisk = "Numbers-at-risk table aligned to a time axis",
  hv_atrisk_compose = "Stack a curve over its numbers-at-risk panel",
  hv_followup = "Follow-up completeness over time",
  hv_balance = "Covariate balance before and after adjustment",
  hv_consort = "CONSORT patient-flow diagram",
  hv_consort_start = "Open a CONSORT flow at the screened cohort",
  hv_consort_exclude = "Add an exclusion arm to a CONSORT flow",
  hv_consort_patients = "Add a retained-patients node to a CONSORT flow",
  hv_consort_summary = "Summarise a CONSORT flow as counts",
  hv_sankey = "Sankey flow between cluster states",
  hv_alluvial = "Alluvial flow across milestones",
  hv_upset = "UpSet plot of set intersections",
  hv_venn = "Venn diagram for two or three sets",
  hv_trends = "Grouped trend series over time",
  hv_spaghetti = "Per-patient trajectories across repeated measures",
  hv_longitudinal = "Longitudinal counts with a composite panel",
  hv_stacked = "Stacked categorical composition",
  hv_mirror_hist = "Back-to-back histogram for two groups",
  hv_eda = "Quick exploratory view of a data frame",

  # hvtiPlotR: finishing and delivery
  theme_hv_manuscript = "Journal theme: small type, minimal chrome, no legend",
  theme_hv_poster = "Poster theme: large type for arm's-length reading",
  theme_hv_ppt_light = "Slide theme for white or light-grey backgrounds",
  theme_hv_ppt_dark = "Slide theme for navy or dark-gradient decks",
  hv_ppt_series = "Matched colour and shape scales plus a slide theme",
  hv_ppt_palette = "The slide palette on its own",
  hv_legend_inside = "Put a key in the emptiest corner of the panel",
  hv_ggsave_dims = "Fix the panel content area, not the file size",
  hv_ph_location = "Slide coordinates so a panel lands in the same place",
  save_manuscript = "Write a manuscript-geometry figure file",
  save_ppt = "Write the plot onto a slide as editable vector shapes",

  # hvtiRtables
  hv_tbl_summary = "Summarise a cohort into a Table 1",
  hv_man_table = "Render a summary as a manuscript table",
  hv_man_footnotes = "Attach footnotes to a manuscript table",
  hv_man_table_save = "Write a manuscript table to Word",
  hv_check_docx = "Check a written Word table against the format rules",
  hv_man_table_jtcvs = "Render a table in the JTCVS format",
  hv_test_footnotes_jtcvs = "JTCVS footnote conventions for test results",
  hv_man_table_save_jtcvs = "Write a JTCVS-format table to Word",

  # ggRandomForests
  gg_error = "Out-of-bag error against the number of trees",
  gg_rfsrc = "Predicted response or survival from a forest",
  gg_survival = "Empirical survival estimate for comparison",
  gg_vimp = "Permutation variable importance",
  gg_shap = "Per-observation SHAP attributions",
  gg_variable = "Predicted response against a predictor's own values",
  gg_partial_rfsrc = "Partial dependence with other predictors averaged out",
  gg_roc = "ROC curve for a classification forest",
  gg_brier = "Brier score over time",
  gg_varpro = "varPro rule-based variable priority",
  gg_beta_varpro = "Beta-refined varPro importance",
  gg_beta_uvarpro = "Beta-refined importance for an unsupervised fit",
  gg_ivarpro = "Individual-level varPro priority",
  gg_isopro = "Isolation and anomaly scores",
  gg_udependent = "Dependency graph among predictors",
  gg_sdependent = "Ranked signal-variable detection",
  gg_partial_varpro = "Partial dependence for varPro-selected variables",
  gg_rhf = "Pointwise and cumulative hazard from a Random Hazard Forest",
  gg_auct = "Time-varying AUC for an RHF fit",
  gg_rhf_importance = "Time-localized variable priority for an RHF fit",
  gg_tune_rhf = "Read a completed RHF tree-size tuning path",

  # TemporalHazard
  hazard = "Fit an additive multi-phase hazard model",
  hzr_phase = "Define a hazard phase",
  hzr_phase_hazard = "Phase-specific hazard estimates",
  hzr_phase_cumhaz = "Phase-specific cumulative hazard",
  hzr_kaplan = "Kaplan-Meier comparison for a hazard fit",
  hzr_nelson = "Nelson-Aalen comparison for a hazard fit",
  hzr_calibrate = "Calibration of predicted against observed risk",
  hzr_gof = "Goodness-of-fit summary",
  hzr_deciles = "Observed against expected by risk decile",
  hzr_bootstrap = "Bootstrap uncertainty for a hazard fit",
  hzr_competing_risks = "Competing-risks decomposition",
  hzr_stepwise = "Stepwise variable selection for a hazard model",
  stepwise_trace = "Inspect the path a stepwise selection took",
  hzr_argument_mapping = "Map SAS HAZARD arguments onto the R interface",
  hzr_translate_sas = "Translate a SAS HAZARD call into R",
  hzr_read_outhaz = "Read a SAS OUTHAZ data set"
)
```

```{r}
#| label: build-map
## Scan the book's own sources. This chapter is excluded on purpose: it names
## every function in the contract above, so including it would mark everything
## covered by itself.
qmd <- setdiff(list.files(".", pattern = "[.]qmd$"), "function_map.qmd")
bodies <- setNames(
  lapply(qmd, function(f) paste(readLines(f, warn = FALSE), collapse = "\n")),
  sub("[.]qmd$", "", qmd)
)

chapters_calling <- function(fn) {
  pattern <- paste0("\\b", gsub("[.]", "[.]", fn), "\\(")
  names(bodies)[vapply(bodies, grepl, logical(1), pattern = pattern)]
}

map <- do.call(rbind, lapply(names(expected), function(pkg) {
  do.call(rbind, lapply(expected[[pkg]], function(fn) {
    where <- chapters_calling(fn)
    data.frame(
      Package  = pkg,
      Function = sprintf("`%s()`", fn),
      Role     = if (is.na(roles[fn])) "" else unname(roles[fn]),
      Recipe   = if (length(where))
                   paste(sprintf("[%s](%s.qmd)", where, where), collapse = ", ")
                 else "Package reference",
      Coverage = if (length(where)) "Worked recipe" else "Package reference",
      stringsAsFactors = FALSE
    )
  }))
}))
```

```{r}
#| label: coverage-headline
#| results: asis
cat(sprintf(
  "This book works through **%d of the %d** workflow-level functions in the contract.\n",
  sum(map$Coverage == "Worked recipe"), nrow(map)))
```

```{r}
#| label: function-map-tables
#| results: asis
for (pkg in names(expected)) {
  rows <- map[map$Package == pkg, c("Function", "Role", "Recipe", "Coverage")]
  cat("\n\n### ", pkg, "\n\n", sep = "")
  print(knitr::kable(rows, format = "pipe", row.names = FALSE))
}
```

## What the map does not cover

Lower-level helpers, S3 methods reached through `plot()` or `print()`, and
compatibility shims kept for older scripts are all deliberately absent. So are
the deprecated theme aliases: `theme_hv_manuscript()` and its siblings are the
current names, and the older `hv_theme_*()` and `theme_man()` forms still work
but should not be written into new code.

If you are looking for a function that is not here, it is either one of those or
it is new since this edition. `hvtiR::status()` will tell you which.
