# Hazard and nonparametric curves

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

The [survival chapter](survival.qmd) starts with patient-level follow-up and asks
what the observed cohort experienced. This chapter starts one step later: you
already hold a grid of fitted values and want to show what a model says between
the observed event times. That distinction matters. A smooth line is a model
summary, not a more precise Kaplan-Meier curve.

Name the quantity before you choose its column. Survival probability asks what
proportion remains event-free beyond a time. Cumulative hazard asks how much
event intensity has accumulated through that time. Hazard rate asks when events
are occurring fastest among patients still at risk. They can be derived from
one fitted model, but their scales, units, and clinical readings differ.

Reach for `hv_hazard()` when you have a parametric prediction grid and want
to show the smooth curve, often with the empirical Kaplan-Meier points overlaid
so a reviewer can judge how well the model fits. Reach for the nonparametric
temporal-trend curve (`hv_nonparametric()`) when your outcome is a prevalence or
a continuous measurement tracked over follow-up with a bootstrap confidence band
rather than a survival function. Reach for the ordinal curve (`hv_ordinal()`)
when the outcome has graded levels (regurgitation grade, severity class) and you
want one probability curve per grade. All three take pre-computed data and hand
back an S3 object. Inspect that object first, then call `plot()` to get the bare
ggplot you decorate. If the empirical hazard shape suggests distinct early,
constant, or late phases, continue to [fit an additive hazard
model](temporal_fit.qmd); these displays come before that modeling decision.

## The data it needs

`hv_hazard()` takes a parametric prediction grid: one row per time point, with
an estimate column (`survival`, `hazard`, or `cumhaz`) and explicit lower/upper
CI columns that you name in the call. Two helpers generate the demonstration
data. `sample_hazard_data()` builds the prediction grid; `sample_hazard_empirical()`
builds the binned Kaplan-Meier overlay (the discrete empirical estimate the
smooth curve is meant to track). Using the same cohort size for both keeps the
overlay aligned with the curve.

```{r}
#| label: hazard_data
dat_hp <- sample_hazard_data(n = 500, time_max = 10)
emp_hp <- sample_hazard_empirical(n = 500, time_max = 10, n_bins = 6)

haz_surv <- hv_hazard(
  dat_hp,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  empirical     = emp_hp,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
)
haz_surv$meta
head(haz_surv$data)
```

## Build it

The most common figure is a smooth parametric survival curve with its
confidence band plus discrete Kaplan-Meier points and error bars. We named the
estimate and CI columns when we built `haz_surv`; now `plot()` uses those stored
roles. The curve is on the percentage scale, so 80 means 80% surviving, not
0.80.

```{r}
#| label: fig-hazard
#| fig-cap: "Smooth parametric survival curve with its confidence band and overlaid binned Kaplan-Meier points with error bars"
plot(haz_surv) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival (%)") +
  theme_hv_manuscript()
```

## Read it

The point of the overlay is the comparison between the smooth curve and the
discrete points. Look for:

- **Points sitting on the curve.** If the empirical Kaplan-Meier estimates fall
  on or very near the parametric line, the model is describing the data well. If
  the points wander systematically off the curve (above it early, below it
  late), the parametric form is the wrong shape and you should reconsider it.
- **The error bars against the band.** The empirical error bars and the smooth
  confidence ribbon should tell a consistent story. Where the cohort thins, both
  widen; a point with a huge error bar is a region the figure cannot really
  support.
- **The curve shape itself.** A survival curve falls and flattens; a hazard
  curve often peaks early then settles; a cumulative hazard only rises. A shape
  that contradicts the biology is a signal to check the model before the figure.

## Variations

### Hazard rate

Switch `estimate_col` to `"hazard"` and the matching CI columns for the
instantaneous event rate. The sample stores $100h(t)$, so its units are events
per 100 patient-years. This is the curve to show when the question is *when*
events occur fastest rather than *how many* survive.

```{r}
#| label: fig-hazard-rate
#| fig-cap: "Instantaneous event rate per 100 patient-years, showing when post-operative events occur fastest among patients still at risk"
haz_rate <- hv_hazard(
  dat_hp,
  estimate_col = "hazard",
  lower_col    = "haz_lower",
  upper_col    = "haz_upper"
)

plot(haz_rate) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_fill_manual(values = c("firebrick"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 30)) +
  labs(x = "Years", y = "Events per 100 patient-years") +
  theme_hv_manuscript()
```

### Cumulative hazard

The `"cumhaz"` column equals $100\{-\log[S(t)]\}$ because these sample curves
use a per-100 scale. It rises monotonically, is not a percentage, and its local
slope is the event rate per 100 patient-years. Reach for cumulative hazard when
the question concerns accumulated event intensity rather than the surviving
fraction.

```{r}
#| label: fig-hazard-cumhaz
#| fig-cap: "Cumulative hazard per 100 patients, rising with a local slope equal to the event rate per 100 patient-years"
haz_cumhaz <- hv_hazard(
  dat_hp,
  estimate_col  = "cumhaz",
  lower_col     = "cumhaz_lower",
  upper_col     = "cumhaz_upper"
)

plot(haz_cumhaz) +
  scale_colour_manual(values = c("darkorange"), guide = "none") +
  scale_fill_manual(values = c("darkorange"), guide = "none") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(x = "Years", y = "Cumulative hazard (per 100)") +
  theme_hv_manuscript()
```

### Stratified by group

Pass `group_col` to compare two or more groups in one panel, each with its own
curve, band, and empirical overlay. Look for curves whose bands stop overlapping
in the windows where the treatment difference matters.

```{r}
#| label: fig-hazard-stratified
#| fig-cap: "Survival curves, bands, and empirical overlays for two groups in a single panel"
dat_strat <- sample_hazard_data(
  n = 400, time_max = 10,
  groups = c("No Takedown" = 1.0, "Takedown" = 0.65)
)
emp_strat <- sample_hazard_empirical(
  n = 400, time_max = 10, n_bins = 6,
  groups = c("No Takedown" = 1.0, "Takedown" = 0.65)
)

haz_strat <- hv_hazard(
  dat_strat,
  estimate_col  = "survival",
  lower_col     = "surv_lower",
  upper_col     = "surv_upper",
  group_col     = "group",
  empirical     = emp_strat,
  emp_lower_col = "lower",
  emp_upper_col = "upper"
)

plot(haz_strat) +
  scale_colour_manual(
    values = c("No Takedown" = "steelblue", "Takedown" = "firebrick"),
    name   = NULL
  ) +
  scale_fill_manual(
    values = c("No Takedown" = "steelblue", "Takedown" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years after Surgery", y = "Survival (%)") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

### Population life-table overlay

For age-stratified survival you often want to set the study curves against what
the general population would do. Pass a life-table data frame to `reference` and
set `ref_group_col` to draw the population survival as dashed lines per age
group. The gap between a study curve and its dashed reference is the excess
mortality attributable to the condition rather than to ageing.

```{r}
#| label: fig-hazard-lifetable
#| fig-cap: "Age-stratified study survival curves set against dashed population life-table references"
dat_age <- sample_hazard_data(
  n = 600, time_max = 12,
  groups = c("<65" = 0.5, "65–80" = 1.0, "≥80" = 1.8)
)
emp_age <- sample_hazard_empirical(
  n = 600, time_max = 12, n_bins = 6,
  groups = c("<65" = 0.5, "65–80" = 1.0, "≥80" = 1.8)
)
lt <- sample_life_table(
  age_groups = c("<65", "65–80", "≥80"),
  age_mids   = c(55, 72, 85),
  time_max   = 12
)

haz_age <- hv_hazard(
  dat_age,
  estimate_col     = "survival",
  lower_col        = "surv_lower",
  upper_col        = "surv_upper",
  group_col        = "group",
  empirical        = emp_age,
  emp_lower_col    = "lower",
  emp_upper_col    = "upper",
  reference        = lt,
  ref_estimate_col = "survival",
  ref_group_col    = "group"
)

plot(haz_age) +
  scale_colour_manual(
    values = c("<65" = "steelblue", "65–80" = "forestgreen",
               "≥80" = "firebrick"),
    name   = "Age group"
  ) +
  scale_fill_manual(
    values = c("<65" = "steelblue", "65–80" = "forestgreen",
               "≥80" = "firebrick"),
    guide  = "none"
  ) +
  scale_linetype_manual(
    values = c("<65" = "dashed", "65–80" = "dashed", "≥80" = "dashed"),
    name = "Population reference"
  ) +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Years", y = "Survival (%)",
       caption = "Dashed lines: US population life table") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

### Nonparametric temporal-trend curve

When the outcome is a prevalence or a continuous measurement tracked over
follow-up rather than a survival function, `hv_nonparametric()` prepares the
average curve with a bootstrap CI ribbon. `sample_nonparametric_curve_data()`
generates the average curve and CI; `sample_nonparametric_curve_points()`
generates the binned summary points that overlay it. Build the S3 object once
with the matching `lower_col`/`upper_col` and optional `data_points`, then
`plot()`.

```{r}
#| label: np_curve_data
curve_dat <- sample_nonparametric_curve_data(
  n            = 500,
  time_max     = 12,
  outcome_type = "probability",
  ci_level     = 0.68
)
pts_dat <- sample_nonparametric_curve_points(n = 500, time_max = 12)

np <- hv_nonparametric(
  curve_data  = curve_dat,
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = pts_dat
)
```

The bare panel shows the average curve, its CI ribbon, and the binned summary
points in default colours. Read it the same way you read the empirical overlay
in @fig-hazard: the curve should pass through the binned points, and the ribbon should
widen at the extremes where the bootstrap sample thins. A flat curve where you
expected a trend is a hint that the `outcome_type` or CI columns are
misspecified.

```{r}
#| label: np_curve_single_bare
plot(np)
```

The shaded ribbon here is the 68% bootstrap CI (one standard error); pass
`ci_level = 0.95` to `sample_nonparametric_curve_data()` for 95% bands.

```{r}
#| label: fig-hazard-nonparametric
#| fig-cap: "Nonparametric temporal-trend curve with its 68% bootstrap confidence ribbon and binned summary points"
plot(np) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values   = c("steelblue"), guide = "none") +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2),
                     labels = function(x) paste(x, "yr")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                     labels = scales::percent) +
  labs(x = "Follow-up (years)", y = "Prevalence (%)") +
  theme_hv_manuscript()
```

Pass `group_col` to the constructor to compare two average curves in a single
panel; each group gets its own CI ribbon.

```{r}
#| label: fig-hazard-nonparametric-groups
#| fig-cap: "Two nonparametric average curves compared in one panel, each with its own bootstrap confidence ribbon"
curve_grp <- sample_nonparametric_curve_data(
  n            = 400,
  time_max     = 7,
  groups       = c("Ozaki" = 0.7, "CE-Pericardial" = 1.3),
  outcome_type = "continuous",
  ci_level     = 0.68
)
pts_grp <- sample_nonparametric_curve_points(
  n = 400, time_max = 7,
  groups = c("Ozaki" = 0.7, "CE-Pericardial" = 1.3),
  outcome_type = "continuous"
)
np_grp <- hv_nonparametric(
  curve_data  = curve_grp,
  group_col   = "group",
  lower_col   = "lower",
  upper_col   = "upper",
  data_points = pts_grp
)

p_haz_grp <- plot(np_grp) +
  scale_colour_manual(
    values = c("Ozaki" = "steelblue", "CE-Pericardial" = "firebrick"),
    name   = NULL
  ) +
  scale_fill_manual(
    values = c("Ozaki" = "steelblue", "CE-Pericardial" = "firebrick"),
    guide  = "none"
  ) +
  scale_x_continuous(limits = c(0, 7), breaks = 0:7) +
  labs(x = "Follow-up (years)", y = "AV Peak Gradient (mmHg)") +
  theme_hv_manuscript()

# Rising curves leave the upper-left empty; pin the key there.
hv_legend_inside(p_haz_grp, prefer = "topleft")
```

### Nonparametric ordinal-outcome curve

When the outcome has graded levels, `hv_ordinal()` prepares one probability
curve per grade from a cumulative proportional-odds model (the prevalence of
each regurgitation grade over time, say). The curve data is long format, one row
per time by grade. `sample_nonparametric_ordinal_data()` generates the curves
and `sample_nonparametric_ordinal_points()` the binned summary points.

```{r}
#| label: np_ordinal_data
ord_dat <- sample_nonparametric_ordinal_data(
  n = 800, time_max = 5,
  grade_labels = c("None", "Mild", "Moderate", "Severe")
)
ord_pts <- sample_nonparametric_ordinal_points(
  n = 800, time_max = 5,
  grade_labels = c("None", "Mild", "Moderate", "Severe")
)
head(ord_dat)
```

Build the object with both curve data and summary points. Each line is one grade
level; because every patient has exactly one grade at each time, the lines sum
to roughly 1.0 at every time point. Read the figure as a stacked story: the
`"None"` line starts high and declines as patients drift into worse grades, and
the severe line creeps up. Colours run from grey for None through graduated
severity to firebrick for Severe so the eye reads worsening as warming.

```{r}
#| label: fig-hazard-ordinal
#| fig-cap: "One probability curve per regurgitation grade over time, coloured from grey to firebrick as severity worsens"
np_ord <- hv_ordinal(curve_data = ord_dat, data_points = ord_pts)
np_ord$meta

plot(np_ord) +
  scale_colour_manual(
    values = c("None" = "grey40", "Mild" = "steelblue",
               "Moderate" = "darkorange", "Severe" = "firebrick"),
    name = "AR Grade"
  ) +
  scale_x_continuous(limits = c(0, 5), breaks = 0:5) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                     labels = scales::percent) +
  labs(x = "Follow-up (years)", y = "Grade prevalence (%)") +
  theme_hv_manuscript() +
  theme(legend.position = c(0.75, 0.6))
```

To collapse adjacent grades, add their probabilities at each time before
passing the data to the constructor. Do not simply drop the middle rows; then
the displayed levels no longer account for every patient and will not sum to
one. Here we combine None with Mild and Moderate with Severe.

```{r}
#| label: fig-hazard-ordinal-collapsed
#| fig-cap: "Two-level ordinal version after summing adjacent grade probabilities; the two curves account for the full cohort at each time"
ord_two <- transform(
  ord_dat,
  grade = ifelse(grade %in% c("None", "Mild"),
                 "None or Mild", "Moderate or Severe")
)
ord_two <- aggregate(estimate ~ time + grade, data = ord_two, FUN = sum)

np_ord_two <- hv_ordinal(curve_data = ord_two)

plot(np_ord_two) +
  scale_colour_manual(
    values = c("None or Mild" = "steelblue",
               "Moderate or Severe" = "firebrick"),
    name   = NULL
  ) +
  scale_x_continuous(limits = c(0, 5), breaks = 0:5) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                     labels = scales::percent) +
  labs(x = "Follow-up (years)", y = "Grade prevalence (%)") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

## Pitfalls

- **Mismatched CI columns.** Every estimate column has its own pair of CI
  columns. Plot `hazard` with the survival bounds and the band will be nonsense.
  Name the matching `lower_col`/`upper_col` for the column you are plotting.
- **Forgetting the empirical overlay.** A smooth parametric curve always looks
  convincing. Without the Kaplan-Meier points a reviewer cannot tell whether the
  model fits, so include `empirical` whenever you have it.
- **Ordinal lines that do not sum to one.** If your grade curves drift above 1.0
  or cross oddly, the wide-format grade columns were probably reshaped to long
  format incorrectly. Each time point's grades should add to about 1.0.
- **Over-reading the CI level.** The nonparametric ribbon defaults to a 68% band
  (one standard error), which is narrower than the 95% band readers expect.
  State the level in the caption, or switch to `ci_level = 0.95`.
