# Histogram plots

How well do treatment groups overlap after adjustment, and how does the mix of
a cohort change over time? Histograms answer both questions when the bin counts
and their denominator are made explicit.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

Two very different questions both end in a histogram. The first is a
propensity-matching diagnostic: when we compare two treatment groups that were
not randomized, we adjust for confounding by matching or weighting on a
propensity score (the modelled probability of receiving the treatment). The
mirrored histogram is how we show a reviewer that the adjustment worked, by
putting one group's score distribution above the axis and the other's below it
and letting the eye judge the overlap. The second question is compositional: how
the make-up of a cohort shifts year over year. The stacked histogram answers
that one, with each bar split into the categories that compose it.

This chapter covers both. The mirrored histogram (`hv_mirror_hist()`) is the
diagnostic; the stacked histogram (`hv_stacked()`) is the trend. Both come from
hvtiPlotR [@R-hvtiPlotR] and follow the package pattern: a constructor prepares
the data, `plot()` hands back a bare ggplot, and you dress it with colour,
labels, and a theme.

## The data it needs

`hv_mirror_hist()` accepts a data frame with a continuous propensity score, a
binary group indicator, and a third column that controls the overlay. Two
display modes follow from which third column you supply. In **binary-match
mode** you pass `match_col`: the full bars show every observation before
matching and a darker overlay shows the matched subset. In **weighted IPTW
mode** (inverse probability of treatment weighting, the alternative to 1:1
matching) you pass `weight_col`: the overlay height is the sum of weights in each
bin rather than a matched count.

`sample_mirror_histogram_data()` simulates 4,000 patients (2,000 per group) with a continuous
propensity score (`prob_t`), a binary group indicator (`tavr`), and a match flag
(`match`). `score_multiplier = 100` puts the score on a 0 to 100 percent scale
and `binwidth = 5` sets 5-point bins.

```{r}
#| label: mirror_histogram_build
mirror_dta <- sample_mirror_histogram_data(n = 2000, separation = 1.5)

mh <- hv_mirror_hist(
  data             = mirror_dta,
  score_col        = "prob_t",
  group_col        = "tavr",
  match_col        = "match",
  group_levels     = c(0, 1),
  group_labels     = c("SAVR", "TF-TAVR"),
  matched_value    = 1,
  score_multiplier = 100,
  binwidth         = 5
)
```

## Inspect it

Before plotting, check what the constructor kept. The class selects the plot
method, `meta` records the columns and five-point bin width, and `tables` carries
the balance diagnostics we will report with the figure.

```{r}
#| label: mirror_histogram_inspect
class(mh)
mh$meta[c("score_col", "group_col", "binwidth", "n_obs", "n_dropped")]
mh$tables$diagnostics[c("group_counts_before", "group_counts_matched",
                        "smd_before", "smd_matched")]
```

## Build it

The bare panel shows two mirrored bar charts, upper bars for the first group and
lower for the second, with white fill and no scale, label, or theme yet.

```{r}
#| label: mirror_histogram_bare
p <- plot(mh, alpha = 0.8)
p
```

Now dress it. `scale_fill_manual()` maps the four internal fill levels
(`before_g0`, `matched_g0`, `before_g1`, `matched_g1`) to white for the
pre-match bars and two greens for the matched subsets. The annotations use
`y = Inf`/`-Inf` with `vjust` to pin the group labels at the panel edges no
matter the count scale, and `scale_y_continuous(labels = abs)` turns the
internal negative counts of the lower group back into positive numbers. Put each
label in the corner the bars leave empty — the upper group thins out to the
right, so `SAVR` sits top-right, while the lower group thins to the left, so
`TF-TAVR` sits bottom-left. An in-panel label belongs in dead space, never on top
of the data it names.
`theme_hv_manuscript()` sizes the text for a journal column; swap in
`theme_hv_poster()` for a conference poster.

```{r}
#| label: fig-histograms
#| fig-cap: "Mirrored matching histogram of propensity scores for SAVR and TF-TAVR, with the matched subset overlaid"
p +
  ggplot2::scale_fill_manual(
    values = c(
      before_g0  = "white",  matched_g0 = "green1",
      before_g1  = "white",  matched_g1 = "green4"
    ),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10)
  ) +
  ggplot2::scale_y_continuous(labels = abs) +
  ggplot2::annotate("text", x = 85, y =  Inf, vjust =  2,
                    label = mh$meta$group_labels[1], size = 7) +
  ggplot2::annotate("text", x = 20, y = -Inf, vjust = -1,
                    label = mh$meta$group_labels[2], size = 7) +
  ggplot2::labs(x = "Propensity Score (%)", y = "Number of Patients") +
  theme_hv_manuscript()
```

## Read it

A mirrored matching histogram is read as a before-and-after, with the two halves
of the figure carrying different jobs. Look for:

- **Symmetric pre-match bars.** The light bars are the full distributions. Two
  groups that need adjusting will sit at different ends of the score axis: the
  SAVR patients piled toward low scores, the TF-TAVR patients toward high. That
  separation is exactly the confounding the matching is meant to fix.
- **The darker overlay narrowing in.** The matched bars should occupy the region
  of score where the two groups overlap, and the upper and lower matched
  distributions should look like mirror images. That symmetry is the visual
  evidence that matching balanced the groups.
- **A region with no overlay.** Where one group has light bars but no dark ones,
  those patients had no match and were dropped. A lot of unmatched patients at
  the extremes is the cost of matching, and worth noting.
- **Identical-looking panels before and after.** If the matched overlay looks
  the same as the full bars, the `match_col` may not be mapping correctly.

The numbers behind the picture are stored on the object. The constructor records
group counts and standardized mean differences (SMD, a scale-free measure of
group imbalance) before and after matching:

```{r}
#| label: mirror_histogram_diagnostics
mh$tables$diagnostics$smd_before
mh$tables$diagnostics$smd_matched
mh$tables$diagnostics$group_counts_before
mh$tables$diagnostics$group_counts_matched
```

An SMD under about 0.1 after matching is the usual rule of thumb for adequate
balance. Quote the matched SMD next to the figure so the reader does not have to
take the symmetry on trust.

The x-axis is propensity score in percent and the y-axis is the number of
patients in each five-point bin, with SAVR above zero and TF-TAVR below. Compare
the dark matched overlay with the light full-cohort bars, then confirm the visual
comparison with the SMD. This is a balance diagnostic. It does not estimate a
treatment effect or establish that treatment caused an outcome difference.

## Adapt it

### Weighted IPTW mode (Limited vs. Extended)

When the analysis uses inverse probability of treatment weighting rather than
1:1 matching, pass `weight_col` instead of `match_col`. Each bin's overlay
height is now the sum of IPTW weights in that bin. `add_weights = TRUE` in the
sample generator attaches an `mt_wt` column.

```{r}
#| label: mirror_histogram_weighted_build
wt_dta <- sample_mirror_histogram_data(
  n = 2000, separation = 1.5, add_weights = TRUE
)

mh_wt <- hv_mirror_hist(
  data             = wt_dta,
  score_col        = "prob_t",
  group_col        = "tavr",
  group_levels     = c(0, 1),
  group_labels     = c("Limited", "Extended"),
  weight_col       = "mt_wt",
  score_multiplier = 100,
  binwidth         = 5
)

mh_wt$meta[c("score_col", "group_col", "binwidth", "n_obs", "n_dropped")]
weight_total_by_group <- mh_wt$tables$diagnostics$effective_n_by_group
weight_total_by_group
mh_wt$tables$diagnostics[c("smd_before", "smd_weighted")]
```

Despite the component name in hvtiPlotR 2.7.10,
`effective_n_by_group` stores the **sum of weights** in each group. It is not the
Kish effective sample size, $(\sum w)^2 / \sum w^2$. We give it the accurate
local name `weight_total_by_group`; in this normalized example the totals are
2,000 and 2,000, the same as the raw group counts.

The bare weighted panel looks like the binary-match bare plot, but the overlay
encodes IPTW weight sums, not counts. Read it the same way: balanced upper and
lower overlay bars mean good weighting, while bars that stay heavily one-sided
point to extreme weights you should investigate.

```{r}
#| label: mirror_histogram_weighted_bare
p_wt <- plot(mh_wt, alpha = 0.8)
p_wt
```

The IPTW variant uses blue and red for the Limited and Extended groups, with the
group labels coloured to match. The axis and annotation pattern is the same as
the binary-match version; swap in `theme_hv_poster()` when preparing the figure
for a conference poster.

```{r}
#| label: fig-histograms-weighted
#| fig-cap: "Mirrored histogram in weighted IPTW mode, with overlay heights showing the sum of inverse-probability weights for the Limited and Extended groups"
p_wt +
  ggplot2::scale_fill_manual(
    values = c(
      before_g0   = "white", weighted_g0 = "blue",
      before_g1   = "white", weighted_g1 = "red"
    ),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10)
  ) +
  ggplot2::scale_y_continuous(labels = abs) +
  ggplot2::annotate("text", x = 85, y =  Inf, vjust =  2,
                    label = mh_wt$meta$group_labels[1], color = "blue", size = 5) +
  ggplot2::annotate("text", x = 15, y = -Inf, vjust = -1,
                    label = mh_wt$meta$group_labels[2], color = "red",  size = 5) +
  ggplot2::labs(x = "Propensity Score (%)", y = "#") +
  theme_hv_manuscript()
```

The weight totals describe the scale of the weighted histogram, while the
weighted SMD evaluates balance:

```{r}
#| label: mirror_histogram_weighted_diagnostics
mh_wt$tables$diagnostics$smd_weighted
weight_total_by_group
```

Here the x-axis remains propensity score, but the overlay height is the sum of
IPTW weights within a bin, not a patient count. Compare the weighted upper and
lower shapes and report the group weight totals with the weighted SMD. In this
frozen example the weighted SMD is 0.543, still well above the 0.1 rule of thumb,
so the weighting is inadequate; the figure is an example of diagnosing failed
balance, not declaring success. Even adequate balance on the score would not by
itself identify a causal treatment effect.

### Stacked histogram: counts over time

The stacked histogram answers the compositional question: how the mix of a
categorical variable shifts over time. `hv_stacked()` prepares the data and
`plot()` hands back a bare ggplot. `sample_stacked_histogram_data()` generates a
reproducible dataset with `year` and `category` columns.

```{r}
#| label: stacked_histogram_data
hist_dta <- sample_stacked_histogram_data(n_years = 20, start_year = 2000,
                                           n_categories = 3)
head(hist_dta)
```

The default `position = "stack"` shows raw counts within each bin, so the height
of a whole bar is the total volume in that year and each coloured segment is one
category's contribution. Use this when both the total and the mix matter: a
growing bar with a shifting palette tells you the cohort is both larger and
differently composed.

```{r}
#| label: fig-histograms-stacked-count
#| fig-cap: "Stacked histogram of annual counts, with each bar split into category contributions"
sh <- hv_stacked(hist_dta, x_col = "year", group_col = "category")
class(sh)
sh$meta[c("x_col", "group_col", "binwidth", "position", "n_obs", "n_groups")]
p_count <- plot(sh)

p_count +
  scale_fill_brewer(
    palette = "Set1",
    labels = c("1" = "Group A", "2" = "Group B", "3" = "Group C"),
    name = "Category"
  ) +
  scale_color_brewer(palette = "Set1", guide = "none") +
  labs(x = "Year", y = "Count") +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

Year is on the x-axis and the y-axis is the number of observations in that
year; each coloured segment contributes to the annual total. Compare both whole
bar heights and the category segments, using all records in a year as the count
denominator. The display describes cohort volume and composition. It cannot say
why either changed.

### Stacked histogram: proportions over time

`position = "fill"` rescales each bin so the bars sum to 1. Every bar is now full
height and only the split between colours carries information, which makes the
relative composition comparable across years without the total volume getting in
the way. Reach for the fill version when the share is the story and the count is
a distraction.

```{r}
#| label: fig-histograms-stacked-fill
#| fig-cap: "Stacked histogram rescaled to proportions, showing category composition over time independent of cohort volume"
sh2 <- hv_stacked(hist_dta, x_col = "year", group_col = "category",
                  position = "fill")
sh2$meta[c("x_col", "group_col", "binwidth", "position", "n_obs", "n_groups")]
p_fill <- plot(sh2)

p_fill +
  scale_fill_manual(
    values = c("1" = "pink", "2" = "cyan", "3" = "orangered"),
    labels = c("1" = "Group A", "2" = "Group B", "3" = "Group C"),
    name   = "Category"
  ) +
  scale_color_manual(
    values = c("1" = "pink", "2" = "cyan", "3" = "orangered"),
    guide  = "none"
  ) +
  labs(x = "Year", y = "Proportion") +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

The proportion version keeps year on the x-axis and divides every segment by
that year's total, so each bar sums to one. Compare category shares across years,
but do not use this panel to compare annual volume or to attribute a change in
case mix to an intervention.

## Deliver it

Once the denominator and balance statistics are in the caption, take the panel
to [Finish and deliver the result](publications.qmd) for the manuscript, poster,
or slide export path. Keep the matched or weighted SMD with the mirrored figure;
the picture is not a substitute for that number.

## Pitfalls

- **Reading balance off the picture alone.** A mirrored histogram that looks
  symmetric is reassuring, but the SMD is the number a reviewer trusts. Always
  report the matched or weighted SMD alongside the figure.
- **Forgetting `labels = abs` on the y-axis.** The lower group's counts are
  stored negative so the bars mirror. Without `scale_y_continuous(labels = abs)`
  the axis shows negative patient counts, which confuses every reader.
- **Mapping the wrong overlay column.** Pass `match_col` for matching and
  `weight_col` for weighting, not both. The overlay means something different in
  each mode, and a count overlay on a weighted analysis (or vice versa) is
  simply wrong.
- **Confusing count and fill stacked bars.** A fill histogram hides volume: a
  category can hold a steady share while its raw count collapses. If the absolute
  numbers matter, show the count version too, or you will mislead a reader about
  how much data sits behind a thin year.
