# Manuscript tables

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(hvtiRtables)
library(gtsummary)
library(flextable)
```

The [publication tables](qt_tables.qmd) chapter built a Table 1 by hand with
`gt`, which is the right tool when the table's destination is this book or
another HTML document. This chapter covers the other destination. When the table
is going into a manuscript, it has to land in a Word file that a coordinator can
paste into the submission draft, and it has to obey the CORR "Table Construction
for Manuscripts" rules once it gets there. `hvtiRtables` is the package that
takes a `gtsummary` summary the rest of the way.

## When to use it

Most of the rules in that memo come down to one idea: everything a reader sees
in the table should be *in* the table. A header that spans two columns by
merging their cells, a hidden quarter-inch spacer column doing the alignment, a
caption floating in a text box positioned over the grid, a footnote living as an
extra row inside the table's own frame — each of these puts content in a layer
that the table structure itself does not account for. They look fine on your
screen. They come apart when the manuscript is re-flowed into the journal's
template, and they are invisible to a screen reader.

So the house style asks for a flat table: one header row, no merged cells, and
footnotes and the abbreviation key written as ordinary paragraphs below the
table rather than tucked inside it. That is a fussy thing to build by hand in
Word, and a fussy thing to keep correct across the six revisions a table goes
through. Reach for `hvtiRtables` whenever the table is manuscript-bound, and let
the package hold the rules.

There are two house formats, and the difference between them is exactly the
merge question. `hv_man_table()` builds the flat CORR style described above.
`hv_man_table_jtcvs()` builds what JTCVS editorial asks for at submission, which
is the opposite: a two-row header with each group name spanning its own
sub-columns, and lettered footnotes attached to specific cells. Same numbers,
two different shapes, chosen by where the table is going.

## The data it needs

Any data frame `gtsummary::tbl_summary()` accepts. We use
`generate_survival_data()` again, the same simulated cohort the
[Table 1](qt_tables.qmd) and [dataset description](data_tables.qmd) chapters
work from, so you can read the three against one dataset.

```{r}
#| label: hv-data
dta <- hvtiRutilities::generate_survival_data(n = 200, seed = 42)
table(dta$sex)
```

## Build it

### Summarise, then convert — `hv_man_table()`

You build the summary with `gtsummary` the way you normally would. The one thing
worth setting deliberately is `statistic`, because the house convention for a
continuous variable is a median with the 15th and 85th percentiles rather than a
mean and SD. That convention is what the default footnotes describe, so the two
need to agree.

```{r}
#| label: hv-summary
tbl <- tbl_summary(
  dta,
  by        = sex,
  include   = c(age, bmi, gfr_bs, diabetes),
  statistic = list(all_continuous() ~ "{median} ({p15}, {p85})")
)
```

`hv_man_table()` converts that into a `flextable` carrying the house rules: a
single non-merged header row, Times New Roman, and the house font size.

```{r}
#| label: hv-flextable
ft <- hv_man_table(tbl)
ft
```

Font size defaults to 12 points; pass `font_size = 11` for a table wide enough
that 12 will not fit the page. Rounding stays where you set it, in
`tbl_summary()`'s own `statistic` and `digits` arguments, rather than being
re-specified here.

The structural check is easy to run on the object itself. A compliant table has
exactly one header row and no cell spanning more than one column.

```{r}
#| label: hv-flat-check
nrow(ft$header$dataset)              # header rows
max(as.vector(ft$header$spans$rows)) # widest header span
max(as.vector(ft$body$spans$rows))   # widest body span
```

### Write the Word file — `hv_man_table_save()`

`hv_man_table_save()` writes the `.docx`. Footnotes default to
`hv_man_footnotes()`, the two notes that are universal to the house rules rather
than specific to any one study.

```{r}
#| label: hv-footnotes
hv_man_footnotes()
```

The `*` marks the non-missing-value count in the wording the rules require, and
the `†` explains the median-and-percentiles format. Each symbol is appended as a
superscript reference mark on the table's header, and the text itself is written
as a real paragraph after the table. That is the part `flextable::footnote()`
cannot do for you: it renders footnote text as an extra row inside the table's
own frame, which is the violation the whole exercise is meant to avoid.

The abbreviation key works the same way, as a `Key:` paragraph below the
footnotes, sorted alphabetically.

```{r}
#| label: hv-save
#| eval: false
hv_man_table_save(
  ft,
  file = "table1.docx",
  abbreviations = c(
    BMI  = "body mass index",
    eGFR = "estimated glomerular filtration rate"
  )
)
```

Override the footnotes with ordinary list operations. `footnotes = NULL`
suppresses both, ``modifyList(hv_man_footnotes(), list(`†` = "..."))`` replaces
one, and ``c(hv_man_footnotes(), list(`‡` = "..."))`` adds a study-specific note
alongside the defaults.

### Check what you wrote — `hv_check_docx()`

Writing the file and trusting it are different things, and the failure modes
here are structural rather than visual. `hv_check_docx()` reads a `.docx` and
reports the three patterns the rules name: text positioned in a floating layer
over the table, a hidden spacer column, and a footnote embedded as a row inside
the table frame. It returns one row per finding, so zero rows means clean.

```{r}
#| label: hv-check
out <- tempfile(fileext = ".docx")
hv_man_table_save(ft, out, abbreviations = c(BMI = "body mass index"))
hv_check_docx(out)
```

It never modifies the file, and it does not care who wrote it, so it also works
on a table a co-author sends back after editing it in Word. That is the case
worth remembering. Your own output is compliant by construction; the round trip
through somebody else's copy of Word is where a merged cell or a text box
actually gets introduced.

Two of the three detectors are structural and reliable. The hidden-column test
deliberately requires a column to be *both* completely empty and narrower than a
tenth of an inch, because the house JTCVS templates use empty half-inch gutter
columns on purpose and either condition alone would flag them. The
embedded-footnote detector is a pattern match on the last row of a table, so
treat what it reports as a prompt to look rather than proof.

## Variations

### JTCVS submissions

The JTCVS path starts somewhere different. `hv_tbl_summary()` is a wrapper over
`gtsummary::tbl_summary()` written to the interface the biostats team already
knows from the `%summarytable` SAS macro: you give it a grouped, ordered
variable list and sort the variables into type buckets, instead of using
`gtsummary`'s tidyselect arguments.

```{r}
#| label: hv-jtcvs-summary
tbl_j <- hv_tbl_summary(
  dta,
  by     = "sex",
  groups = list(
    "Demographics" = c("age", "bmi"),
    "Comorbidity"  = c("gfr_bs", "diabetes")
  ),
  continuous  = c("age", "bmi", "gfr_bs"),
  categorical = c("diabetes"),
  compare     = "pvalue"
)
```

`groups` is the macro's `LIST=`, and it does double duty: it fixes the display
order and it supplies the section headers that break the table into blocks.
`continuous`, `binary`, and `categorical` are `CON1=`, `CAT1=`, and `CAT2=`.
Every variable named in `groups` has to appear in exactly one bucket and every
bucketed variable has to appear in `groups`, which is a strict rule but a
forgiving one to debug, since the error names the variables that fell through.
Continuous variables are summarised as median with the percentile pair from
`percentiles` (15th and 85th by default) and tested non-parametrically
throughout, so there is no per-variable normality decision to make or defend.

`compare` adds the trailing comparison column: `"pvalue"`, `"smd"`, `"both"`, or
`"none"`. The result carries two attributes that the renderer wants, and you
pass them straight through.

```{r}
#| label: hv-jtcvs-attrs
attr(tbl_j, "hv_stat_label")
attr(tbl_j, "hv_trailing")
```

`hv_man_table_jtcvs()` then builds the merged shape. Its own `groups` argument
is a different thing entirely from `hv_tbl_summary()`'s: here it maps each
`stat_` column to its spanning header label, and you write the group's N into
the label text yourself.

```{r}
#| label: hv-jtcvs-table
n_sex <- table(dta$sex)

ft_j <- hv_man_table_jtcvs(
  tbl_j,
  groups = c(
    stat_1 = sprintf("Female (n=%d)", n_sex[["Female"]]),
    stat_2 = sprintf("Male (n=%d)",   n_sex[["Male"]])
  ),
  trailing   = attr(tbl_j, "hv_trailing"),
  stat_label = attr(tbl_j, "hv_stat_label")
)
ft_j
```

Compare the structure with the flat table above and the two modes separate
cleanly: two header rows here instead of one, spanning header cells, and
full-width shaded section rows in the body.

```{r}
#| label: hv-jtcvs-structure
nrow(ft_j$header$dataset)
max(as.vector(ft_j$header$spans$rows))
max(as.vector(ft_j$body$spans$rows))
```

JTCVS marks each p-value with a superscript letter naming the test behind it,
which is the same thing `%summarytable` does. `hv_test_footnotes_jtcvs()` reads
the test `gtsummary` recorded for each row and builds those footnotes, including
the body-row indices, so you are not counting rows by hand.

```{r}
#| label: hv-jtcvs-footnotes
fn <- hv_test_footnotes_jtcvs(tbl_j)
fn
```

Letters follow a fixed order rather than order of appearance, so two tables in
one manuscript that use the same tests get the same letters. Pass the result
to `hv_man_table_save_jtcvs()`, which adds the bold caption paragraph above the
table and renders the lettered notes below it.

```{r}
#| label: hv-jtcvs-save
#| eval: false
hv_man_table_save_jtcvs(
  ft_j,
  file      = "table1_jtcvs.docx",
  caption   = "Table 1. Baseline Characteristics",
  footnotes = fn,
  abbreviations = c(eGFR = "estimated glomerular filtration rate")
)
```

The caption is not auto-numbered; write the number into the string yourself.

## Read it

The two chunks worth reading closely in either format are the header and the
denominators.

In the flat table, the header carries each group's N, so the column tells you
its own denominator without a trip to the caption. Read it before you read any
percentage in the column beneath it. The `*` footnote is doing real work here,
because the count it describes is the count of *non-missing* values, which is
not always the group N. When a variable has missing data, those two numbers
separate, and the footnote is what tells a reader which one they are looking at.

In the JTCVS table, the N sits in its own sub-column beside each group's
statistic, one pair per group. That layout exists so a reader can see per-variable
availability at a glance rather than inferring it, which is the same information
the flat table's footnote conveys in prose.

Then the section headers. Both formats break the variable list into labelled
blocks, and those blocks are an editorial claim about what belongs with what. A
reader scanning a Table 1 reads the block labels before the rows. Get them
wrong and a comorbidity reads as a demographic.

## Pitfalls

- **`hv_tbl_summary()` output goes to `hv_man_table_jtcvs()`, not
  `hv_man_table()`.** This is the one that will actually bite you, because it
  fails silently rather than erroring. `hv_tbl_summary()` packs each cell as
  `"{N_obs} ||| {stat}"`, and the `|||` is a sentinel that
  `hv_man_table_jtcvs()` splits on to build its paired N and statistic
  sub-columns. `hv_man_table()` does not split on it, so the cells render
  literally as `77 ||| 46 (30, 62)` and you get a table that looks like a bug in
  the package. Build the flat CORR table from a plain `tbl_summary()`, as the
  first recipe above does, and keep `hv_tbl_summary()` for the JTCVS path.

- **`groups` means two different things.** In `hv_tbl_summary()` it is a named
  *list* of section label to variable names, the row structure. In
  `hv_man_table_jtcvs()` it is a named *character vector* of `stat_` column to
  spanning header label, the column structure. They sit two calls apart in the
  same pipeline and neither will accept the other's value, so the error you get
  is at least immediate.

- **A standardised mean difference needs exactly two groups.** `compare = "smd"`
  and `compare = "both"` are only defined between two groups and error
  otherwise. The version of this that costs an afternoon is a `by` variable that
  is a factor carrying a level nothing in the data has — after a cohort filter,
  say. It looks like two groups when you `table()` it and counts as three here.
  `droplevels()` on the filtered frame is the fix. `compare = "pvalue"` works at
  any number of groups.

- **Footnote row indices count the section-header rows.** When you write
  footnotes for `hv_man_table_save_jtcvs()` by hand, `row` indexes the rendered
  body rows *including* the section headers the renderer interleaves, not the
  rows of your original summary. Count off the printed table, not the data
  frame. For test footnotes, let `hv_test_footnotes_jtcvs()` compute the indices,
  which is most of why it exists.

- **The check is not automatic.** `hv_man_table_save()` and
  `hv_man_table_save_jtcvs()` write compliant output, but neither one inspects
  the file afterwards. If you want the check, run `hv_check_docx()` on the path
  yourself. It is cheap, and the moment to do it is after the table has been
  through anyone else's hands.
