# Manuscript tables

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(hvtiRtables)
library(gtsummary)
library(flextable)
```

The [publication tables](qt_tables.qmd) chapter built a Table 1 by hand with
`gt`. This chapter takes the same summary to a different destination. When the
table is going into a manuscript, it has to land in a Word file that a
coordinator can paste into the submission draft, and it has to obey the CORR
"Table Construction for Manuscripts" rules once it gets there. `hvtiRtables` is
the package that takes a `gtsummary` summary the rest of the way.

`gt` writes Word too, so the obvious shortcut is `gtsave()`, and it is a fair
thing to ask about before installing another package. We take that question
seriously below, after the recipe, and answer it by saving one summary both ways
and unzipping the two files.

## When to use it

Most of the rules in that memo come down to one idea: everything a reader sees
in the table should be *in* the table. A header that spans two columns by
merging their cells, a hidden quarter-inch spacer column doing the alignment, a
caption floating in a text box positioned over the grid, a footnote living as an
extra row inside the table's own frame — each of these puts content in a layer
that the table structure itself does not account for. They look fine on your
screen. They come apart when the manuscript is re-flowed into the journal's
template, and they are invisible to a screen reader.

So the house style asks for a flat table: one header row, no merged cells, and
footnotes and the abbreviation key written as ordinary paragraphs below the
table rather than tucked inside it. That is a fussy thing to build by hand in
Word, and a fussy thing to keep correct across the six revisions a table goes
through. Reach for `hvtiRtables` whenever the table is manuscript-bound, and let
the package hold the rules.

There are two house formats, and the difference between them is exactly the
merge question. `hv_man_table()` builds the flat CORR style described above.
`hv_man_table_jtcvs()` builds what JTCVS editorial asks for at submission, which
is the opposite: a two-row header with each group name spanning its own
sub-columns, and lettered footnotes attached to specific cells. Same numbers,
two different shapes, chosen by where the table is going.

## The data it needs

Any data frame `gtsummary::tbl_summary()` accepts. We use
`generate_survival_data()` again, the same simulated cohort the
[Table 1](qt_tables.qmd) and [dataset description](data_tables.qmd) chapters
work from, so you can read the three against one dataset.

```{r}
#| label: hv-data
dta <- hvtiRutilities::generate_survival_data(n = 200, seed = 42)
table(dta$sex)
```

## Build it

### Summarise, then convert — `hv_man_table()`

You build the summary with `gtsummary` the way you normally would. Two things
are worth setting deliberately. The first is `statistic`, because the house
convention for a continuous variable is a median with the 15th and 85th
percentiles rather than a mean and SD, and that convention is what the default
footnotes describe. The second is less obvious: `gtsummary` writes its own
footnote describing the same statistics, and marks the group headers with a
superscript `1` pointing at it. Leave it in and the finished document carries two
footnote systems saying the same thing in two different marker schemes, one of
which lands as an extra row inside the table's own frame. Drop it, and the house
footnotes are the only ones in the file.

```{r}
#| label: hv-summary
tbl <- tbl_summary(
  dta,
  by        = sex,
  include   = c(age, bmi, gfr_bs, diabetes),
  statistic = list(all_continuous() ~ "{median} ({p15}, {p85})")
) |>
  remove_footnote_header() |>
  remove_footnote_body()
```

`hv_man_table()` converts that into a `flextable` carrying the house rules: a
single non-merged header row, Times New Roman, and the house font size.

```{r}
#| label: hv-flextable
ft <- hv_man_table(tbl)
ft
```

Font size defaults to 12 points; pass `font_size = 11` for a table wide enough
that 12 will not fit the page. Rounding stays where you set it, in
`tbl_summary()`'s own `statistic` and `digits` arguments, rather than being
re-specified here.

The structural check is easy to run on the object itself. A compliant table has
exactly one header row and no cell spanning more than one column.

```{r}
#| label: hv-flat-check
shape <- function(x) {
  c(header_rows = nrow(x$header$dataset),
    header_span = max(as.vector(x$header$spans$rows)),
    body_span   = max(as.vector(x$body$spans$rows)))
}
rbind(hv_man_table  = shape(ft),
      as_flex_table = shape(as_flex_table(tbl)))
```

Read that as a check on your own table rather than as a scoreboard. Plain
`gtsummary::as_flex_table()` reports the same three numbers on the same summary,
because a `by`-grouped summary has nothing in it that wants merging in the first
place. The numbers earn their keep further down, as the contrast against the
JTCVS shape.

### Write the Word file — `hv_man_table_save()`

`hv_man_table_save()` writes the `.docx`. Footnotes default to
`hv_man_footnotes()`, the two notes that are universal to the house rules rather
than specific to any one study.

```{r}
#| label: hv-footnotes
hv_man_footnotes()
```

The `*` gives the non-missing-value count the wording the rules require, and the
`†` explains the median-and-percentiles format. Each symbol is appended as a
superscript reference mark on the `N` column header, or on the first column when
there is no `N` column, which is the case here: both marks land on
`Characteristic`. The text itself is written as a real paragraph after the table.
That is the part
`flextable::footnote()` cannot do for you: it renders footnote text as an extra
row inside the table's own frame, which is the violation the whole exercise is
meant to avoid.

The abbreviation key works the same way, as a `Key:` paragraph below the
footnotes, sorted alphabetically.

```{r}
#| label: hv-save
#| eval: false
hv_man_table_save(
  ft,
  file = "table1.docx",
  abbreviations = c(
    BMI  = "body mass index",
    eGFR = "estimated glomerular filtration rate"
  )
)
```

Override the footnotes with ordinary list operations. `footnotes = NULL`
suppresses both, ``hv_man_footnotes()["†"]`` keeps one, ``modifyList(hv_man_footnotes(),
list(`†` = "..."))`` replaces one, and ``c(hv_man_footnotes(), list(`‡` = "..."))``
adds a study-specific note alongside the defaults. Keeping only the `†` is worth
knowing about, for reasons the reading guide below gets to.

### Check what you wrote — `hv_check_docx()`

Writing the file and trusting it are different things, and the failure modes
here are structural rather than visual. `hv_check_docx()` reads a `.docx` and
reports the three patterns the rules name: text positioned in a floating layer
over the table, a hidden spacer column, and a footnote embedded as a row inside
the table frame. It returns one row per finding, so zero rows means clean.

```{r}
#| label: hv-check
out <- tempfile(fileext = ".docx")
hv_man_table_save(ft, out, abbreviations = c(BMI = "body mass index"))
hv_check_docx(out)
```

It never modifies the file, and it does not care who wrote it, so it also works
on a table a co-author sends back after editing it in Word. That is the case
worth remembering. Your own output is compliant by construction; the round trip
through somebody else's copy of Word is where a merged cell or a text box
actually gets introduced.

Two of the three detectors are structural and reliable. The hidden-column test
deliberately requires a column to be *both* completely empty and narrower than a
tenth of an inch, because the house JTCVS templates use empty half-inch gutter
columns on purpose and either condition alone would flag them. The
embedded-footnote detector is a pattern match on the last row of a table, so
treat what it reports as a prompt to look rather than proof.

### Why not `gt::gtsave()`?

The honest answer starts with a concession: `gt` writes Word, and writes it
well. `gt::gtsave(as_gt(tbl), "table1.docx")` hands you a `.docx` with the right
numbers on an unmerged grid, and `hv_check_docx()` finds nothing wrong with it.
The shortcut does not mangle the table.

What `gt` has no mechanism for is putting anything *outside* the table. The
house rules want the footnotes and the abbreviation key as ordinary document
paragraphs below the table; a note in `gt` is a row in the table. Save the same
summary both ways and the difference is countable. A `.docx` is a zip archive
whose `word/document.xml` holds the body: `<w:tbl>` wraps the table, `<w:tr>` is
a row, and `<w:gridSpan>` is a merged cell.

```{r}
#| label: hv-docx-scan
docx_scan <- function(path) {
  d <- tempfile()
  dir.create(d)
  unzip(path, exdir = d)
  x <- paste(readLines(file.path(d, "word", "document.xml"), warn = FALSE),
             collapse = "")
  tbl_xml <- regmatches(x, regexpr("<w:tbl xmlns.*</w:tbl>", x))
  paras <- regmatches(sub(".*</w:tbl>", "", x),
                      gregexpr("<w:p[ >].*?</w:p>", sub(".*</w:tbl>", "", x)))[[1]]
  data.frame(
    rows_in_table     = sum(gregexpr("<w:tr[ >]", tbl_xml)[[1]] > 0),
    merged_cells      = sum(gregexpr("<w:gridSpan", tbl_xml)[[1]] > 0),
    paras_after_table = sum(grepl("<w:t[ >]", paras))
  )
}
```

`out` is still the file `hv_man_table_save()` wrote a moment ago. Give `gt` the
same footnote to place and let it choose where.

```{r}
#| label: hv-vs-gtsave
gt_out <- tempfile(fileext = ".docx")
gt::gtsave(
  gt::tab_source_note(as_gt(tbl), "Median (15th, 85th percentile)."),
  gt_out
)

rbind(hv_man_table_save = docx_scan(out),
      gt_gtsave         = docx_scan(gt_out))
```

Five rows against six. `gt`'s extra row is the footnote itself, merged across
all three columns and sitting inside the table frame, which is the violation the
package exists to route around, arriving by the shortest available path. The
three paragraphs after `hv_man_table_save()`'s table are the two default
footnotes and the `Key:` line. `gt`'s file has nothing after its table, because
in `gt` there is nowhere else for that text to go.

That is the whole of the difference, and it is enough. The grid `gt` draws is
fine. It is the material around the grid, which is most of what the CORR memo
legislates, that the house functions exist to place.

## Variations

### JTCVS submissions

The JTCVS path starts somewhere different. `hv_tbl_summary()` is a wrapper over
`gtsummary::tbl_summary()` written to the interface the biostats team already
knows from the `%summarytable` SAS macro: you give it a grouped, ordered
variable list and sort the variables into type buckets, instead of using
`gtsummary`'s tidyselect arguments.

```{r}
#| label: hv-jtcvs-summary
tbl_j <- hv_tbl_summary(
  dta,
  by     = "sex",
  groups = list(
    "Demographics" = c("age", "bmi"),
    "Comorbidity"  = c("gfr_bs", "diabetes")
  ),
  continuous  = c("age", "bmi", "gfr_bs"),
  categorical = c("diabetes"),
  compare     = "pvalue"
)
```

`groups` is the macro's `LIST=`, and it does double duty: it fixes the display
order and it supplies the section headers that break the table into blocks.
`continuous`, `binary`, and `categorical` are `CON1=`, `CAT1=`, and `CAT2=`.
Every variable named in `groups` has to appear in exactly one bucket and every
bucketed variable has to appear in `groups`, which is a strict rule but a
forgiving one to debug, since the error names the variables that fell through.
Continuous variables are summarised as median with the percentile pair from
`percentiles` (15th and 85th by default) and tested non-parametrically
throughout, so there is no per-variable normality decision to make or defend.

`compare` adds the trailing comparison column: `"pvalue"`, `"smd"`, `"both"`, or
`"none"`. The result carries two attributes that the renderer wants, and you
pass them straight through.

```{r}
#| label: hv-jtcvs-attrs
attr(tbl_j, "hv_stat_label")
attr(tbl_j, "hv_trailing")
```

`hv_man_table_jtcvs()` then builds the merged shape. Its own `groups` argument
is a different thing entirely from `hv_tbl_summary()`'s: here it maps each
`stat_` column to its spanning header label, and you write the group's N into
the label text yourself.

```{r}
#| label: hv-jtcvs-table
n_sex <- table(dta$sex)

ft_j <- hv_man_table_jtcvs(
  tbl_j,
  groups = c(
    stat_1 = sprintf("Female (n=%d)", n_sex[["Female"]]),
    stat_2 = sprintf("Male (n=%d)",   n_sex[["Male"]])
  ),
  trailing   = attr(tbl_j, "hv_trailing"),
  stat_label = attr(tbl_j, "hv_stat_label")
)
ft_j
```

Compare the structure with the flat table above and the two modes separate
cleanly: two header rows here instead of one, spanning header cells, and
full-width shaded section rows in the body.

```{r}
#| label: hv-jtcvs-structure
rbind(hv_man_table       = shape(ft),
      hv_man_table_jtcvs = shape(ft_j))
```

JTCVS marks each p-value with a superscript letter naming the test behind it,
which is the same thing `%summarytable` does. `hv_test_footnotes_jtcvs()` reads
the test `gtsummary` recorded for each row and builds those footnotes, including
the body-row indices, so you are not counting rows by hand.

```{r}
#| label: hv-jtcvs-footnotes
fn <- hv_test_footnotes_jtcvs(tbl_j)
fn
```

Letters follow a fixed order rather than order of appearance, so two tables in
one manuscript that use the same tests get the same letters. Pass the result
to `hv_man_table_save_jtcvs()`, which adds the bold caption paragraph above the
table and renders the lettered notes below it.

```{r}
#| label: hv-jtcvs-save
#| eval: false
hv_man_table_save_jtcvs(
  ft_j,
  file      = "table1_jtcvs.docx",
  caption   = "Table 1. Baseline Characteristics",
  footnotes = fn,
  abbreviations = c(eGFR = "estimated glomerular filtration rate")
)
```

The caption is not auto-numbered; write the number into the string yourself.

## Read it

The part worth reading closely in either format is the denominator: which N
belongs to the percentage in front of you, and where the table says so.

The flat table says so once, in the header. `N = 77` and `N = 123` are group
sizes, and every percentage in the column beneath is taken against them. That is
all the counting the table does. Where a variable has missing data, plain
`tbl_summary()` reports it as an `Unknown` row underneath that variable rather
than by adjusting a denominator, so read the rows, not the header, for
availability. Note what this means for the default `*` footnote: it describes a
count of non-missing values, and a flat table built this way prints no such
count. On a table with no missing data the note is harmless but idle; on one
with missing data it points at a number that is not there. Either give the table
that count or leave the `*` out, with ``footnotes = hv_man_footnotes()["†"]``.

The JTCVS table says so on every row. The N sub-column beside each group's
statistic repeats the group N alongside the number it divides, so a reader
comparing two rows never has to track back to the header. It is the same
quantity the flat header carries, printed where it is used rather than once at
the top.

Then the section headers, which are a JTCVS feature and not a flat-table one.
`hv_tbl_summary()`'s `groups` breaks the variable list into labelled blocks, and
those blocks are an editorial claim about what belongs with what. A reader
scanning a Table 1 reads the block labels before the rows. Get them wrong and a
comorbidity reads as a demographic. The flat table has no blocks at all. It
lists the variables in the order you gave `include`, and that order is the only
structural claim you get to make, so make it deliberately.

## Pitfalls

- **`hv_tbl_summary()` output goes to `hv_man_table_jtcvs()`, not
  `hv_man_table()`.** This is the one that will actually bite you, because it
  fails silently rather than erroring. `hv_tbl_summary()` packs each cell as
  `"{N_obs} ||| {stat}"`, and the `|||` is a sentinel that
  `hv_man_table_jtcvs()` splits on to build its paired N and statistic
  sub-columns. `hv_man_table()` does not split on it, so the cells render
  literally as `77 ||| 46 (30, 62)` and you get a table that looks like a bug in
  the package. Build the flat CORR table from a plain `tbl_summary()`, as the
  first recipe above does, and keep `hv_tbl_summary()` for the JTCVS path.

- **`groups` means two different things.** In `hv_tbl_summary()` it is a named
  *list* of section label to variable names, the row structure. In
  `hv_man_table_jtcvs()` it is a named *character vector* of `stat_` column to
  spanning header label, the column structure. They sit two calls apart in the
  same pipeline and neither will accept the other's value, so the error you get
  is at least immediate.

- **A standardised mean difference needs exactly two groups.** `compare = "smd"`
  and `compare = "both"` are only defined between two groups and error
  otherwise. The version of this that costs an afternoon is a `by` variable that
  is a factor carrying a level nothing in the data has — after a cohort filter,
  say. It looks like two groups when you `table()` it and counts as three here.
  `droplevels()` on the filtered frame is the fix. `compare = "pvalue"` works at
  any number of groups.

- **Footnote row indices count the section-header rows.** When you write
  footnotes for `hv_man_table_save_jtcvs()` by hand, `row` indexes the rendered
  body rows *including* the section headers the renderer interleaves, not the
  rows of your original summary. Count off the printed table, not the data
  frame. For test footnotes, let `hv_test_footnotes_jtcvs()` compute the indices,
  which is most of why it exists.

- **The check is not automatic.** `hv_man_table_save()` and
  `hv_man_table_save_jtcvs()` write compliant output, but neither one inspects
  the file afterwards. If you want the check, run `hv_check_docx()` on the path
  yourself. It is cheap, and the moment to do it is after the table has been
  through anyone else's hands.
