# Using legends

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

ggplot2 [@R-ggplot2] places the legend outside the panel by default. For publication figures
we often move it inside, drop it, or shrink it, since the axis labels frequently do
the identification work already. Legend behaviour is controlled through
`theme()` (or directly via a theme's `...`), and per-aesthetic via `guide =`.

## When to use it

A legend is the key that tells a reader which colour or shape means which group.
You need one whenever the grouping is not already named on the panel, and you
want it gone the moment it is. The default position, off to the right, is safe
but wastes width: it shrinks the panel to make room for a box the reader glances
at once. So the real decisions are where to put the legend and whether to keep
it at all.

Move the legend inside the panel when there is empty space the data do not use
(a survival curve leaves a wide gap in one corner) and you would rather spend
that gap than narrow the plot. Move it to the bottom when the labels are long
and a corner would crowd. Drop it entirely when an annotation or the axis labels
already identify the groups. Shrink its text and swatches when it has to sit
inside a busy panel without stealing attention from the data. We use a grouped
trends plot below so there is a multi-level legend to position.

```{r}
#| label: base
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))
```

## Inside the panel

Pass fractional coordinates `c(x, y)` to `legend.position`. `c(0, 0)` is
bottom-left; `c(1, 1)` is top-right. A light background box keeps it legible
over data.

```{r}
#| label: fig-legends
#| fig-cap: "Grouped trends plot with the legend moved inside the panel via fractional coordinates and given a light background box"
p_base +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = NULL
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_manuscript() +
  theme(
    legend.position   = c(0.15, 0.2),
    legend.background = element_rect(fill = "white", colour = "grey80",
                                     linewidth = 0.3)
  )
```

## Automatic placement

Choosing the corner by hand means re-checking it every time the data, axis
range, or cohort change, which is the first pitfall below. `hv_legend_inside()` does the
choosing for you: it inspects the built plot, finds the panel corner the data
leave emptiest, and anchors the legend there, falling back to an outside legend
when no corner is clear. Apply it last, after the theme (which sets
`legend.position = "none"`), so its position wins.

```{r}
#| label: fig-legends-auto
#| fig-cap: "The grouped trends plot with hv_legend_inside() choosing the emptiest corner automatically"
p_auto <- p_base +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = NULL
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_manuscript()

hv_legend_inside(p_auto)
```

Because it reads the rendered data, the same call keeps the legend off the data
after you change the cohort or axis range. On a dense panel where every corner is
occupied, as with stratified survival or hazard curves, it returns an outside legend
instead of forcing the key over the data. Add a background box with
`+ theme(legend.background = element_rect(...))` when the chosen corner is not
fully empty.

## Outside the panel (explicit sides)

Set `legend.position` to `"right"`, `"left"`, `"top"`, or `"bottom"`.
`"bottom"` with `legend.direction = "horizontal"` is the common choice for
multi-group figures where a corner legend would crowd the panel.

```{r}
#| label: fig-legends-bottom
#| fig-cap: "The same plot with the legend placed horizontally below the panel, the common choice for multi-group figures"
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_manuscript() +
  theme(
    legend.position  = "bottom",
    legend.direction = "horizontal"
  )
```

## Suppress all legends

`theme(legend.position = "none")` hides every legend in one shot.
Alternatively, `guide = "none"` on each `scale_*()` call drops only that
aesthetic's legend, handy when you want a colour legend but no shape legend.

```{r}
#| label: fig-legends-none
#| fig-cap: "Single-stratum survival curve with all legends suppressed, since the axis labels already identify the group"
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)

plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_manuscript()
```

## Legend text and key size

The legend inherits the theme's font size, often larger than needed inside a
crowded panel. `legend.text` controls the label font; `legend.key.size` shrinks
the colour/shape swatch.

```{r}
#| label: text
p_base +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15, "Group II" = 19,
               "Group III" = 17, "Group IV" = 18),
    name = "Group"
  ) +
  labs(x = "Surgery Year", y = "Outcome") +
  theme_hv_manuscript() +
  theme(
    legend.text     = element_text(size = 9),
    legend.key.size = unit(0.4, "cm")
  )
```

## Pitfalls

- **The legend covers data.** Moving a legend inside the panel saves width, but
  the inside of the panel is where the data live. Drop the box on a corner that
  looks empty today and it can sit squarely on a curve once you change the data,
  the axis range, or the cohort. Check the placement against the actual figure,
  not a mental image of it, and prefer a corner the data genuinely avoid (the
  lower-left of a survival curve, the upper-left of a rising trend). When no
  corner is safe, move the legend to the bottom rather than forcing it over the
  data. `hv_legend_inside()` (above) automates exactly this check, taking the
  emptiest corner or an outside fallback when none is clear, so it does not have to be
  redone by hand each time the data change.
- **Redundant legends.** A legend that repeats what the panel already says is
  clutter. If an annotation labels each line, or the axis names the single
  group, the legend adds nothing and costs space. The same redundancy creeps in
  when colour and shape both map to the same grouping and each gets its own
  legend box: suppress one with `guide = "none"` so the reader sees a single key,
  not two that say the same thing. Every element on the figure should earn its
  place; a legend that names something already named does not.
