# Preparing figures for manuscripts

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

Journals impose hard constraints on figures: a fixed column or page width, a
maximum height, and a vector format (PDF/EPS) so type and lines stay crisp at
print resolution. The trap is that two figures saved at the same `ggsave()`
dimensions rarely share the same *data region*: a longer axis label or a wider
legend steals space from the panel, so the plotting area drifts from figure to
figure. `theme_hv_manuscript()` sets a print-appropriate base size, and
`hv_ggsave_dims()` pins the panel content area to a constant rectangle across a
figure set.

## A manuscript-styled plot

`theme_hv_manuscript()` uses a 12 pt base on a white background, sized for the
final printed page rather than the screen.

```{r}
#| label: build-manuscript
p    <- plot(hv_survival(sample_survival_data(n = 500, seed = 42)))
p_ms <- p + theme_hv_manuscript()
p_ms
```

## Manuscript PDF

For most figures the quickest path is `save_manuscript()`, which writes a
finished plot at the house manuscript size, 6 × 4 inches, in one call.

```{r}
#| label: save-manuscript-default
#| eval: false
save_manuscript(p_ms, "trends_manuscript.pdf")
```

It is the manuscript counterpart of `save_ppt()`: it fixes the output geometry,
not the theme, so the 12 pt type comes from the `theme_hv_manuscript()` you have
already applied. Override the size when a journal asks for it, for a single
column, say `save_manuscript(p_ms, "fig.pdf", width = 3.5, height = 3)`, and
pass a `device` when you need font embedding, with the caveat below. The chunks
in this section are `eval: false` because each writes a binary file.

### Holding the panel constant across a figure set

`save_manuscript()`, like `ggsave()`, fixes the *file* dimensions but lets the
panel float inside them. When you need the panel content area identical across
several figures, so the data regions line up regardless of how much room each
plot's axis labels take, `hv_ggsave_dims()` inverts the problem: you specify the
panel content area you want (the rectangular bounding box of the gtable cells
tagged `panel`) and it returns the `width`/`height`/`units` such that `ggsave()`
yields exactly that panel size, expanding the file to absorb axes, titles, and
any remaining chrome. The result is a named list shaped to splat directly into
`ggsave()` via `do.call()`.

```{r}
#| label: save-manuscript
#| eval: false
dims <- hv_ggsave_dims(p_ms, width = 6.5, height = 4.0, units = "in")

do.call(
  ggsave,
  c(
    list(filename = "trends_manuscript.pdf", plot = p_ms,
         device = grDevices::quartz, type = "pdf"),
    dims
  )
)
```

The device matters, and the usual recommendation has a failure mode worth
knowing. `cairo_pdf` embeds fonts and renders transparency and custom typefaces
correctly, where the base `pdf` device substitutes them. But on a Mac without
XQuartz it warns "failed to load cairo DLL", returns **without error**, and writes
no file. Measured on this machine: `ggsave()` reported success and the target PDF
did not exist. A silent no-op is worse than a failure you can see, so check that
the file appeared before assuming a cairo export worked.

The quartz device above is the macOS answer: part of the operating system,
nothing to install, fonts embedded, full Unicode range. It is what this book's own
PDF build uses, for the same reason. On Linux, `cairo_pdf` with a working cairo is
the right call. The chunk is `eval: false` because it writes a binary file; the
code is shown, not run during the book build.

For multi-panel figure sets where the data region must be identical across
several PDFs regardless of label length, call `hv_ggsave_dims()` with the same
`width`/`height` for each figure. Its slide-layout analogue,
`hv_ph_location()`, computes `officer::ph_location()` coordinates so a panel
lands at the same slide position in a deck.

## PowerPoint export

`save_ppt()` writes the plot into a PowerPoint slide as **editable DrawingML
vector graphics** via the `officer` and `rvg` packages: shapes, lines, and
text remain selectable in PowerPoint, so co-authors can re-label or recolor
without round-tripping back to R. It appends a slide to a template `.pptx`
rather than overwriting it.

```{r}
#| label: save-ppt
#| eval: false
template <- system.file("ClevelandClinic.pptx", package = "hvtiPlotR")

p_ppt <- p + theme_hv_ppt_dark()

save_ppt(
  object       = p_ppt,
  template     = template,
  powerpoint   = "trends_slides.pptx",
  slide_titles = "Overall Survival",
  width        = 10.1,
  height       = 5.8
)
```

This chunk is `eval: false` because it writes a `.pptx` file. To anchor the
panel content area at identical slide coordinates across a multi-slide deck
(useful when axis-label widths differ between plots), pass
`panel_box = list(width, height, left, top)`; `save_ppt()` delegates to
`hv_ph_location()` per slide.
