# Number needed to treat and survival difference

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

A hazard ratio is a relative comparison. It does not tell a patient how much
survival changes at five years, and it can hide when that change occurs. The
absolute survival difference answers that first question directly: how many
more patients out of 100 are alive in group 2 than group 1 at time $t$. Number
needed to treat (NNT) asks a different question by inverting the benefit: how
many patients must receive group 2 rather than group 1 to prevent one event by
that same time.

Direction is part of both estimands. In this chapter the difference is
$S_2(t)-S_1(t)$. A positive value favors group 2, and when the value is expressed
in percentage points, $NNT(t)=100/[S_2(t)-S_1(t)]$. If group 2 has lower
survival, call that harm and reverse the contrast before reporting an NNT.

Reach for the survival difference when you want to show the size and timing of a
benefit and whether its confidence band clears zero. Reach for NNT when you want
the bedside number, the figure that makes a small absolute benefit feel concrete
or an over-hyped one feel modest. Both constructors, from hvtiPlotR
[@R-hvtiPlotR], take a pre-computed data frame. Build and inspect the S3 object
first; then `plot()` returns the ggplot you decorate.

## The data it needs

Both functions expect one row per time point with an estimate column and
explicit lower/upper CI columns that you name in the call. `sample_nnt_data()`
generates two survival curves from a hazard-ratio contrast and computes NNT and
ARR (absolute risk reduction) at each time point with CI bounds. The `groups`
argument names the arms and sets their hazard multipliers; here the internal
thoracic artery graft (ITA) carries 0.75 times the hazard of the saphenous vein
graft (SVG) reference.

```{r}
#| label: nnt_data
nnt_dat <- sample_nnt_data(
  n        = 500,
  time_max = 20,
  groups   = c("SVG" = 1.0, "ITA" = 0.75)
)
head(nnt_dat)

nnt <- hv_nnt(
  nnt_dat,
  lower_col = "nnt_lower",
  upper_col = "nnt_upper"
)
nnt$meta

report_times <- c(5, 10, 20)
report_rows <- vapply(
  report_times,
  function(x) which.min(abs(nnt_dat$time - x)),
  integer(1)
)
nnt_report <- nnt_dat[report_rows, c(
  "arr", "arr_lower", "arr_upper", "nnt", "nnt_lower", "nnt_upper"
)]
nnt_report <- cbind(report_time = report_times, nnt_report)
rownames(nnt_report) <- NULL
round(nnt_report, 1)
```

## Build it

The NNT object stores the pre-computed estimate and its CI columns. The default
`estimate_col` is `"nnt"`, so we named only the bounds above. The plotted
contrast is ITA minus SVG survival: a positive benefit and finite NNT favor ITA.

```{r}
#| label: fig-nnt
#| fig-cap: "Number needed to treat with ITA rather than SVG; NNT falls as the benefit grows to about 10 years, then rises as that benefit narrows"
plot(nnt) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  scale_y_continuous(breaks = seq(0, 50, 10)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 50)) +
  labs(x = "Years", y = "Number Needed to Treat") +
  theme_hv_manuscript()
```

## Read it

This simulated NNT curve falls while the ITA survival benefit widens, reaches
about 6 near 10 years, then rises to about 17 at 20 years as that benefit
narrows. The report table above ties those readings to named horizons. Look for:

- **The direction of change.** Early in follow-up the two arms have barely diverged,
  so you must treat a great many patients to prevent one event and the NNT is
  large. NNT drops while the survival gap widens and rises if that gap later
  narrows. Read the value at the time horizon that matters for your decision,
  not at the visual minimum or endpoint by default.
- **The confidence band.** A wide band, especially early when NNT is large and
  unstable, means the count is poorly determined. When the survival-difference
  interval includes zero, a finite two-sided NNT interval does not exist; the
  ribbon may therefore be one-sided or absent at those times.
- **The y-axis scale.** NNT is a ratio and blows up when the absolute benefit is
  near zero. A capped axis (here 0 to 50) keeps the figure readable, but
  remember the curve is heading to infinity at time zero, not flattening.

## Variations

### Absolute risk reduction

The same data can show absolute risk reduction (ARR) in percentage points by
setting `estimate_col = "arr"` and naming the matching CI columns. Here ARR is
the ITA minus SVG survival difference, which is also the reduction in event
risk. NNT equals 100 divided by ARR, not the raw reciprocal of the plotted
percentage-point value.

```{r}
#| label: fig-nnt-arr
#| fig-cap: "Absolute event-risk reduction for ITA versus SVG in percentage points; positive values favor ITA and the zero line marks no difference"
arr <- hv_nnt(
  nnt_dat,
  estimate_col = "arr",
  lower_col    = "arr_lower",
  upper_col    = "arr_upper"
)

plot(arr) +
  scale_colour_manual(values = c("firebrick"), guide = "none") +
  scale_fill_manual(values = c("firebrick"), guide = "none") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  scale_y_continuous(breaks = seq(-10, 30, 10)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(-10, 30)) +
  labs(x = "Years", y = "Absolute risk reduction (percentage points)") +
  theme_hv_manuscript()
```

### Survival difference

`hv_survival_difference()` plots the difference `S_2(t) - S_1(t)` between two
groups over time with an optional confidence band.
`sample_survival_difference_data()` generates the point-wise difference with
bootstrap CI bounds from a hazard-ratio contrast. Both arms are simulated at the
same n to keep the band width realistic.

```{r}
#| label: surv_diff_data
diff_dat <- sample_survival_difference_data(
  n      = 500,
  groups = c("Control" = 1.0, "Treatment" = 0.70)
)
head(diff_dat)

surv_diff <- hv_survival_difference(
  diff_dat,
  lower_col = "diff_lower",
  upper_col = "diff_upper"
)
surv_diff$meta
```

The dashed line at zero is the no-difference baseline. Because Control is group
1 and Treatment is group 2, a positive value means higher Treatment survival;
a negative value would favor Control. No grouping argument is needed for this
two-group case. Read the figure two ways: the height is the percentage-point
difference at each time, and the lower edge of the 95% band shows whether the
data remain compatible with no difference.

```{r}
#| label: fig-nnt-survival-difference
#| fig-cap: "Treatment minus Control survival difference in percentage points with its 95% confidence band; positive values favor Treatment"
plot(surv_diff) +
  scale_colour_manual(values = c("steelblue"), guide = "none") +
  scale_fill_manual(values = c("steelblue"), guide = "none") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(breaks = 0:10) +
  scale_y_continuous(breaks = seq(-5, 35, 5)) +
  coord_cartesian(xlim = c(0, 10), ylim = c(-5, 35)) +
  labs(x = "Years", y = "Survival difference (percentage points)") +
  theme_hv_manuscript()
```

### Multiple comparisons

Combine several two-group differences into one long data frame and use
`group_col` to overlay them. This is how we put one reference arm against
several alternatives in a single panel: each line is one contrast, and the ones
that sit highest and clear zero earliest are the comparisons with the strongest
benefit.

```{r}
#| label: fig-nnt-survival-difference-multi
#| fig-cap: "Several two-group survival differences overlaid in one panel, one line per contrast against a shared reference arm"
d1 <- sample_survival_difference_data(
  groups = c("Medical Mgmt" = 1.0, "TF-TAVR" = 0.70), seed = 1L
)
d1$comparison <- "TF-TAVR vs Medical Mgmt"

d2 <- sample_survival_difference_data(
  groups = c("TA-TAVR" = 0.90, "TF-TAVR" = 0.70), seed = 2L
)
d2$comparison <- "TF-TAVR vs TA-TAVR"

d3 <- sample_survival_difference_data(
  groups = c("AVR" = 0.80, "TF-TAVR" = 0.70), seed = 3L
)
d3$comparison <- "TF-TAVR vs AVR"

surv_diff_multi <- hv_survival_difference(
  rbind(d1, d2, d3),
  group_col = "comparison"
)

plot(surv_diff_multi) +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(x = "Years", y = "Survival difference (percentage points)") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

## Pitfalls

- **Quoting NNT at time zero.** NNT is a reciprocal, so it explodes near the
  start of follow-up where the arms have not yet diverged. Always report it at a
  named horizon, never as a single headline number with no time attached.
- **Mismatched CI columns.** NNT and ARR each have their own pair of bounds.
  Plot ARR with the NNT bounds and the band will be wrong. Name the CI columns
  that go with the estimate you are showing.
- **Reading the survival difference where the band straddles zero.** A positive
  point estimate whose confidence band still includes zero is compatible with
  no difference. Keep the estimate and interval together.
- **Leaving the contrast unnamed.** The sign means nothing until the reader
  knows which survival curve was subtracted from which. Put “group 2 minus
  group 1” in the text or caption, and reverse the labels if you want positive
  values to mean benefit.
- **Forgetting the zero line.** The dashed reference at zero is what makes the
  survival-difference figure readable. Add the `geom_hline()` every time; without
  it the eye has no anchor for "no benefit".
