# Packages

A short stack does most of the work in this book. `ggplot2` [@R-ggplot2] is the
foundation; everything else either builds figures on top of it or supplies the
modeling output we plot. If you have not installed any of it yet, skip to
@sec-installing at the end of this chapter and come back.

```{r}
#| label: load-packages
#| message: false
#| warning: false
library(ggplot2)
library(hvtiPlotR)        # house plots, themes, and PPT/PDF export
library(hvtiRutilities)   # data dictionaries, labels, manifests
library(gt)               # publication tables
library(patchwork)        # compose multi-panel figures
```

What each one is for:

- **hvtiR** [@R-hvtiR]: the family installer and health check. It puts the
  related packages on the same GitHub-main release path, tells you whether an
  installed member has moved, updates only the members that need it, and
  diagnoses an R or GitHub setup that cannot get that far.
- **hvtiPlotR** [@R-hvtiPlotR]: the core of the book. It wraps the figures we
  publish into a two-step API: an `hv_*()` constructor validates the data and
  returns an object, and `plot()` turns that object into a bare ggplot. It also
  carries the house themes (`theme_hv_manuscript()`, `theme_hv_poster()`, the
  PowerPoint themes) and the save helpers.
- **hvtiRutilities** [@R-hvtiRutilities]: the data-governance side: data
  dictionaries, variable labels, dataset manifests, and the synthetic-data
  generators we use so the examples run without touching real patient data.
- **ggRandomForests** [@R-ggRandomForests], **randomForestSRC**
  [@R-randomForestSRC; @randomForestSRC2007; @randomForestSRC2008], and **varPro**
  [@R-varPro]: the random-forest stack. `randomForestSRC` fits the forest,
  `varPro` does variable selection, and `ggRandomForests` turns either one into a
  ggplot. These load in the random forest chapters where they are used.
- **TemporalHazard** [@R-TemporalHazard]: our R port of the SAS/C HAZARD
  module, for parametric and nonparametric hazard modeling. It loads in its own
  chapter.
- **gt** [@R-gt]: publication-quality tables, and the right choice when the
  table's destination is this book or another HTML document.
- **hvtiRtables** [@R-hvtiRtables]: manuscript tables. It takes a `gtsummary`
  summary and renders it through `flextable` into a Word table that follows the
  CORR "Table Construction for Manuscripts" rules: flat headers with nothing
  merged, no hidden spacer columns, and footnotes and the abbreviation key as
  text below the table. A JTCVS mode swaps in that journal's merged spanning
  headers and lettered footnotes instead.
- **gtsummary** [@R-gtsummary] and **flextable** [@R-flextable]: the two
  packages `hvtiRtables` is built on. `gtsummary::tbl_summary()` computes the
  baseline summary and `flextable` is the Word rendering layer. You call the
  first one directly in the manuscript tables chapter; the second mostly stays
  under the hood.
- **kernelshap** [@R-kernelshap]: the SHAP engine behind `gg_shap()`. It loads
  in the SHAP attribution chapter, and it is the one dependency in the book
  whose cost you feel, so that chapter says what the two size arguments buy you.
- **patchwork** [@R-patchwork]: composes several plots into one figure with
  shared labels and panel tags.
- **survival** [@R-survival; @survival-book]: the `Surv()` objects and
  Kaplan-Meier machinery underneath the survival figures.
- **mlbench** [@R-mlbench]: example data only: the ROC chapter borrows its
  `SynthDiabetes` set for a classification forest that discriminates well enough
  to draw a ROC curve worth reading. Loaded just in that chapter. Version 2.1-11
  or newer, which is where that synthetic set replaced the withdrawn Pima Indians
  data. On an older `mlbench` the chapter stops at `object not found`.

You do not need all of these for any single figure. Each chapter loads what it
uses in its first chunk.

## Installing the stack {#sec-installing}

Install the family from GitHub. Two of these packages, `ggRandomForests` and
`TemporalHazard`, are on CRAN as well, but the GitHub release comes first and
CRAN follows it, so GitHub is where the whole family stays in step. `hvtiR`
[@R-hvtiR] is the small control panel for that lifecycle. Use the command that
matches the state you are in:

```r
install.packages("pak")
pak::pak("ehrlinger/hvtiR")
hvtiR::install()       # fresh machine
hvtiR::status()        # compare installed packages with GitHub main
hvtiR::update()        # install only missing or stale family members
hvtiR::doctor()        # diagnose R, pak, GitHub, and member status
```

On a fresh machine, run the first three lines in a fresh R session, before
anything attaches a family package. `install()` stops rather than replacing a
loaded namespace, so calling it after the `library()` chunk above fails with
`already loaded in this session`. Restart R before installing or updating, then
use `status()` to check a working installation. It compares installed versions
with GitHub `main` and, when `pak` recorded the source commit, catches a new
commit even when the package version did not change.

When `status()` marks a member stale or missing, restart R and run `update()`.
It installs only those family members, still following the GitHub-main release
order. If the setup itself is puzzling, `doctor()` checks the R version, `pak`,
GitHub access, and each member's status in one report. Avoid
`update.packages()` for this family: it can replace a GitHub install with an
older CRAN release, leaving a family member out of step. Run `status()` and
then `update()` instead.

That call installs the eleven family packages, and their dependencies bring in
most of the rest of the stack with them: `ggplot2`, `patchwork`, `survival` and
`ggalluvial` arrive with `hvtiPlotR`, `gtsummary` and `flextable` with
`hvtiRtables`, `randomForestSRC` and `varPro` with `ggRandomForests`. Four
packages the book uses are not reachable that way, and install on their own:

```r
pak::pak(c("gt", "kernelshap", "mlbench", "ehrlinger/ggsankey"))
```

Expect that second list to get shorter. It holds only what no family package
depends on yet, so a package the family absorbs drops out of it. `gt` is the
likely next one to go, as table work consolidates into `hvtiRtables`.

Later, to find out whether an installed package has fallen behind GitHub `main`:

```r
hvtiR::status()
```

It prints one row per family package, installed version against what is on
GitHub `main`. That is the check worth running before you trust a figure you
rendered a while ago.

## What this build used {#sec-provenance}

Every figure in this book is served from a cached render, so the versions that
produced it are a property of the cache rather than of your session. The table
below is written by the render itself, and it is the only trustworthy record of
what the current cache was built against.

```{r}
#| label: sibling-versions
#| echo: false
#| message: false
#| tbl-cap: "Family package versions this build was rendered against."
siblings <- c(
  "hvtiR", "hvtiPlotR", "hvtiRutilities", "hvtiRtables",
  "TemporalHazard", "ggRandomForests", "ggsankey"
)
versions <- vapply(
  siblings,
  function(p) as.character(utils::packageVersion(p)),
  character(1)
)

# Written to the repository root, where the scheduled version check reads it.
jsonlite::write_json(
  list(
    rendered = format(Sys.Date()),
    r = paste(R.version$major, R.version$minor, sep = "."),
    packages = as.list(versions)
  ),
  "sibling-versions.json",
  auto_unbox = TRUE, pretty = TRUE
)

knitr::kable(
  data.frame(Package = siblings, Version = unname(versions)),
  row.names = FALSE
)
```

If that table is older than a family release you know about, the figures are
too. `hvtiR::status()` tells you which packages have moved since, and
@sec-installing has the re-render.
