# Postage stamp plots

Does the relationship between two measurements look the same across clinical
groups or eras? A postage stamp plot gives each subgroup the same small panel so
the layout carries the comparison.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

Sometimes overlaying every group on one panel turns into spaghetti: too many
colours, too many overlapping clouds, and the comparison you wanted is lost.
The fix is to give each group its own small panel and line them up on a shared
grid. The result is a "postage stamp" plot, the same plot repeated across the
levels of a categorical variable, each panel reading like a stamp on a sheet.
Because every panel shares the same axes and geometry, your eye compares them
the way it compares photographs in a contact sheet: the layout does the
alignment for you. Reach for it when one figure has to carry a per-subgroup
comparison and an overlay would be too busy.

hvtiPlotR 2.7.10 has no postage-stamp or faceting constructor, so we build this
layout directly with ggplot2's
[@R-ggplot2] `facet_wrap()` and apply the house theme `theme_hv_manuscript()` for
visual consistency. The faceting is the whole trick: you write one `ggplot()`
call describing a single panel, add `facet_wrap()`, and ggplot2 repeats the panel
once per group.

## The data it needs

A faceted scatter needs an x and a y for the relationship inside each panel, plus
a categorical column to split the panels on. We use
`hvtiPlotR::sample_eda_data()` with `op_years` (years since first surgery) for x
and `ef` (ejection fraction) for y, and we facet on operative `year`. A postage
stamp only earns its keep when there are *many* small panels, so we restrict to a
nine-year window, which lays out as a clean three-by-three sheet.

```{r}
#| label: data
dat <- sample_eda_data()
# A nine-year window facets into a 3 x 3 sheet of small multiples.
dat      <- dat[dat$year %in% 2012:2020, ]
dat$year <- factor(dat$year)
xvar <- "op_years"
yvar <- "ef"
grp  <- "year"
c(x = xvar, y = yvar, facet = grp, panels = nlevels(dat[[grp]]))
```

The `.data[[xvar]]` idiom is how ggplot2 reads a column name held in a variable,
so the recipe runs against whatever columns you name. Faceting on a window of
years turns one busy overlay into a contact sheet you read year by year;
substitute your own grouping (centre, severity class, era) and the mechanics are
identical.

## Inspect it

There is no S3 constructor object in this direct faceting recipe. Check the
factor order and panel counts before smoothing, because a sparse or misordered
panel looks just as finished as a well-supported one.

```{r}
#| label: postagestamp_inspect
levels(dat[[grp]])
table(dat[[grp]], useNA = "ifany")
summary(dat[c(xvar, yvar)])
```

## Build it

Each panel shows the relationship between the two numeric variables for one level
of the faceting variable, with a straight fitted line to summarize the direction
within each subgroup. A linear fit is stable in these small, narrow panels;
`facet_wrap()` lays them out left to right and wraps to a new row when it runs
out of width.

```{r}
#| label: fig-postagestamp
#| fig-cap: "Ejection fraction against years from surgery, faceted into a three-by-three sheet of small multiples by operative year on shared axes"
#| fig-height: 5
ggplot(dat, aes(x = .data[[xvar]], y = .data[[yvar]])) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(vars(.data[[grp]]), ncol = 3) +
  scale_x_continuous(breaks = c(8, 12, 16)) +
  labs(
    title = "Small-multiple panels by operative year",
    x = "Years from first surgery",
    y = "Ejection fraction (%)"
  ) +
  theme_hv_manuscript()
```

The key is that all nine panels share one x-axis and one y-axis. The smooth in
each panel is computed from that panel's data alone, so the curves are directly
comparable: a higher curve in one panel really does mean a higher y there, not a
trick of a rescaled axis.

## Read it

Read a postage stamp plot by comparing panels, not by studying any one in
isolation. Years from first surgery are on the x-axis, ejection fraction is on
the y-axis, and operative year defines the panels. Each fitted line uses only
the observations in its panel. Differences among lines are descriptive
era comparisons; they do not show that operative year caused EF to differ or
that a cross-sectional pattern is a within-patient change.

- **Compare on the shared scale.** The point of the layout is that the axes are
  fixed across panels, so a difference in height or slope between two panels is a
  real difference in the data. Trust the comparison only as long as the scales
  are shared; the moment you free the scales (see below) the panels no longer
  line up and cross-panel reading breaks.
- **Look at the trend per panel.** Each panel's smooth summarises that subgroup.
  Do the curves rise in some panels and fall in others? A pattern that differs by
  panel is exactly the interaction a single overlaid plot would have buried.
- **Watch the density of points.** A panel with few points (a year with fewer
  operations) gives a smooth you should not over-read; the curve is drawn with the
  same confidence as a panel backed by ten times the data.

## Adapt it

### Fixing the panel layout

`facet_wrap()` chooses a roughly square grid by default. When you want the panels
in a single row to emphasise a left-to-right comparison, or in a single column to
stack them, set `nrow` or `ncol` directly.

```{r}
#| label: fig-postagestamp-row
#| fig-cap: "The same panels forced into a single row to emphasise a left-to-right comparison"
#| fig-width: 12
#| fig-height: 3.2
ggplot(dat, aes(x = .data[[xvar]], y = .data[[yvar]])) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(vars(.data[[grp]]), nrow = 1) +
  scale_x_continuous(breaks = c(8, 12, 16)) +
  labs(
    title = "Panels in a single row",
    x = "Years from first surgery",
    y = "Ejection fraction (%)"
  ) +
  theme_hv_manuscript()
```

### Free scales, used deliberately

By default every panel shares both axes, which is what makes cross-panel reading
honest. Sometimes one group lives on a different range and the shared scale
squashes it flat. `scales = "free_y"` lets each panel set its own y-axis. Use it
only when you care about the shape inside each panel rather than comparing
heights across them, and say so in the caption, because the panels no longer line
up.

```{r}
#| label: fig-postagestamp-free
#| fig-cap: "Panels with a free y-axis per group, for comparing the shape within each panel rather than heights across them"
#| fig-height: 5
ggplot(dat, aes(x = .data[[xvar]], y = .data[[yvar]])) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(vars(.data[[grp]]), scales = "free_y") +
  scale_x_continuous(breaks = c(8, 12, 16)) +
  labs(
    title = "Free y-axis per panel (compare shape, not height)",
    x = "Years from first surgery",
    y = "Ejection fraction (%)"
  ) +
  theme_hv_manuscript()
```

## Deliver it

Once the facet order and shared scales are fixed, use
[Finish and deliver the result](publications.qmd) for the manuscript, poster, or
slide export path. Keep the panel variable, per-panel smoother, and any free
scale choice in the caption so the exported sheet remains interpretable.

## Pitfalls

- **Free scales by accident.** The default shared scale is what makes the panels
  comparable. Switching to `scales = "free"` or `"free_y"` without flagging it
  lets a reader compare panel heights that are no longer on the same axis, which
  is the most common way a small-multiples figure misleads. Free the scales only
  on purpose, and label it.
- **Too many panels.** A facet over a high-cardinality variable produces a wall
  of tiny panels nobody can read. If a categorical has more than eight or so
  levels, collapse the rare ones into an "Other" group, or pick the levels worth
  showing, before you facet.
- **Unhelpful panel order.** `facet_wrap()` orders panels by the factor levels,
  which default to alphabetical. When the variable has a natural order (severity,
  era, anatomical class), reorder the factor first so the panels read in that
  order; otherwise the trend across panels is scrambled.
- **Over-reading a thin panel.** A subgroup with few observations gets a panel and
  a smooth that look just as authoritative as a well-populated one. Note the
  per-panel counts, or let the sparse panel speak for itself, rather than reading
  a confident-looking curve off a dozen points.
