# Slides and presentations

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

A slide is read from across a room and sits on a coloured background, not white
paper. Two things change from a manuscript figure. The type scales up the way a
poster's does, so it carries to the back of the room; and the theme has to match
the deck: dark for a projected slide, light or transparent for a printed
handout or a light template. hvtiPlotR ships a pair of PowerPoint themes for
exactly this, plus `save_ppt()`, which drops the finished plot onto a slide as an
*editable* vector graphic rather than a flat image.

## A slide-styled plot

`theme_hv_ppt_dark()` styles a plot for a projected deck: white ink, no
gridlines, and a large base size so the figure reads from the back of the room.
It deliberately leaves the figure background *transparent* so the slide's own
background shows through, which is what you want on a real deck, but it makes the
white axis text vanish on this white page. For the preview below we add a dark
`plot.background` to stand in for the slide; on the actual deck, drop that line
and let `save_ppt()` place the transparent figure on the slide.

```{r}
#| label: build-ppt-dark
p <- plot(hv_survival(sample_survival_data(n = 500, seed = 42)))
p + theme_hv_ppt_dark(base_size = 18) +
  theme(plot.background = element_rect(fill = "grey15", colour = NA))
```

`theme_hv_ppt_light()` is the same typography on a light or transparent
background, for a printed handout or a light template.

```{r}
#| label: build-ppt-light
p + theme_hv_ppt_light(base_size = 18)
```

Both themes default to a 32 pt Arial base, sized for a slide projected across a
room. We drop `base_size` to 18 here so the figure previews legibly on the page;
leave it at the default for the real deck, and raise it further for a large hall.

## One decorator for the whole deck

A theme cannot finish a grouped figure on its own. `theme()` governs the
non-data ink, the text, panel, grid and ticks, and nothing that is drawn from
the data. Add `theme_hv_ppt_dark()` to a four-group trends plot and you restyle
the axes while leaving every series in ggplot2's default hues. Series colour and
shape live in the scales, so a finished slide figure is a theme call plus
`scale_colour_manual()` plus `scale_shape_manual()`.

Three lines is nothing once. On a twelve-figure deck it is thirty-six, and the
cost is not the typing. It is that one figure drifts out of step with the rest,
usually the one you added last and looked at least.

`hv_ppt_series()` bundles the three into a single object. It returns a plain
list holding the theme and both scales, and ggplot2's `+` unrolls a list element
by element, so the bundle composes exactly the way a theme does. Define it once
near the top of the script, then add that same object to every plot in the deck.

```{r}
#| label: build-ppt-series
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_trends   <- plot(hv_trends(dta_trends))

ppt <- hv_ppt_series(mode = "dark", base_size = 18)

p_trends +
  labs(x = "Surgery Year", y = "Outcome") +
  ppt +
  theme(plot.background = element_rect(fill = "grey15", colour = NA))
```

`mode` names the background the figure will sit on, and it picks both the theme
and the colour ordering: `"dark"` wraps `theme_hv_ppt_dark()`, `"light"` wraps
`theme_hv_ppt_light()`. Anything else you pass goes through to the wrapped
theme, which is how `base_size = 18` got there above. Leave the 32 pt default in
place for a real deck.

Two arguments cover the cases where the defaults are wrong. Pass `colours` when
a deck calls for a particular set, and `shapes` when the default glyphs collide
with something else on the slide.

```{r}
#| label: build-ppt-series-light
p_trends +
  labs(x = "Surgery Year", y = "Outcome") +
  hv_ppt_series(
    mode      = "light",
    colours   = c("#0072B2", "#D55E00", "#009E73", "#CC79A7"),
    shapes    = c(16, 17, 15, 18),
    base_size = 18
  )
```

Four values here for four groups. The defaults carry six of each, which is the
ceiling for a grouping variable before a discrete scale runs out of values and
errors. Ask `hv_ppt_palette()` for a seventh colour and it stops rather than
recycling, on the reasoning that two series sharing a hue is worse than a
stopped script.

### Why colour and shape both carry the group

Both scales map the same variable, so each series is told apart twice. On your
monitor that looks like belt and braces. In the room it is not. A projector in a
dim hall flattens colour differences that are perfectly clean on screen, and a
black-and-white handout drops them altogether, but the point shapes survive
both. This is the same redundancy the [colour chapter](colors.qmd) argues for in
a manuscript figure, and the case for it on a slide is stronger, not weaker.

The shape scale still earns its keep with no legend drawn. Where two curves
cross, the plotted summary points are what tells them apart.

### Naming the series

A finished CORR figure carries no legend, on a slide or on a journal page.
`theme_hv_ppt_dark()` and `theme_hv_ppt_light()` both set
`legend.position = "none"`, and `hv_ppt_series()` deliberately leaves that
alone. You name the groups on the panel instead, in the theme's own ink.
@sec-annotation-series has that recipe and the reasoning behind the ink.

If a working draft does want a key, pass `legend.position` straight through and
give both scales a shared title, so ggplot2 merges them into one legend instead
of stacking two:

```
hv_ppt_series(mode = "dark", legend.position = "top", name = "Group")
```

## Editable PowerPoint

A figure exported as a PNG is frozen: a reviewer who wants the title reworded or
a colour swapped has to come back to you. `save_ppt()` instead writes the plot
onto a slide as a native vector graphic, so its text, lines, and colours stay
editable in PowerPoint itself. It anchors the plot in the panel rectangle of a
house slide template, so every figure lands in the same place across a deck.

```{r}
#| label: save-ppt
#| eval: false
save_ppt(
  p + theme_hv_ppt_dark(),
  template   = "RD-dark.pptx",
  powerpoint = "survival_slide.pptx"
)
```

The chunk is `eval: false` because it writes a binary `.pptx`. The code is shown,
not executed during the book build. Pair the dark theme with a dark `template`
and the light theme with a light one, and keep `theme_hv_ppt_dark()` at its 32 pt
default here, since the template is laid out for that size. Pass `panel_box` to move or
resize the plot rectangle, and `slide_titles` to set the slide heading.
