# Publication tables

For a thorough treatment of effective display tables in `gt` [@R-gt], see
Rich Iannone's guide:
<https://rich-iannone.github.io/gt-effective-display-tables/>.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(gt)
library(dplyr)
library(hvtiPlotR)
```

This chapter builds the table by hand, directly with `gt`. That is worth
doing for two reasons. Assembling a Table 1 a row group at a time is how
you learn what is actually in one: which variables earn a row, which
summary each variable type asks for, and where a reader expects the
denominators. And `gt` is the right tool when the table's destination is
an HTML document, this book included, rather than a Word file bound for a
journal.

That splits the work along a clean line: hand-built `gt` for understanding and
for the web, the `hv_*()` functions in
[manuscript tables](hv_tables.qmd) when it goes to a manuscript.

## When to use it

Open almost any clinical manuscript and the first table you meet is a
baseline-characteristics summary, the one journals number "Table 1." It
describes who was in the study before any outcome is discussed: how old the
patients were, how their risk factors split, and whether the comparison groups
looked alike at the start. Reach for it whenever you need to convince a reader
that two arms are comparable, or simply to put a face on the cohort a figure
later summarises. The job is description, not inference. A good Table 1 lets a
reader decide for themselves whether a later result is believable, because they
can see what the groups were made of.

The layout is always the same: rows of characteristics, columns split by the
grouping variable (treatment, exposure, sex). Continuous variables get a centre
and a spread (mean and standard deviation, or median and interquartile range);
categorical variables get a count and a percentage. We build the table here from
`hvtiRutilities::generate_survival_data()`, which returns a richly labelled
patient-level cohort with a natural grouping variable (`sex`), so the recipe
runs end to end before you point it at your own data. It is the same cohort
[describing a dataset](data_tables.qmd) took inventory of, so if you want to see
where the variables below came from, that chapter is one click back.

```{r}
#| label: t1_data
dta <- hvtiRutilities::generate_survival_data(n = 200, seed = 42)
str(dta[, c("sex", "age", "bmi", "gfr_bs", "diabetes")])
```

## Build it

A Table 1 is assembled one variable type at a time, then joined and rendered.
We summarise the continuous variables first, the categorical variable second,
and hand the combined frame to `gt()` for formatting. Doing it in pieces keeps
each summary readable and lets you check the numbers before they are dressed up.

### Summarise numeric variables as mean ± SD

We summarise three continuous baseline variables (age, BMI, baseline
eGFR) as mean and standard deviation within each `sex` group. Mean and SD are
the textbook pair when a variable is roughly symmetric; for a skewed
variable (cost, time on bypass) you would report median and interquartile range
instead, since a mean dragged by a long tail misrepresents the typical patient.

Know before you copy this that mean and SD are not what a CORR manuscript table
reports. The house convention is a median with the 15th and 85th percentiles,
whatever the shape, so that one summary covers the symmetric and the skewed
variable alike and there is no per-variable normality call to defend to a
reviewer. [Manuscript tables](hv_tables.qmd) builds the table that way and
prints the footnote that documents it. We stay with mean and SD here because
this chapter is assembling a table by hand for the web, and because holding the
two conventions side by side is the fastest way to see what the house one buys
you.

```{r}
#| label: t1_numeric
num_summary <- dta |>
  group_by(sex) |>
  summarise(
    n        = dplyr::n(),
    age_mean = mean(age),     age_sd = sd(age),
    bmi_mean = mean(bmi),     bmi_sd = sd(bmi),
    gfr_mean = mean(gfr_bs),  gfr_sd = sd(gfr_bs),
    .groups  = "drop"
  )
num_summary
```

### Summarise a categorical variable as n (%)

Diabetes status is summarised as a count and within-group percentage. The
percentage here is *within* each sex group, the column denominator, which is
almost always what a reader expects: "what fraction of the women had diabetes,"
not "what fraction of all diabetics were women." Make the denominator explicit
in your own head before you compute it, because the same count divided by a
different base tells a different story.

```{r}
#| label: t1_categorical
cat_summary <- dta |>
  group_by(sex) |>
  summarise(
    diabetes_n   = sum(diabetes == "Yes"),
    diabetes_pct = 100 * mean(diabetes == "Yes"),
    .groups      = "drop"
  )
cat_summary
```

### Render with `gt()`

We join the two summaries and present the result with a header, relabelled
columns, and consistent numeric formatting. Mean ± SD pairs are formatted
to one decimal place; the diabetes percentage to one decimal place.

```{r}
#| label: t1_gt
table1 <- num_summary |>
  left_join(cat_summary, by = "sex") |>
  gt(rowname_col = "sex") |>
  tab_header(
    title    = "Table 1. Baseline Characteristics by Sex",
    subtitle = "Simulated cohort, n = 200"
  ) |>
  cols_label(
    n            = "N",
    age_mean     = "Age (mean)",
    age_sd       = "Age (SD)",
    bmi_mean     = "BMI (mean)",
    bmi_sd       = "BMI (SD)",
    gfr_mean     = "eGFR (mean)",
    gfr_sd       = "eGFR (SD)",
    diabetes_n   = "Diabetes (n)",
    diabetes_pct = "Diabetes (%)"
  ) |>
  fmt_number(
    columns  = c(age_mean, age_sd, bmi_mean, bmi_sd,
                 gfr_mean, gfr_sd, diabetes_pct),
    decimals = 1
  ) |>
  tab_spanner(label = "Age (years)",  columns = c(age_mean, age_sd)) |>
  tab_spanner(label = "BMI (kg/m²)",  columns = c(bmi_mean, bmi_sd)) |>
  tab_spanner(label = "eGFR (mL/min)", columns = c(gfr_mean, gfr_sd))

table1
```

The table renders as static HTML here, which is what `gt` is for. When the same
numbers have to reach a journal, hand the summary to
[manuscript tables](hv_tables.qmd) instead and let `hvtiRtables` place the
footnotes and the abbreviation key where the CORR rules want them.

## Pitfalls

- **Significance tests in Table 1.** It is tempting to add a p-value column
  comparing the groups, and many journals still ask for one. In a randomised
  trial that column is close to meaningless: any baseline difference is by
  construction due to chance, so testing it asks whether randomisation worked,
  not whether the groups differ in a way that matters. In an observational
  cohort the test is at least answering a real question, but a small p-value on
  a tiny clinically irrelevant difference (a half-year age gap in 5,000
  patients) misleads more than it informs. Describe the groups; let the reader
  judge balance from the numbers and the standardised differences.
- **Inconsistent rounding.** Decide on decimal places per variable and hold to
  it down the whole column. Age to one decimal in one row and whole numbers in
  the next reads as carelessness and makes the eye work harder. `fmt_number()`
  with a fixed `decimals` keeps a column honest; the recipe above pins the
  continuous summaries and the percentage to one decimal each.
- **Ambiguous denominators for percentages.** A percentage means nothing
  without its base. State whether it is computed within the column group, across
  the whole cohort, or among only the patients with a non-missing value, and be
  consistent. When missingness is common, report the n the percentage rests on
  so a reader does not assume a denominator you did not use.
