# Posters and research day

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

A poster is read from a metre away, not held at arm's length. Text that looks
generous on a manuscript page disappears on a poster panel. The fix is a larger
base size so axis labels, titles, and any on-panel series labels scale up
together, plus a panel content area sized to the physical poster layout rather
than a screen. `theme_hv_poster()` handles the typography; `hv_ggsave_dims()`
pins the data region to the dimensions you reserve on the poster.

Like the other house themes, `theme_hv_poster()` sets `legend.position = "none"`,
so a poster figure names its groups on the panel rather than in a key. At poster
scale that is not just convention: a reader standing a metre back cannot hold a
margin key in mind while tracking a curve. @sec-annotation-series has the recipe.

## A poster-styled plot

`theme_hv_poster()` uses a 16 pt base, and every text element scales from it, so a
single argument controls overall legibility.

```{r}
#| label: build-poster
p        <- plot(hv_survival(sample_survival_data(n = 500, seed = 42)))
p_poster <- p + theme_hv_poster()
p_poster
```

For an even larger viewing distance, bump the base: `theme_hv_poster(base_size
= 24)` enlarges the whole text hierarchy proportionally.

## Poster PDF

Poster figures are bigger than manuscript figures and usually slot into a fixed
panel of the poster template. As with manuscripts, fix the *panel content area*
rather than the file size: `hv_ggsave_dims()` returns `width`/`height`/`units`
so the data region matches the rectangle you allotted on the poster, expanding
the file to fit the (now larger) axis text and on-panel labels.

```{r}
#| label: save-poster
#| eval: false
dims <- hv_ggsave_dims(p_poster, width = 12, height = 9, units = "in")

do.call(
  ggsave,
  c(
    list(filename = "trends_poster.pdf", plot = p_poster,
         device = grDevices::quartz, type = "pdf"),
    dims
  )
)
```

The chunk is `eval: false` because it writes a binary PDF. The code is shown, not
executed during the book build.

The device is worth a word, because the obvious choice fails quietly. `cairo_pdf`
is the usual recommendation for embedding fonts, and on a Mac without XQuartz it
warns "failed to load cairo DLL", returns without error, and writes **no file at
all**. Measured on this machine: `ggsave()` reported success and the target PDF
did not exist. The quartz device is part of macOS, needs nothing installed,
embeds fonts, and handles the full Unicode range, which is why this book's own
PDF build uses it. The cost is that it is macOS only, so a poster built on Linux
needs `cairo_pdf` and a working cairo.

Calling `hv_ggsave_dims()` with the same `width`/`height` for every figure on the
poster keeps the data regions aligned across panels, regardless of how much room
each plot's labels consume.
