# Variable and partial dependence

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

[Variable importance](rf_vimp.qmd) ranks predictors but says nothing about *how*
a predictor moves the prediction. Once VIMP tells you bilirubin matters, the
next question is the shape of that effect: does predicted survival
fall steadily as bilirubin climbs, drop off a cliff past some threshold, or
level out at the high end? Dependence plots answer that. Reach for them whenever
you have a forest that predicts well and you need to describe what a specific
variable is doing, for a results figure, a clinical discussion, or your own
understanding of the model.

Two plots fill the gap, and the difference between them is the whole point of
this chapter:

- **Variable dependence** shows the forest's predicted value for every
  observation against that observation's actual predictor value. It is the raw,
  marginal picture, one point per patient, scatter and all.
- **Partial dependence** averages out every other predictor to isolate the
  effect of one variable. It is a smoother, model-level summary, the same trend
  with the confounding from other variables removed.

Both come from ggRandomForests [@R-ggRandomForests]: `gg_variable()` for the
first, `gg_partial_rfsrc()` for the second. Each hands back a bare ggplot you
finish with a house theme.

## The data it needs

Like VIMP, dependence plots read from a fitted `rfsrc` object, not a raw data
frame. We reuse the `pbc` survival forest from the
[variable importance](rf_vimp.qmd) chapter, fit on the primary biliary cirrhosis
cohort: one row per patient, a follow-up time, an event indicator, and a panel
of liver labs.

```{r}
#| label: rf-fit
data(pbc, package = "randomForestSRC")
rf <- rfsrc(Surv(days, status) ~ ., data = pbc, ntree = 100,
            importance = TRUE)
```

For a survival forest there is one extra ingredient: a follow-up time at which
to read the prediction. Survival probability is not a single number, it is a
curve over time, so dependence is always evaluated at a chosen horizon. The
median follow-up is a natural default.

```{r}
#| label: tval
t_eval <- median(pbc$days)
t_eval
```

## Build it

`gg_variable()` returns one predicted survival probability per observation. For
a survival forest the `plot()` method needs both a `time` to evaluate at and an
`xvar` selection naming which predictors to draw; without a time point it has
nothing to put on the y-axis. Setting `panel = TRUE` lays the chosen predictors
out side by side. Each point is one patient and the blue curve is a loess smooth
through the cloud.

```{r}
#| label: fig-rf_dependence
#| fig-cap: "Variable dependence for bilirubin, age, and albumin: predicted survival at the median follow-up against each patient's predictor value, with a loess smooth through the cloud"
#| fig-height: 4
gg_v <- gg_variable(rf, time = t_eval)
plot(gg_v, xvar = c("bili", "age", "albumin"), time = t_eval,
     panel = TRUE) +
  labs(y = paste0("Survival at ", t_eval, " days")) +
  theme_hv_manuscript()
```

The horizon goes on `gg_variable()`, not on `plot()`, and that is the easiest
thing to get wrong here. `gg_variable()` slices the predicted survival matrix at
a time when it builds the frame, and it defaults to the median of the forest's
own event-time grid (`median(rf$time.interest)`, which is `r median(rf$time.interest)`
days here) when you do not give it one. Passing `time` only to `plot()` therefore
selects a horizon that was never extracted, and you silently get the forest's
median instead of yours. The two differ: `t_eval` is `r t_eval`.

That `labs(y = ...)` is not decoration. `plot.gg_variable()` builds its survival
y-axis label as `paste("Survival at", time, "year")`, with the unit hardcoded, so
on `pbc`, whose follow-up is recorded in days, the default axis reads "Survival
at `r t_eval` year" for a horizon that is `r t_eval` **days**. Set the label
yourself on any survival dependence figure whose time is not in years.

Reading the smooths: predicted survival at `r t_eval` days falls steeply as
serum bilirubin rises, declines gently with age, and tracks upward with higher
albumin. These are the marginal trends actually present in the data, scatter
included, and that scatter is information: it warns you how much patient-to-patient
spread underlies each trend.

## Read it

The two plots answer different questions, and reading them well means keeping
the difference straight.

- **Variable dependence shows the raw trend, confounding and all.** The
  bilirubin panel in @fig-rf_dependence mixes bilirubin's own effect with the fact that
  high-bilirubin patients also tend to differ in age, albumin, and the rest. The
  scatter tells you how tight the relationship is once you account for that
  spread.
- **Partial dependence shows the adjusted trend.** The partial-dependence plot
  (@fig-rf_dependence-partial) holds the
  other predictors at their observed distribution and varies bilirubin alone, so
  the curve is bilirubin's marginal effect with the other variables averaged out.
  When the two plots disagree, the gap is the confounding.
- **The shape is the message.** Look for whether the curve is monotone (steady
  in one direction), thresholded (flat then steep, or steep then flat), or
  genuinely flat (the variable has little marginal pull at this time point even
  if VIMP ranked it). Report the time you read it at, because a different horizon
  can give a different shape.

To see partial dependence, `gg_partial_rfsrc()` drives randomForestSRC's
partial-prediction machinery directly. For a grid of `bili` values it predicts
every patient as if they had that value, then averages, giving the marginal
effect of `bili` with everything else held at its observed distribution. This is
the slow step, so we restrict it to a single variable at the median time and a
modest evaluation grid (`n_eval`) to keep the render quick.

```{r}
#| label: fig-rf_dependence-partial
#| fig-cap: "Partial dependence for bilirubin at the median follow-up: the adjusted dose-response curve with the other predictors averaged out"
gg_p <- gg_partial_rfsrc(rf, xvar.names = "bili",
                         partial.time = t_eval, n_eval = 25)
plot(gg_p) + geom_point() + theme_hv_manuscript()
```

The partial-dependence curve is the adjusted version of the bilirubin panel
in @fig-rf_dependence: the same downward trend, but now free of the influence of age, albumin,
and the rest. The flat scatter has collapsed into a single, interpretable
dose-response line, the cleanest summary of how bilirubin moves predicted
survival.

## Variations

### A different follow-up time

Because survival dependence is read at a horizon, the natural variation is to
read it at another one. Swap `t_eval` for an early landmark and the same
variable can show a flatter or steeper effect, since the predictors that drive
early events are not always the ones that drive late ones. Whenever you change
the time, change the caption too.

### Readable variable names

The panels above are labelled `bili`, `age`, and `albumin`, which is what the
data frame calls them and not what a manuscript reader calls them. Pass `labels`
a named character vector and the facet strips take the readable name instead.
Variables you do not name keep their raw column name, so you can label only the
ones a reviewer will see.

```{r}
#| label: fig-rf_dependence-labels
#| fig-cap: "The same variable dependence with reader-facing labels on the facet strips, showing age in the days the dataset actually stores rather than the years a reader would assume"
#| fig-height: 4
pbc_labels <- c(
  bili    = "Serum bilirubin (mg/dL)",
  age     = "Age (days)",
  albumin = "Serum albumin (g/dL)"
)

plot(gg_v, xvar = c("bili", "age", "albumin"), time = t_eval,
     panel = TRUE, labels = pbc_labels) +
  labs(y = paste0("Survival at ", t_eval, " days")) +
  theme_hv_manuscript()
```

Labelling is where unit errors surface, which is a good reason to do it early.
`age` in `pbc` is recorded in **days**, not years: the median is `r round(median(pbc$age))`,
which is about `r round(median(pbc$age) / 365.25)` years. A figure labelled
"Age (years)" would have been wrong on an axis nobody would think to re-check.

`labels` also accepts a labelled data frame, in which case it reads each
column's `label` attribute, or a two-column key/label frame. The same argument
works on `gg_udependent()`, `gg_sdependent()`, `gg_vimp()`, `gg_shap()`, and the
varPro plot methods, so one label vector can carry across every figure in a
manuscript.

### Lowering the cost of partial dependence

Partial dependence refits predictions across the whole grid, so it scales with
`n_eval`, the number of variables, and `ntree`. If a figure is too slow, lower
`n_eval` first (a coarser grid still shows the shape), then restrict to the one
or two variables you actually need to discuss, and only then reduce `ntree`.

## Pitfalls

- **Set the horizon on `gg_variable()`, not on `plot()`.** Predicted survival is
  undefined without a time, so `gg_variable()` slices the prediction matrix as it
  builds the frame. Give it no `time` and it quietly falls back to
  `median(rf$time.interest)`, the median of the forest's event-time grid, which is
  not the median follow-up and is rarely the horizon you meant. `plot()` then
  labels whatever was extracted, so a `time` passed only to `plot()` produces a
  figure that disagrees with your own caption. Pass it to both and they agree.
  `panel = TRUE` still needs its `xvar` selection.
- **Pick the time deliberately and report it.** Changing `time` or
  `partial.time` can change the shape of the curve. Choose a clinically
  meaningful horizon, state it, and do not quietly compare a figure read at the
  median with one read at five years.
- **Order of operations.** Use VIMP to choose *which* variables to profile here.
  There is no point spending a slow partial-dependence run on a variable the
  forest already told you it ignores.
