# Did the forest stabilize?

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

A forest prediction is an average over trees, so our first question is whether
that average has stopped changing materially as trees are added. `gg_error()`
shows the cumulative out-of-bag (OOB) error recorded while the forest grew. Use
it before reading an importance ranking or a predicted curve.

The display is a tree-count diagnostic. A roughly level tail says the fitted
forest is no longer sensitive to adding a few more trees. It does not prove
that the model transports to another population, that its predictors are
causal, or even that its absolute prediction error is acceptable.

## Fit a bounded teaching forest

We use Fisher's public iris data because a three-class outcome lets us inspect
the overall error and each class-specific error on the same object. The
teaching fit uses 300 trees and records error after every tree. A production
analysis should use an adequate tree count for its own cohort and verify the
plateau rather than copying 300 blindly.

```{r}
#| label: rf-fit
set.seed(20260828)
rf <- rfsrc(
  Species ~ .,
  data = iris,
  ntree = 300,
  block.size = 1
)
rf
```

`block.size = 1` matters here. Without it, `rfsrc()` may retain only a sparse
error history or the final value, leaving no trajectory to assess.

## Build and inspect it

The constructor returns one row per recorded tree. The `all` column is overall
misclassification error; the remaining columns are class-specific errors.

```{r}
#| label: rf-error-object
error_dta <- gg_error(rf)
error_dta

tail_summary <- data.frame(
  outcome = setdiff(names(error_dta), "ntree"),
  final_error = unlist(error_dta[nrow(error_dta),
                                  setdiff(names(error_dta), "ntree")]),
  mean_last_50 = colMeans(tail(
    as.data.frame(error_dta)[setdiff(names(error_dta), "ntree")], 50
  ))
)
tail_summary$final_error <- round(tail_summary$final_error, 3)
tail_summary$mean_last_50 <- round(tail_summary$mean_last_50, 3)
tail_summary
```

Now we plot the inspected object and add the house theme.

```{r}
#| label: fig-rf_error
#| fig-cap: "Cumulative OOB misclassification error by tree count, overall and for each iris class"
plot(error_dta) +
  theme_hv_manuscript() +
  labs(x = "Number of trees", y = "OOB misclassification error",
       color = "Outcome") +
  theme(legend.position = "bottom")
```

## Read it

The early path is noisy because only a few trees have contributed to each OOB
prediction. After that, the overall and class-specific curves fluctuate around
fairly stable levels. We do not require a monotone decline. Adding one tree can
move a cumulative OOB estimate up or down, especially for a small class.

Read the curves separately. A stable overall curve can hide a class whose error
is still moving, just as an overall event rate can hide a subgroup problem.
Here *setosa* settles quickly; *versicolor* and *virginica* carry most of the
remaining error. That class-specific pattern is more informative than the
overall line alone.

## Adapt it

For regression and survival forests, `gg_error()` usually returns one error
column plus `ntree`. The survival error is `1 - C`, where `C` is concordance;
lower is better. Classification returns overall and class-specific columns as
shown here. Always inspect `names(error_dta)` before deciding which trajectory
you are reporting.

If the right edge still has a sustained trend, refit with a larger explicit
`ntree` and inspect again. If it merely jitters around a level, summarize the
tail rather than declaring that every local rise means the forest has started
to fail.

One more reason to summarize the tail rather than read a single value. These
error rates are not stable across machines. Rebuilding the book on a Linux
runner moved the overall rate from 0.053 to 0.054 and versicolor's from 0.060
to 0.062, on the same seed and the same `randomForestSRC` version. Both of the
chapters that fit on `iris` do this, and the book's other forest chapters came
back unchanged from the same comparison, so it looks like a property of a
dataset measured to the nearest millimetre, where many values tie exactly and a
split has to break those ties somehow. Quote the level the trajectory settles
at, not the third decimal.

## Pitfalls

- **Calling a noisy curve proof of convergence.** We are checking practical
  tree-count stability. Repeated fits with different seeds can show whether the
  apparent plateau itself is stable.
- **Reading OOB as external validation.** OOB predictions reduce training
  optimism, but they do not establish transportability to another hospital or
  time period.
- **Reading a low error as a treatment effect.** This figure assesses a fitted
  prediction rule. It does not estimate a causal contrast.
- **Forgetting the outcome scale.** Misclassification error, mean squared
  error, and `1 - C` are different quantities. Name the forest family and the
  y-axis measure in the figure caption.
