# What does the forest predict?

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

After the tree-count diagnostic is stable, we can ask what the forest predicts.
For a survival forest the response is not one number. Each patient receives a
survival function, the predicted probability of remaining event-free beyond
each follow-up time.

Use `gg_rfsrc()` for model predictions and `gg_survival()` for the empirical
Kaplan-Meier estimate. They answer related but different questions. One shows
what the fitted forest predicts from all included covariates; the other
summarizes the observed time-to-event experience without that model.

## Fit a survival forest

The public `veteran` trial data contain follow-up time, death status, treatment
arm, and baseline predictors. We use 300 trees to keep the recipe bounded. A
production fit still needs its own tree-count stability check.

```{r}
#| label: rf-fit
data(veteran, package = "randomForestSRC")
veteran$trt <- factor(
  veteran$trt,
  levels = c(1, 2),
  labels = c("Standard", "Test")
)

set.seed(20260828)
rf <- rfsrc(
  Surv(time, status) ~ .,
  data = veteran,
  ntree = 300
)
rf
```

## Inspect OOB predicted survival

`oob = TRUE` is the constructor default for a training forest. Each patient's
curve is built from trees that did not use that patient to grow the tree.

```{r}
#| label: rf-predicted-object
pred_oob <- gg_rfsrc(rf, oob = TRUE)
pred_oob
head(as.data.frame(pred_oob), 4)

pred_check <- c(
  minimum = min(pred_oob$value),
  maximum = max(pred_oob$value),
  monotone_subjects = sum(vapply(
    split(pred_oob$value, pred_oob$obs_id),
    function(z) all(diff(z) <= sqrt(.Machine$double.eps)),
    logical(1)
  )),
  subjects = length(unique(pred_oob$obs_id))
)
pred_check
```

The inspected object is long: `obs_id` identifies the patient, `variable` is
the event-time grid, and `value` is predicted survival. Values stay between 0
and 1, and every subject's curve is non-increasing.

```{r}
#| label: fig-rf_predicted
#| fig-cap: "OOB predicted survival curves for the veteran cohort, one curve per patient"
plot(pred_oob, alpha = 0.18) +
  theme_hv_manuscript() +
  labs(x = "Follow-up time (days)", y = "Predicted survival probability") +
  coord_cartesian(ylim = c(0, 1))
```

Read the fan as predicted heterogeneity. Curves that fall earlier correspond to
higher predicted risk. They are not observed patient trajectories, and the
separation does not identify which covariate caused an outcome.

## OOB versus in-sample predictions

Setting `oob = FALSE` uses the full forest, including trees that may have seen
the patient during training. The two objects share the same scale, but the
full-forest values are in-sample descriptions and can look more decisive.

```{r}
#| label: rf-predicted-oob-check
pred_in <- gg_rfsrc(rf, oob = FALSE)
pred_in

data.frame(
  comparison = "Mean absolute difference",
  survival_probability = mean(abs(pred_oob$value - pred_in$value))
)
```

Use OOB predictions for an internal performance view. Use predictions from a
separate validation cohort when the question is transportability.

## Predicted and empirical survival by treatment arm

A grouped model display summarizes the patient-level predictions within each
arm. It is descriptive of this fitted forest, not an adjusted treatment effect
and not a hypothesis test.

```{r}
#| label: rf-predicted-group-object
pred_group <- gg_rfsrc(rf, by = "trt")
pred_group
head(as.data.frame(pred_group), 4)
```

```{r}
#| label: fig-rf_predicted-grouped
#| fig-cap: "Median OOB predicted survival within each veteran treatment arm"
plot(pred_group) +
  theme_hv_manuscript() +
  labs(x = "Follow-up time (days)", y = "Predicted survival probability",
       color = "Treatment arm", fill = "Treatment arm") +
  coord_cartesian(ylim = c(0, 1)) +
  theme(legend.position = "bottom")
```

For the observed comparison, `gg_survival()` handles right-censoring directly.
The `censor` argument names the event indicator here: `status = 1` is death and
`status = 0` is censored.

```{r}
#| label: rf-empirical-object
km <- gg_survival(
  interval = "time",
  censor = "status",
  by = "trt",
  data = veteran
)
km
head(as.data.frame(km)[c("time", "n", "dead", "surv", "groups")], 4)
```

```{r}
#| label: fig-rf_predicted-empirical
#| fig-cap: "Empirical Kaplan-Meier survival by veteran treatment arm"
plot(km) +
  theme_hv_manuscript() +
  labs(x = "Follow-up time (days)", y = "Observed survival probability",
       color = "Treatment arm", fill = "Treatment arm") +
  coord_cartesian(ylim = c(0, 1)) +
  theme(legend.position = "bottom")
```

The two displays should not be expected to coincide. The empirical curves use
only treatment groups; the forest predictions use the full predictor set and
then summarize those predictions by treatment.

## Pitfalls

- **Calling predictions outcomes.** `pred_oob$value` is a predicted survival
  probability. Death and censoring are stored separately in the observed data.
- **Ignoring the tail.** Late estimates depend on fewer patients who remain
  under follow-up. Read apparent separation there cautiously.
- **Calling a grouped prediction causal.** Averaging predictions by treatment
  does not control confounding, estimate a treatment effect, or test the arms.
- **Treating OOB as transport validation.** OOB predictions are internal. A new
  CORR era, registry, or institution remains a new-data question.
