# What if a predictor changes during follow-up?

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

Every survival recipe so far has assumed one row per patient. Baseline creatinine,
baseline ejection fraction, the covariate values as they stood on day zero, and a
single follow-up time. That assumption is usually a simplification we accept, and
sometimes it is the thing that ruins the analysis. A patient on temporary
mechanical support has a risk profile that changes week by week. Freezing them at
their admission values throws away the part of the record that was doing the work.

A Random Hazard Forest takes the counting-process form instead: several rows per
patient, each covering an interval of follow-up, with covariates allowed to change
between intervals. The forest sees what changed and when, so its predictions can be
read on the same clinical timeline as the measurements.

Reach for this when the exposure is time-varying and the timing matters, not just
its presence. Support escalation, a therapy that starts partway through admission, a
lab that drifts. If your predictors are genuinely fixed at baseline, the
[survival](survival.qmd) and [forest prediction](rf_predicted.qmd) recipes are the
simpler and better fit, and nothing here improves on them for that case.

This chapter reads a fitted forest. Growing one is the job of `randomForestRHF`
[@R-randomForestRHF]; the reading is done by `ggRandomForests`
[@R-ggRandomForests]. The boundary between the two is worth keeping straight, so
it has its own section near the end.

## The data it needs

The counting-process layout gives each patient one row per interval `(start, stop]`,
and `event` records whether the event happened at that interval's right endpoint. We
work from the reviewed bundle that ships with the `ggRandomForests` article, so the
chapter renders without refitting.

```{r}
#| label: bundle
bundle <- readRDS("data/rhf_precomputed.rds")

stopifnot(
  identical(unname(bundle$versions[["ggRandomForests"]]), "4.0.0"),
  identical(unname(bundle$versions[["randomForestRHF"]]), "2.0.0")
)

rhf_data <- bundle$data
one_subject <- rhf_data[rhf_data$id == rhf_data$id[1L],
                        c("id", "start", "stop", "event", "xtd")]
knitr::kable(one_subject, digits = 3,
             caption = "One simulated subject's counting-process records")
```

Read down the `xtd` column, then consider `x.1`, which is not shown because it never
moves. That is the whole distinction: `xtd` is the time-varying covariate and the
`x.1` through `x.10` columns are fixed at baseline. A counting-process frame carries
both, and the forest treats them differently only because the data says they behave
differently.

The rule that makes this legitimate is **predictability**: the value used at time
`t` must be known just before the event decision at `t`. A lab drawn after an
interval ends cannot be copied backward into it. In this simulation `xtd` is a
continuous function of `stop` and the subject's own fixed covariates, so its value at
`stop` is also its left-hand limit there, and it is available for the interval ending
at that time.

That is not a technicality. Copying a later measurement backward is how a model
learns to predict an event from evidence produced by the event, and it is the most
common way a time-varying analysis flatters itself. The no-lookahead routing is
enforced by `randomForestRHF::rhf()` when the forest is grown, not by anything in
this chapter.

```{r}
#| label: data-shape
records_per_subject <- as.integer(table(rhf_data$id))
events_by_subject   <- tapply(rhf_data$event, rhf_data$id, max)

data.frame(
  measure = c("Subjects", "Counting-process records", "Records per subject",
              "Subjects with an event"),
  value = c(length(unique(rhf_data$id)),
            nrow(rhf_data),
            sprintf("%d to %d (median %g)", min(records_per_subject),
                    max(records_per_subject), stats::median(records_per_subject)),
            sum(events_by_subject))
) |>
  knitr::kable(col.names = c("Measure", "Value"))
```

## Where these fitted objects came from

The bundle holds the fit and three upstream results that are far too slow to
recompute on every render. The calls that produced it are shown but not run.

```{r}
#| label: provenance
#| eval: false
set.seed(bundle$seed)

fit <- randomForestRHF::rhf(Surv(start, stop, event) ~ ., data = rhf_data)

auct_incident <- randomForestRHF::auct.rhf(fit, marker = "hazard")
importance    <- randomForestRHF::importance.rhf(
  randomForestRHF::varpro.cache.rhf(fit)
)
tune_risk     <- randomForestRHF::tune.treesize.rhf(rhf_data, metric = "risk")
tune_iauc     <- randomForestRHF::tune.treesize.rhf(rhf_data, metric = "iauc")
```

Supplying the saved results to the extractors is the pattern to copy for any analysis
you expect to revisit. Each `gg_*` function below will compute its own input when you
do not hand it one, which is convenient once and expensive every time after that.

## Read the hazard over time

`gg_rhf()` pulls both the pointwise hazard and the cumulative hazard into one tidy
frame, one row per subject per point on the forest's working time grid.

```{r}
#| label: fig-rhf-hazard
#| fig-cap: "Out-of-bag pointwise hazard for five of six requested subjects, each curve ending where that subject's follow-up ends; subject 22 has no grid point inside its records and so draws nothing"
case_ids   <- c(1, 2, 3, 11, 14, 22)
rhf_curves <- gg_rhf(bundle$fit)

p_hazard <- plot(rhf_curves, idx = case_ids) + theme_hv_manuscript()
hv_legend_inside(p_hazard)
```

All four house themes set `legend.position = "none"` on purpose: a CORR figure
carries no legend, and groups get named on the panel instead, which is the recipe
in @sec-annotation-series. This figure is the documented exception rather than a
departure from it. The six lines are simulated subject identifiers, not clinical
groups, so there is nothing meaningful to annotate beside each curve, and
[legends](legends.qmd) is explicit that you want a key whenever the grouping is
not already named on the panel. `hv_legend_inside()` is how you add one without
giving up panel width: apply it after the theme so its position wins, and it
anchors the key in whichever corner the data leave emptiest.

Hazard here is a local event rate: a high point says events are occurring at a high
instantaneous rate among the subjects still at risk around that time. It is not the
probability that this subject has the event then, and neither its height nor its
scale should be reported as a percentage. We plot six subjects because the extracted
object holds all of them and the panel becomes unreadable well before that.

The curves stop at different places, and that is deliberate rather than missing data.
From version 2.0.0, `randomForestRHF::rhf()` returns a hazard only where a grid point
falls inside one of that subject's supplied `(start, stop]` records, and `NA` in the
gaps and after the final stop. Each line therefore covers the range the forest was
actually asked about.

Count the legend and you get five, though we asked for six. Subject 22 has **no**
grid point inside any of its records, so every one of its hazard values is `NA` and
it draws nothing at all. That is the masking rule taken to its limit rather than a
plotting failure, and it is worth knowing that a requested subject can disappear from
this panel entirely. It comes back in the next one.

### Cumulative hazard accumulates the local rates

The same object carries the cumulative hazard. Set `hazard.only = FALSE` to draw it.

```{r}
#| label: fig-rhf-cumhaz
#| fig-cap: "Cumulative hazard for the same six subjects, all present this time, drawn across the whole grid because it accumulates interval overlap rather than being masked outside the records"
p_cumhaz <- plot(rhf_curves, idx = case_ids, hazard.only = FALSE) +
  theme_hv_manuscript()
hv_legend_inside(p_cumhaz)
```

Cumulative hazard is not masked the way the pointwise hazard is. It accumulates the
exact overlap between the working grid and the supplied records, so it stays flat
through a gap instead of going missing, and every grid point is drawn. Subject 22 is
back, as a flat line just above zero: it accumulated almost no hazard, which is a
different statement from having none to report.

The practical consequence is small and easy to trip over: if you summarise the
`hazard` column yourself, pass `na.rm = TRUE`. Counting subjects off a hazard panel
will also undercount them.

## Does discrimination hold up over time?

A single concordance number for the whole of follow-up hides the thing you usually
want to know, which is whether the model still separates patients late as well as it
did early. Time-varying AUC answers that, and which AUC you want depends on the
question.

`gg_auct()` with `marker = "haz"` gives the **incident/dynamic** definition: does a
subject who has the event near time `t` rank above a subject in the risk set at `t`?
That is a question about local failures, so it uses the local hazard marker.

```{r}
#| label: fig-rhf-auct
#| fig-cap: "Incident/dynamic time-varying AUC from the saved calculation, with the 0.5 chance line for reference"
auct_incident <- gg_auct(bundle$fit, marker = "haz", method = "incident",
                         auct_fit = bundle$auct_incident)

plot(auct_incident) + theme_hv_manuscript()
```

The saved curve has a Uno integrated AUC of
`r sprintf("%.3f", attr(auct_incident, "iauc")$uno)`, and its finite values run from
`r sprintf("%.3f", min(auct_incident$auc, na.rm = TRUE))` to
`r sprintf("%.3f", max(auct_incident$auc, na.rm = TRUE))`. AUC is a ranking
probability on a 0 to 1 scale, so 0.5 is the chance reference drawn on the panel.
This calculation was not bootstrapped, so there is no confidence ribbon; supply a
bootstrapped `auct_fit` and one appears.

The companion definition, **cumulative/dynamic**, asks whether a subject who has had
the event by a horizon ranks above one still event-free at that horizon, and its
matching marker is cumulative hazard. It targets a different quantity rather than a
better version of the same one. On in-sample data of this shape it is currently
awkward to read, because holding the cumulative hazard flat after follow-up ends
makes it track observation length as well as risk. Reach for it on genuinely
out-of-sample predictions, and read the incident/dynamic curve here.

## Which variables matter, and when?

Ordinary variable importance gives one number per predictor for the whole study. With
time-varying data that is the wrong shape: a covariate can dominate early and stop
mattering entirely once patients are past the risky window.

```{r}
#| label: fig-rhf-importance
#| fig-cap: "Time-localized variable priority across five time windows, with point size and colour showing the same priority value"
#| fig-width: 8
#| fig-height: 5.25
rhf_priority <- gg_rhf_importance(bundle$fit, importance_fit = bundle$importance)

windows <- unique(as.data.frame(rhf_priority)[, c("time_window", "start", "stop")])
windows <- windows[order(windows$start), ]
window_labels <- setNames(sprintf("%.2f to %.2f", windows$start, windows$stop),
                          windows$time_window)

plot(rhf_priority) +
  scale_x_discrete(labels = window_labels) +
  theme_hv_manuscript() +
  theme(legend.position = "right", plot.margin = margin(l = 80))
```

Two adjustments earn their place here. The window labels arrive as full-precision
interval strings such as `(0.00000000, 0.01055113]`, and five of those overlap into
an unreadable band along the axis, so we rebuild them from the `start` and `stop`
columns at two decimals. And the legend goes back on the outside rather than inside:
point size and colour both encode priority, so the key is a continuous scale rather
than a short list of series, and dropping it into the panel would sit on top of the
matrix it explains.

RHF priority is a time-local rule-release contrast. Within a window it compares the
fitted working response for a rule against its near-miss set, and asks how much that
response moves when rules involving a variable are released. Larger nonnegative values
mean a larger contrast.

Read that sentence again before reporting anything from this figure, because of what
it does **not** say. Priority does not tell you the direction of an effect, so it
never licenses "higher `xtd` raises hazard". It is not a z-score, not a p-value, and
carries no selection threshold. It ranks, and that is all.

In this simulation `xtd` carries the largest early-window value while `x.1` leads
several later windows, which is exactly the pattern a single whole-study importance
number would have averaged away. Note also that the number at risk falls from
`r max(rhf_priority$n_risk)` to `r min(rhf_priority$n_risk)` across the five windows,
so the late-window comparisons rest on much less information than the early ones. Any
window-to-window story should be told with that in view.

## Which tree size did the search pick?

`gg_tune_rhf()` reads a completed tuning path. It never reruns the search, and it
preserves the upstream evaluation order, so the connecting line follows the order
candidates were actually tried rather than a tidied-up sort.

```{r}
#| label: fig-rhf-tune-risk
#| fig-cap: "Out-of-bag risk tuning path, with the emphasized point marking the size the upstream search selected"
risk_tuning <- gg_tune_rhf(bundle$tune_risk)

plot(risk_tuning) + theme_hv_manuscript()
```

The risk search evaluated sizes `r paste(risk_tuning$treesize, collapse = ", ")` and
selected `r risk_tuning$treesize[risk_tuning$selected]`, where mean out-of-bag risk
was smallest. The saved risk path carries no standard-error field, and
`gg_tune_rhf()` sets `se` to missing rather than inventing uncertainty, so this plot
has no ribbon.

### Tuning by integrated AUC instead

```{r}
#| label: fig-rhf-tune-iauc
#| fig-cap: "The same tuning display for the integrated-AUC search, which selected a different tree size than the risk search did"
iauc_tuning <- gg_tune_rhf(bundle$tune_iauc)

plot(iauc_tuning) + theme_hv_manuscript()
```

This search evaluated `r paste(iauc_tuning$treesize, collapse = ", ")` and selected
`r iauc_tuning$treesize[iauc_tuning$selected]`, where out-of-bag iAUC was largest. The
two searches choose differently, which is the useful part: risk and discrimination are
not the same objective, and a tree size that minimises one need not maximise the
other. Say which criterion you tuned on when you report a size. This path has
`r sum(is.finite(iauc_tuning$se))` finite standard errors, so again no ribbon.

## What each extractor returns

Each `gg_*` function returns a data frame you can work with directly, which matters
when a figure is not the deliverable.

| Function | One row per | Key columns |
|---|---|---|
| `gg_rhf()` | subject and grid time | `id`, `time`, `hazard`, `chf`, `source` |
| `gg_auct()` | grid time | `time`, `auc`, `se`, `lower`, `upper`, `marker`, plus an `iauc` attribute |
| `gg_rhf_importance()` | variable and time window | `variable`, `time_window`, `time`, `n_risk`, `n_rules`, `priority` |
| `gg_tune_rhf()` | candidate tree size | `treesize`, `metric`, `value`, `se`, `selected` |

The `iauc` attribute on `gg_auct()` is the one people miss. It carries the Uno and
standardized integrated AUC, and `attr(x, "iauc")` is how you reach it for a table.

## Where ggRandomForests ends and randomForestRHF begins

This chapter draws pictures of decisions made elsewhere, and it is worth being precise
about which package owns what, so that an upstream modelling choice is not read as a
plotting defect.

`randomForestRHF` owns the model. It sets the working time grid, applies the
no-lookahead routing, decides that the pointwise hazard exists only inside supplied
records, computes the AUC under its chosen definition, evaluates variable priority,
and runs the tuning search. From 2.0.0 it also defaults to adaptive hazard
aggregation, which shifts fitted values relative to earlier versions.

`ggRandomForests` reads those results and tidies them. It passes the `NA` mask through
unchanged, drops those points at draw time so a curve ends with its subject's
follow-up, preserves the upstream tuning order, and declines to fabricate standard
errors the upstream object does not carry.

So a hazard curve that stops early, a missing late-window priority point, and a tuning
plot with no ribbon are all faithful reports of the fitted object. If you want
different behaviour there, the change belongs upstream.

## Pitfalls

- **A counting-process frame is not a wide longitudinal frame.** One row per interval,
  with `(start, stop]` half-open so consecutive intervals do not double-count the
  boundary. Reshaping one repeated-measures column per visit into `start`/`stop` rows
  is the step most likely to go wrong, and it goes wrong silently.
- **Predictability is the assumption that carries the analysis.** A value used at time
  `t` must be knowable just before `t`. Carrying a later lab backward into an earlier
  interval produces a model that predicts events from their own consequences, and
  reports excellent discrimination while doing it.
- **A subject can vanish from a hazard panel.** If no grid point falls inside any of
  its records, every hazard value is `NA` and the line is not drawn. The subject is
  still in the extracted frame and still appears in the cumulative-hazard panel. Never
  read the number of lines on a hazard figure as the number of subjects you asked for.
- **Hazard is a rate, not a probability.** Do not report a hazard height as a
  percentage, and do not compare hazard heights across figures drawn on different
  grids.
- **Priority ranks; it does not direct.** `gg_rhf_importance()` says a variable
  mattered in a window, never which way it pushed. Pair it with a dependence view
  before writing a direction into a manuscript.
- **Say which criterion you tuned on.** The risk and iAUC searches selected different
  tree sizes on the same data. A reported tree size is uninterpretable without its
  objective.
- **Supply the saved upstream objects.** `gg_auct()`, `gg_rhf_importance()` and
  `gg_tune_rhf()` will each compute their input when you omit it. That is fine once
  and painful in a document you re-render, which is why this chapter passes
  `auct_fit`, `importance_fit`, and a completed tuning path.
- **The house themes suppress legends deliberately, so add one deliberately.**
  All four `theme_hv_*()` functions set `legend.position = "none"` because a CORR
  figure names its groups on the panel rather than in a margin key. When the
  grouping genuinely cannot be named that way, restore the key after the theme:
  `hv_legend_inside()` for a handful of series, an explicit `legend.position` for
  a continuous scale that would occlude the data if placed inside.
- **Late windows rest on fewer patients.** The number at risk falls sharply across the
  importance windows here. Read late-window contrasts with that in mind rather than as
  equals of the early ones.
