# How well does the forest predict?

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
library(mlbench)
```

## When to use it

Model performance is not one question. For a binary outcome, an ROC curve asks
whether predicted scores rank events above non-events across thresholds. For a
right-censored survival outcome, a Brier curve asks how far predicted survival
probabilities sit from observed survival status over follow-up, with censoring
handled in the score.

Neither display validates transportability. And neither estimates a causal
treatment effect. They assess predictions from a fitted model.

## Discrimination for a binary outcome

We fit a classification forest to the public Pima diabetes data. The factor
levels are printed before we choose the positive class, because
`which_outcome = 2` means the second level, `pos`, in this fit.

```{r}
#| label: rf-classification-fit
data(PimaIndiansDiabetes, package = "mlbench")
levels(PimaIndiansDiabetes$diabetes)

set.seed(20260828)
rf_class <- rfsrc(
  diabetes ~ .,
  data = PimaIndiansDiabetes,
  ntree = 300
)
rf_class
```

The teaching value keeps execution bounded. Before reporting a study AUC, we
would verify the error trajectory with enough trees and evaluate the model on
the intended validation population.

## Build and inspect the ROC object

`gg_roc()` defaults to OOB class probabilities for an `rfsrc` training fit. Its
returned columns are sensitivity, specificity, and the threshold used at each
point.

```{r}
#| label: rf-roc-object
roc_dta <- gg_roc(rf_class, which_outcome = 2, oob = TRUE)
roc_dta
head(as.data.frame(roc_dta), 4)

roc_check <- data.frame(
  positive_class = levels(PimaIndiansDiabetes$diabetes)[2],
  auc = calc_auc(roc_dta),
  min_fpr = min(1 - roc_dta$spec),
  max_fpr = max(1 - roc_dta$spec),
  min_tpr = min(roc_dta$sens),
  max_tpr = max(roc_dta$sens)
)
roc_check[-1] <- lapply(roc_check[-1], round, digits = 3)
roc_check
```

```{r}
#| label: fig-rf_roc
#| fig-cap: "OOB ROC curve for positive diabetes, with false-positive rate on the x-axis and sensitivity on the y-axis"
plot(roc_dta) +
  theme_hv_manuscript()
```

The curve runs from the lower-left to the upper-right. The red 45-degree line
is chance ranking; a useful curve bows toward the upper-left. AUC is on the 0
to 1 scale, with 0.5 representing chance and 1 perfect ranking. It says nothing
by itself about whether a predicted probability of 0.8 occurs 80% of the time.

For a multi-class forest, request one class index at a time and name the
one-vs-rest comparison in the caption. Current `rfsrc` behavior does not build a
multi-class object from `per_class = TRUE`, so explicit class-specific calls are
the maintained, auditable route here.

## Time-resolved survival prediction error

Now we change both the outcome and the metric. The `veteran` data provide a
right-censored survival outcome. `gg_brier()` wraps the censoring-aware Brier
calculation from randomForestSRC and returns one score per event time.

```{r}
#| label: rf-survival-fit
data(veteran, package = "randomForestSRC")

set.seed(20260828)
rf_surv <- rfsrc(
  Surv(time, status) ~ .,
  data = veteran,
  ntree = 300
)
rf_surv
```

```{r}
#| label: rf-brier-object
brier_dta <- gg_brier(rf_surv, cens.model = "km")
brier_dta
head(as.data.frame(brier_dta)[c("time", "brier", "crps")], 5)

brier_check <- data.frame(
  time_min = min(brier_dta$time),
  time_max = max(brier_dta$time),
  brier_min = min(brier_dta$brier),
  brier_max = max(brier_dta$brier),
  censoring_model = attr(brier_dta, "cens.model"),
  integrated_crps = attr(brier_dta, "crps_integrated")
)
brier_check
```

```{r}
#| label: fig-rf_roc-brier
#| fig-cap: "OOB time-resolved Brier prediction error for the veteran survival forest; lower is better"
plot(brier_dta) +
  theme_hv_manuscript() +
  labs(x = "Follow-up time (days)", y = "Brier score")
```

The Brier score compares predicted survival probability with whether each
patient is event-free at that time. Because some outcomes are unknown after
censoring, the calculation uses inverse-probability-of-censoring weighting;
`cens.model = "km"` estimates that censoring distribution with Kaplan-Meier.
Whenever randomForestSRC can do so, it uses OOB values.

Lower is better, but the curve is not a calibration curve and not a
discrimination curve. It combines both properties of probabilistic prediction.
The far-right tail rests on very few patients with long follow-up, so we do not
read its drop toward zero as evidence of improving performance.
To study calibration, compare predicted and observed risks by horizon; to study
survival discrimination, use a censoring-aware concordance or time-dependent
ROC analysis.

## Pitfalls

- **Choosing the wrong positive class.** Print the factor levels, then map the
  numeric `which_outcome` index to the clinical event in the caption.
- **Reversing the ROC axes.** The plot uses false-positive rate (`1 -
  specificity`) on x and sensitivity on y. The chance line rises from `(0, 0)`
  to `(1, 1)`.
- **Calling AUC calibration.** AUC measures ranking discrimination.
- **Calling Brier score calibration alone.** Brier score is overall
  probabilistic prediction error, affected by calibration and discrimination.
- **Ignoring censoring and time units.** Report the IPCW censoring model and put
  the cohort's follow-up units on the x-axis.
- **Calling OOB external validation.** OOB metrics are internal estimates. They
  do not show that performance will carry to another registry or era.
