# SHAP attribution

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

The previous chapter ranked predictors by permutation importance: shuffle a
variable among the out-of-bag patients, see how much worse the forest predicts,
and the size of that drop is the variable's importance. One number per variable,
and it is a statement about the forest. VIMP answers "what is this model built
on?"

Then a different question arrives, usually from a clinician looking at one
patient's chart: this patient scored high, why? VIMP has nothing to say. It has
no per-patient parts to it.

SHAP does. Think of a SHAP attribution as an itemized receipt for a single
prediction. Start from the baseline, what the forest predicts about a patient it
knows nothing about, then credit or debit each variable with the amount it moved
that patient's prediction, and the line items add back up to the score the
patient actually got. Every patient gets their own receipt, so a variable that
lifts one patient's prediction can pull another's down.

So reach for VIMP when the question is about the model and SHAP when it is about
the individual. The two are not symmetric, which is worth knowing before you
pick: average the SHAP receipts over patients and you get a ranking that reads a
lot like VIMP, but you cannot go the other way and recover a patient from a VIMP
bar.

`gg_shap()`, from ggRandomForests [@R-ggRandomForests], computes those
attributions using the kernelshap package and hands back one row per
observation per variable. Three renderers draw it: `shap_importance()` for the
cohort ranking, `shap_beeswarm()` for the per-patient spread behind that
ranking, and `shap_dependence()` for one variable across its range. As
everywhere else in this part, each returns a bare ggplot you finish with a house
theme and the usual `+`.

## The data it needs

A fitted `rfsrc` object, with one restriction that shapes the rest of this
chapter: `gg_shap()` supports regression and classification forests. The `pbc`
survival fit the other recipes in this part lean on will not work here, and
handing one over gets you an error naming the family rather than a plot.

So we switch to `breast`, the Wisconsin breast cancer data that ships with
randomForestSRC and that the varPro chapter also uses: 32 tumour measurements
per patient and a two-level outcome, `N` for nonrecurrence and `R` for
recurrence.

```{r}
#| label: shap-fit
set.seed(42)
data(breast, package = "randomForestSRC")
dta <- na.omit(breast)
rf  <- rfsrc(status ~ ., data = dta, ntree = 200)

# newdata must carry predictors only -- see Pitfalls
nd <- dta[1:20, setdiff(names(dta), "status"), drop = FALSE]
gs <- gg_shap(rf, newdata = nd, bg_n = 20)
dim(gs)
```

Two arguments set both the cost and the meaning. `newdata` holds the patients
you want explained, twenty of them here, and `bg_n` is the size of the
background sample drawn from the training data that each attribution is measured
against. Twenty patients times 32 predictors is the 640 rows that come back, one
attribution apiece.

Because this is a classification forest, the attributions are on the predicted
probability of a single class, and `which.class = 1` picks the first one. Here
that is `N`, so everything below reads as movement in the predicted probability
of nonrecurrence. Pass `which.class = 2` to explain recurrence instead.

## Build it

`shap_importance()` collapses the receipts into a ranking. It takes the mean
absolute attribution for each variable across the explained patients and draws
the bars sorted, which gives you the SHAP counterpart to the VIMP figure in the
last chapter.

```{r}
#| label: fig-rf_shap-importance
#| fig-cap: "Global SHAP importance for the breast classification forest, ranking predictors by mean absolute attribution across the explained observations"
shap_importance(gs) + theme_hv_manuscript()
```

Two of the three variables at the top, tumour size and positive nodes, are the
staging measurements rather than the morphology panel, which is a reassuring
place for a breast recurrence model to land. Read the scale carefully: this is a
probability, so a bar at 0.03 says the variable moved the predicted probability
of nonrecurrence by about three percentage points on average, in one direction
or the other. Absolute value is doing work in that sentence, and it is exactly
what the next figure takes back apart.

## Read it

The mean absolute value that builds the ranking throws away two things a
clinician will ask about: the sign, and whether every patient looks alike.
`shap_beeswarm()` puts both back by plotting all 640 attributions, one point per
patient per variable.

```{r}
#| label: fig-rf_shap-beeswarm
#| fig-cap: "The per-observation attributions behind that ranking, one point per patient per variable, showing the spread a mean absolute value hides"
#| fig-height: 5.5
shap_beeswarm(gs) + theme_hv_manuscript()
```

The plot encodes two variables at once, and getting them straight is the whole
skill of reading it:

- **Horizontal position is the attribution.** Right of the dashed zero line, the
  variable pushed that patient toward nonrecurrence; left of it, away. Distance
  from the line is how hard it pushed.
- **Colour is the patient's own value of that variable**, rescaled low to high
  within each row. The legend reads Low to High, not in the variable's units,
  because each row is on its own scale.
- **The two together give you direction.** A row that runs high-value on the
  left and low-value on the right means the variable pushes the prediction down
  as it increases. That is what the top rows here do: patients with larger
  tumours and more involved nodes sit left of zero, away from nonrecurrence.
  A row where colour is scattered with no left-right order is the interesting
  case, since it means the variable's effect depends on what else the patient
  has.
- **The width of a row is heterogeneity.** A tight row acts on everyone the same
  way. A row that spreads wide is the forest treating patients differently, and
  that spread is precisely what the importance bar averaged away.

One caution when you export this figure. Twenty patients is a thin beeswarm, and
a row can look structured on twenty points and dissolve on two hundred. Read the
direction, not the individual points.

## Variations

### One variable across its range

The beeswarm compresses each variable into a strip. `shap_dependence()` pulls
one variable back out and plots each patient's attribution against their own
value of it, which is the closest SHAP gets to a partial dependence curve, built
per patient rather than per grid point.

```{r}
#| label: fig-rf_shap-dependence
#| fig-cap: "SHAP dependence for mean_radius, plotting each patient's attribution against their own value of the predictor"
shap_dependence(gs, xvar = "mean_radius") + theme_hv_manuscript()
```

This one comes back flat. The attributions scatter inside a couple of percentage
points of zero with no trend against radius, and that is an answer rather than a
failed plot: on these patients, this forest was not using `mean_radius`, which
is consistent with where it sits in the ranking above. Swap in a variable from
the top of that ranking, `pnodes` say, and the same call draws a clear downward
band, more involved nodes pulling the prediction away from nonrecurrence. Leave
`xvar` off entirely and you get the top-ranked variable by default.

For a factor predictor the same function draws boxplots by level instead of a
scatter, so the call does not change when the variable does.

## Pitfalls

- **`newdata` must hold predictors only.** Pass the fitted data frame whole,
  outcome column and all, and the call fails deep inside kernelshap with
  `all(colnames(X) %in% colnames(bg_X)) is not TRUE`, an error that names
  neither `newdata` nor the offending column. The fix is the
  `setdiff(names(dta), "status")` in the fit chunk above: drop the outcome, keep
  everything else.
- **Cost scales with `nrow(newdata)` times `bg_n`.** On this 32-predictor fit,
  twenty patients against a twenty-row background takes roughly 48 seconds;
  thirty against thirty takes roughly 99. Both arguments are levers, but they
  are not the same lever. Shrinking `newdata` explains fewer patients, while
  shrinking `bg_n` makes every attribution noisier, since the background is the
  reference the attribution is measured against. Cut `newdata` first, and keep
  `bg_n` as large as your patience allows.
- **Attribution is not causation.** A large SHAP value says the forest leaned on
  that variable for that patient. It does not say the variable caused the
  outcome, and it does not survive a change of model. Two correlated predictors
  will split their attributions the same way they split VIMP, so a variable can
  look quiet simply because a collinear partner absorbed the credit. Present
  SHAP as an explanation of the model, which is all it is, not as an effect
  estimate.
