# Which variables contribute to prediction?

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

Once a forest predicts adequately, a co-author usually asks which variables
carry that prediction. Permutation variable importance (VIMP) answers by
breaking one predictor's information in the OOB data and measuring how much
prediction error changes.

That is a model question. VIMP is not an effect size, does not give the
direction of association, and does not show that a variable causes the
outcome. Correlated predictors can share predictive information, so permuting
one may redistribute or understate the importance we would see if it stood
alone.

## Fit a classification forest

We again use iris so the returned object exposes both global and class-specific
VIMP. `importance = TRUE` asks `rfsrc()` to retain permutation importance. The
bounded 300-tree fit is for teaching; a study analysis should verify both the
error trajectory and the ranking's stability across an adequate forest.

```{r}
#| label: rf-fit
set.seed(20260828)
rf <- rfsrc(
  Species ~ .,
  data = iris,
  ntree = 300,
  importance = TRUE
)
rf
```

## Build and inspect it

```{r}
#| label: rf-vimp-object
vimp_dta <- gg_vimp(rf)
vimp_dta
head(as.data.frame(vimp_dta), 8)

with(as.data.frame(vimp_dta),
     aggregate(vimp, list(outcome = set), max))
```

The `vars` column holds raw predictor identities. `set = "all"` is the global
permutation score; the named class sets show the one-vs-rest contribution for
each outcome. Positive VIMP means permutation worsened prediction. Negative
VIMP means permutation improved it in this fit, often a sign of noise or an
unstable small contribution. The sign is not the direction of a clinical
association.

## Use display labels at plot time

Human-readable labels belong in the display, not in the model object. The
current plot method uses `labels`. We confirm that applying those labels leaves
the returned variables and their order unchanged.

```{r}
#| label: rf-vimp-label-check
feature_labels <- c(
  Sepal.Length = "Sepal length",
  Sepal.Width = "Sepal width",
  Petal.Length = "Petal length",
  Petal.Width = "Petal width"
)

raw_ids <- as.character(vimp_dta$vars)
vimp_plot <- plot(vimp_dta, labels = feature_labels)

data.frame(
  raw_names_unchanged = identical(raw_ids, as.character(vimp_dta$vars)),
  rows = nrow(vimp_dta),
  outcome_sets = length(unique(vimp_dta$set))
)
```

```{r}
#| label: fig-rf_vimp
#| fig-cap: "Permutation VIMP overall and by iris class; display labels do not alter the stored predictor names"
vimp_plot +
  theme_hv_manuscript() +
  scale_y_continuous(breaks = c(0, 0.4, 0.8), limits = c(0, 0.8)) +
  labs(x = "Predictor", y = "Permutation VIMP", fill = "VIMP > 0") +
  theme(axis.text.x = element_text(size = 7))
```

## Read it

Start with the `all` facet for the global ranking, then inspect whether the
class-specific facets tell the same story. Petal width and petal length carry
most of the prediction in this fit. A predictor can rank differently by class
because the forest may need it to distinguish one outcome more than another.

The object contains point estimates, not confidence intervals. Small gaps near
the bottom should not be treated as settled ranks. If the order matters to a
manuscript conclusion, refit across seeds or use a resampling analysis and
report that uncertainty separately.

## Adapt it

For a survival or regression forest there is usually one outcome set. For a
classification forest, inspect `unique(vimp_dta$set)` before deciding whether
to report the global ranking or a class-specific one. `gg_vimp(rf, nvar = 10)`
can reduce a long display, but keep the full ranking in a supplement when the
cutoff affects interpretation.

## Pitfalls

- **Reading sign as clinical direction.** Positive VIMP means worse prediction
  after permutation, not that larger predictor values increase risk.
- **Calling VIMP an effect size.** Scores depend on this forest, this outcome,
  this predictor set, and the prediction-error scale.
- **Ignoring correlated predictors.** Shared information can be split or
  redistributed across variables.
- **Treating labels as data changes.** Use `labels` in `plot()`. Keep the raw
  predictor names intact for downstream code and audit trails.
- **Claiming causality.** VIMP describes predictive contribution. It does not
  identify an intervention effect.
