# Which variables contribute to prediction?

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

Once a forest predicts adequately, a co-author usually asks which variables
carry that prediction. Permutation variable importance (VIMP) answers by
breaking one predictor's information in the OOB data and measuring how much
prediction error changes.

That is a model question. VIMP is not an effect size, does not give the
direction of association, and does not show that a variable causes the
outcome. Correlated predictors can share predictive information, so permuting
one may redistribute or understate the importance we would see if it stood
alone.

## Fit a classification forest

We again use iris so the returned object exposes both global and class-specific
VIMP. `importance = TRUE` asks `rfsrc()` to retain permutation importance. The
bounded 300-tree fit is for teaching; a study analysis should verify both the
error trajectory and the ranking's stability across an adequate forest.

```{r}
#| label: rf-fit
set.seed(20260828)
rf <- rfsrc(
  Species ~ .,
  data = iris,
  ntree = 300,
  importance = TRUE
)
rf
```

## Build and inspect it

```{r}
#| label: rf-vimp-object
vimp_dta <- gg_vimp(rf)
vimp_dta
head(as.data.frame(vimp_dta), 8)

with(as.data.frame(vimp_dta),
     aggregate(vimp, list(outcome = set), max))
```

The `vars` column holds raw predictor identities. `set = "all"` is the global
permutation score; the named class sets show the one-vs-rest contribution for
each outcome. Positive VIMP means permutation worsened prediction. Negative
VIMP means permutation improved it in this fit, often a sign of noise or an
unstable small contribution. The sign is not the direction of a clinical
association.

## Expect the numbers to move

Run this chapter on your own machine and the VIMP values will not match the
ones printed here. That is not a seeding mistake, and it is worth knowing
before you quote a number from a forest into a manuscript.

`set.seed()` fixes the forest for a given machine, and it holds. What it does
not survive is a change of platform. Rebuilding the book on a Linux runner
moved petal width's global score from 0.619 to 0.633 and reordered the rows
beneath it, with the same seed, the same `randomForestSRC` version and the same
R.

What moves is the forest, not the importance calculation. The plainest evidence
is [Did the forest stabilize?](rf_error.qmd), which grows the same iris
forest on the same seed and never asks for importance at all, and whose
out-of-bag error still shifts in the third decimal across the same two
machines. Nor is this a property of classification forests, since the
classification forest in [How well does the forest predict?](rf_roc.qmd)
reproduces exactly.

It is this dataset. Of the eight forests in the book, the two grown on `iris`
both move across platforms and the other six, on `pbc`, `veteran`,
`SynthDiabetes`, `breast`, `housing` and a simulated cohort, all reproduce to
the last decimal. `iris` is measured to the nearest millimetre, so its 150 rows
carry a great many exactly tied values, and a split has to break those ties
somehow. That is where we would look first, though we have not proved it.

The size of the shift is the useful part. Petal length and petal width sit far
above both sepal measurements in every run we have looked at, and that
separation is what @fig-rf_vimp is for. The order *within* a close pair is not,
at 300 trees. So read the grouping rather than the ranking. If an analysis
genuinely needs the ranking, grow the forest until the ranking stops moving,
and report how many trees that took.

## Use display labels at plot time

Human-readable labels belong in the display, not in the model object. The
current plot method uses `labels`. We confirm that applying those labels leaves
the returned variables and their order unchanged.

```{r}
#| label: rf-vimp-label-check
feature_labels <- c(
  Sepal.Length = "Sepal length",
  Sepal.Width = "Sepal width",
  Petal.Length = "Petal length",
  Petal.Width = "Petal width"
)

raw_ids <- as.character(vimp_dta$vars)
vimp_plot <- plot(vimp_dta, labels = feature_labels)

data.frame(
  raw_names_unchanged = identical(raw_ids, as.character(vimp_dta$vars)),
  rows = nrow(vimp_dta),
  outcome_sets = length(unique(vimp_dta$set))
)
```

```{r}
#| label: fig-rf_vimp
#| fig-cap: "Permutation VIMP overall and by iris class; display labels do not alter the stored predictor names"
vimp_plot +
  geom_hline(yintercept = 0, color = "grey45", linewidth = 0.35) +
  theme_hv_manuscript() +
  scale_y_continuous(
    breaks = function(x) pretty(x, n = 3),
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  labs(x = "Predictor", y = "Permutation VIMP", fill = "VIMP > 0") +
  theme(axis.text.x = element_text(size = 7))
```

## Read it

Start with the `all` facet for the global ranking, then inspect whether the
class-specific facets tell the same story. Petal width and petal length carry
most of the prediction in this fit. A predictor can rank differently by class
because the forest may need it to distinguish one outcome more than another.
The grey zero line separates positive from negative VIMP, and the axis range is
set from the returned scores so neither negative values nor larger positive
values are clipped.

The object contains point estimates, not confidence intervals. Small gaps near
the bottom should not be treated as settled ranks. If the order matters to a
manuscript conclusion, refit across seeds or use a resampling analysis and
report that uncertainty separately.

## Adapt it

For a survival or regression forest there is usually one outcome set. For a
classification forest, inspect `unique(vimp_dta$set)` before deciding whether
to report the global ranking or a class-specific one. `gg_vimp(rf, nvar = 10)`
can reduce a long display, but keep the full ranking in a supplement when the
cutoff affects interpretation.

## Pitfalls

- **Reading sign as clinical direction.** Positive VIMP means worse prediction
  after permutation, not that larger predictor values increase risk.
- **Calling VIMP an effect size.** Scores depend on this forest, this outcome,
  this predictor set, and the prediction-error scale.
- **Ignoring correlated predictors.** Shared information can be split or
  redistributed across variables.
- **Treating labels as data changes.** Use `labels` in `plot()`. Keep the raw
  predictor names intact for downstream code and audit trails.
- **Claiming causality.** VIMP describes predictive contribution. It does not
  identify an intervention effect.
