# Variable importance

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(randomForestSRC)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

You have fit a random forest and it predicts well, and now a co-author asks the
obvious next question: which variables are actually carrying the prediction?
That is what a variable importance plot answers. Reach for it whenever you want
to rank predictors by how much the forest leans on them, whether you are
trimming a wide cohort down to a workable model, sanity-checking that the
clinically expected markers rise to the top, or building the figure that opens
the results section of a CORR paper.

The standard random forest answer is **permutation variable importance** (VIMP).
The idea is simple to picture. For one predictor, take the out-of-bag patients
(the cases each tree never saw during fitting), shuffle that predictor's values
among them so it no longer lines up with the outcome, and re-run those patients
down the forest. If prediction error jumps, the forest was relying on that
variable; if error barely moves, the variable was doing little work. The size of
that jump is the variable's importance.

`gg_vimp()`, from ggRandomForests [@R-ggRandomForests], pulls the VIMP for every
predictor out of a fitted forest, sorts it, and hands back a tidy object whose
`plot()` method draws the ranking. As with the rest of the recipes here, what
comes back is a bare ggplot you finish with a house theme and the usual `+`.

## The data it needs

VIMP is computed from a fitted forest, so the real input is an `rfsrc` object,
not a data frame. The one thing to get right at fit time is importance itself:
pass `importance = TRUE` so the forest stores permutation VIMP as it builds.
Skip it and `gg_vimp()` still works, but it has to auto-calculate the scores on
the fly and warns you that it did so.

We use the primary biliary cirrhosis (`pbc`) survival data that ships with
randomForestSRC: one row per patient with a follow-up time, an event indicator,
and a panel of liver-function lab values. The outcome is time to death, so this
is a survival forest, but VIMP reads the same way for regression and
classification forests too.

```{r}
#| label: rf-fit
data(pbc, package = "randomForestSRC")
rf <- rfsrc(Surv(days, status) ~ ., data = pbc, ntree = 100,
            importance = TRUE)
rf
```

The summary confirms the family (survival) and the predictors the forest had to
work with. A small `ntree` keeps this example fast; for a real analysis you
would let it grow to the default and confirm the error curve has flattened
before trusting any ranking built on top of it.

## Build it

`gg_vimp()` extracts the VIMP for every variable and orders them from most to
least important. The `plot()` method draws a horizontal bar per predictor, so
the ranking reads top to bottom.

```{r}
#| label: fig-rf_vimp
#| fig-cap: "Permutation variable importance for every predictor in the pbc survival forest, ranked from most to least important"
plot(gg_vimp(rf)) + theme_hv_manuscript()
```

Bars are coloured by the sign of the importance. A handful of clinical markers,
bilirubin, albumin, prothrombin time, separate clearly from the pack, which is
exactly what we would expect for liver disease. Swap `theme_hv_manuscript()` for
`theme_hv_poster()` when you want the larger-text version.

## Read it

A VIMP plot is a ranking, and that framing is the whole skill of reading one.
Look for:

- **The break in the ranking.** Scan top to bottom for the gap where a few
  high-importance predictors give way to a long flat tail of near-zero
  variables. The handful above that gap are the ones the forest actually uses;
  the tail is mostly noise.
- **The sign of each bar.** Importance is colour-coded by sign. A negative VIMP
  means permuting the variable *improved* out-of-bag error, a fairly strong
  signal that the variable is noise the forest was better off ignoring.
- **Ranking, not effect size.** VIMP tells you *whether* a variable matters to
  the forest, never the *direction* or *shape* of its effect. Bilirubin sitting
  at the top says the forest leans on it heavily, not that higher bilirubin
  means worse survival (it does, but the plot does not show that). For the shape
  of an effect, see the [variable and partial dependence](rf_dependence.qmd)
  chapter.

## Variations

### Show only the top predictors

For a manuscript figure you rarely want every variable cluttering the panel. The
`nvar` argument keeps just the top-N, giving a clean, focused figure that puts
the eye straight on the predictors that matter.

```{r}
#| label: fig-rf_vimp-top
#| fig-cap: "The same variable importance ranking restricted to the top ten predictors for a cleaner manuscript figure"
plot(gg_vimp(rf, nvar = 10)) + theme_hv_manuscript()
```

Ten is a sensible default for a results figure; the full ranking is better kept
as a supplement so a reader can confirm nothing surprising hides in the tail.

## Pitfalls

- **Fit with `importance = TRUE`.** If you do not, `gg_vimp()` auto-calculates
  VIMP on the spot and warns you. The result is the same numbers, but the warning
  is easy to miss and the extra pass is wasted work. Set it at fit time and the
  scores are already there.
- **VIMP is relative ranking, not a causal claim.** A high bar means the forest
  found the variable useful for prediction given everything else in the model. It
  is not an effect size and it is not evidence that the variable *causes* the
  outcome. Two correlated predictors can split their importance between them, and
  dropping one can lift the other; read VIMP as a ranking within this forest, not
  a verdict on any single variable.
- **Stability.** VIMP inherits the forest's randomness, so adjacent
  low-importance variables can trade places between fits. Confirm the forest has
  converged (`gg_error()`) before you trust the ranking, and resist reading fine
  distinctions near the bottom of the list.
