# Alluvial and Sankey plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
library(ggalluvial)
library(ggsankey)
```

This chapter covers two figures that both answer the same kind of question:
where did the patients go? An alluvial diagram traces how a cohort redistributes
across a small number of clinical states. A cluster-stability Sankey traces how
a single cohort gets carved into more and more groups as you turn up the number
of clusters. They share a visual grammar (bands whose width is a count, flowing
left to right) but the data they want and the story they tell are different
enough that we treat them as two recipes under one roof.

The decision is in the unit of analysis. Use the alluvial recipe when each row
is an observed pathway with a count and you need to compare movement across
clinical states. Use the cluster-stability Sankey when each row is one patient,
each column is a value of K, and you need to see how cluster membership splits
as K increases.

## Alluvial plots

### When to use it

Reach for an alluvial diagram when you can describe each patient by two or three
categorical states measured in sequence, and the interesting thing is the
*movement* between them. The canonical CORR example is a valve cohort: an aortic
regurgitation (AR) grade before surgery, the procedure they received, and the AR
grade after. A bar chart of post-operative grades would tell you the endpoint;
an alluvial tells you who came from where. You can see at a glance that most of
the severe-AR patients ended up at none-or-mild, and you can see the handful who
did not.

`hv_alluvial()`, from hvtiPlotR [@R-hvtiPlotR], prepares the diagram on top of
ggalluvial [@R-ggalluvial]. Each row of the input is one unique combination of
the axis values together with a count, and `plot()` draws a band for that
combination sized in proportion to the count. As with the rest of the package,
`plot()` returns a bare ggplot you finish with the usual `+`.

### The data it needs

`hv_alluvial()` wants one row per unique combination of axis values, plus a
column holding the patient count for that combination. The `axes` vector names
the categorical columns in left-to-right display order; `y_col` names the count.
`sample_alluvial_data()` simulates 300 patients with pre-operative AR grade,
procedure type, and post-operative AR grade, already collapsed to the
one-row-per-combination shape the constructor expects.

```{r}
#| label: alluvial_data
dta_al  <- sample_alluvial_data(n = 300, seed = 42)
axes    <- c("pre_ar", "procedure", "post_ar")
head(dta_al)
```

### Inspect it

Build and inspect the returned `hv_alluvial` object before plotting. Its
metadata fixes the left-to-right axis order, while its prepared data holds one
row per pathway and the count that sets band width. The count audit should sum
to the cohort size, 300 here.

```{r}
#| label: alluvial_inspect
al <- hv_alluvial(dta_al, axes = axes, y_col = "freq")
al
al$meta[c("axes", "y_col", "n_obs")]
head(al$data)
sum(al$data$freq)
```

### Build it

Start from the bare panel so you can see what the constructor produced before
any styling. Build the object once, then call `plot()` on it.

```{r}
#| label: alluvial_bare
plot(al)
```

That is the raw material: bands connecting every level of `pre_ar` through
`procedure` to every level of `post_ar`, each band as wide as its count, and
nothing else. Now layer on the house style. Passing `fill_col` to the
constructor colours each band by a categorical column; here we colour by
pre-operative grade so you can follow a single starting grade across the whole
diagram. The matching `scale_colour_manual()` (with `guide = "none"`) keeps the
thin band outlines the same colour as their fill. We redraw the stratum labels
at a smaller size and leave strata under 10 patients unlabelled so the rare
post-operative grades do not print on top of one another. Their bands remain in
the count display.

```{r}
#| label: fig-sankey-alluvial
#| fig-cap: "Alluvial diagram tracing AR-grade transitions from before surgery, through procedure, to after, with bands coloured by pre-operative grade"
al_filled <- hv_alluvial(dta_al, axes = axes, y_col = "freq",
                         fill_col = "pre_ar")

plot(al_filled, show_labels = FALSE) +
  geom_text(
    stat = ggalluvial::StatStratum,
    aes(label = ifelse(
      after_stat(count) >= 10,
      as.character(after_stat(stratum)),
      ""
    )),
    size = 3
  ) +
  scale_fill_manual(
    values = c(None     = "steelblue",
               Mild     = "goldenrod",
               Moderate = "darkorange",
               Severe   = "firebrick"),
    name = "Pre-op AR"
  ) +
  scale_colour_manual(
    values = c(None     = "steelblue",
               Mild     = "goldenrod",
               Moderate = "darkorange",
               Severe   = "firebrick"),
    guide = "none"
  ) +
  scale_x_continuous(
    breaks = 1:3,
    labels = c("Pre-op AR", "Procedure", "Post-op AR"),
    expand = c(0.05, 0.05)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.02))) +
  labs(y = "Patients (n)",
       title = "AV Regurgitation: Pre- to Post-operative") +
  theme_hv_manuscript()
```

### Read it

An alluvial diagram rewards reading the bands, not the columns. A few things to
look for:

- **Follow one colour across the panel.** With `fill_col = "pre_ar"`, each
  colour is a starting grade. Trace the firebrick (severe) bands from left to
  right: where do they land at post-op? A figure where the severe colour
  concentrates at the none-or-mild end on the right is the story you want from a
  valve operation.
- **Band width is a count, not a rate.** A wide band is a common pathway; a
  hairline band is a rare one. Do not confuse a thin band for a small *effect*,
  it is a small *number* of patients taking that route.
- **Missing bands are missing combinations.** If there is no band from a given
  grade through a given procedure, that combination never occurred in the
  cohort. That absence is sometimes the finding (a procedure no one offered to
  the mild-AR group, say).

### Variations

#### Two-axis before / after comparison

When you only need to compare two time points, drop the middle axis and pass
just two columns to `axes`. Use `axis_labels` to replace the raw column names
with readable stage labels, and `annotate()` to call out the direction of
change directly on the panel. A diverging Brewer palette (`RdYlGn`, reversed)
maps naturally onto an ordered grade where green is good.

```{r}
#| label: fig-sankey-alluvial-twoaxis
#| fig-cap: "Two-axis alluvial comparing AR grade before and after surgery, with annotation marking the direction of improvement"
al2 <- hv_alluvial(
  dta_al,
  axes        = c("pre_ar", "post_ar"),
  y_col       = "freq",
  fill_col    = "pre_ar",
  axis_labels = c("Pre-operative", "Post-operative")
)

plot(al2, show_labels = FALSE) +
  geom_text(
    stat = ggalluvial::StatStratum,
    aes(label = ifelse(
      after_stat(count) >= 10,
      as.character(after_stat(stratum)),
      ""
    )),
    size = 3
  ) +
  scale_fill_brewer(palette = "RdYlGn", direction = -1,
                    name = "AR Grade") +
  scale_colour_brewer(palette = "RdYlGn", direction = -1,
                      guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.02))) +
  annotate("text", x = 1.5, y = 250,
           label = "Improvement after surgery",
           size = 3.5, fontface = "italic") +
  labs(y = "Patients (n)",
       title = "AV Regurgitation Before and After Surgery") +
  theme_hv_manuscript()
```

#### Milestone patient-flow with a clean y-axis

A patient-flow alluvial across clinical *milestones* — admission, a device
placement, its removal, discharge — is often read for its *shape*: where the
flow goes, not the exact count at each stage. When that is the focus, dropping
the count axis declutters the panel. It is a trade-off, though: as the pitfalls
below note, the axis (or the node counts) is what lets a reader recover absolute
numbers, so keep it whenever the denominators matter. Here the movement is the
story, so we pass `show_yaxis = FALSE`. One wrinkle: a full house theme re-styles
the axes it would reinstate, so when you finish with `theme_hv_manuscript()` you
blank the y elements again after it.

```{r}
#| label: fig-sankey-alluvial-milestones
#| fig-cap: "Milestone patient-flow alluvial across the mechanical circulatory support device course, with the count axis suppressed to emphasise pathway shape"
#| fig-width: 8
#| fig-height: 5
set.seed(1)
n_pt <- 444
flow <- data.frame(
  Admission = sample(c("MCS", "No MCS"), n_pt, replace = TRUE, prob = c(.48, .52)),
  Placement = sample(c("MCS", "No MCS"), n_pt, replace = TRUE, prob = c(.20, .80)),
  Removal   = sample(c("MCS", "No MCS"), n_pt, replace = TRUE, prob = c(.18, .82)),
  Discharge = sample(c("NHS", "HTx", "LVAD", "Dead"), n_pt, replace = TRUE,
                     prob = c(.30, .29, .16, .25))
)
flow_agg <- aggregate(list(freq = rep(1, n_pt)), flow, length)

al_ms <- hv_alluvial(
  flow_agg,
  axes     = c("Admission", "Placement", "Removal", "Discharge"),
  y_col    = "freq",
  fill_col = "Discharge"
)

plot(al_ms, show_labels = FALSE, show_yaxis = FALSE) +
  geom_text(
    stat = ggalluvial::StatStratum,
    aes(label = after_stat(stratum)),
    size = 3
  ) +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  scale_colour_brewer(palette = "Set1", guide = "none") +
  labs(title = "Mechanical circulatory support across the device course") +
  theme_hv_manuscript() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), axis.line.y  = element_blank())
```

Each band is one *pathway* — the patients who share the same admission-to-
discharge route — its width the number who took it (the aggregated `freq`), and
its colour where it ends. With the axis gone the figure reads as a flow, not a
stacked bar, which is the point of an alluvial in the first place.

## Cluster-stability Sankey plots

### When to use it

A clustering analysis leaves you with an awkward choice: how many clusters? Pick
two and you may be lumping distinct groups; pick nine and you may be splitting
noise. The cluster-stability Sankey turns that choice into a picture. It lines
up the cluster assignments at K = 2, K = 3, and so on across the panel, and
draws a band for every patient flow between consecutive K values. Where a single
cluster splits cleanly into two as K rises, you see one band fork into two
orderly bands. Where the partition is unstable, you see patients shuffling
sideways and bands crossing. The point where the orderly splitting gives way to
churn is the point where K has outrun the data.

`hv_sankey()` [@R-hvtiPlotR] builds this diagram on top of ggsankey
[@R-ggsankey]. It reads a block of cluster-assignment columns and works out the
flows for you; node labels show the cluster letter and its patient count.

### The data it needs

`hv_sankey()` wants one row per patient and one column per value of K, each
column holding that patient's cluster label at that K.
`sample_cluster_sankey_data()` returns 300 patients with nine such columns,
`C2` through `C9`, where `C2` is the PAM (partitioning-around-medoids) label at
K = 2 and so on up to K = 9. A quick `table()` on the last column confirms the
marginal cluster sizes before you build the figure.

```{r}
#| label: sankey_data
dta_san <- sample_cluster_sankey_data(n = 300, seed = 42)
head(dta_san)
table(dta_san$C9)
```

### Inspect it

Build the Sankey object and inspect its prepared flow data before plotting.
`node_levels` gives the vertical order used throughout the figure, and the
cross-tabulation checks every node count at every K. Each column must sum to
300, because the same synthetic patients are followed across all eight
partitions.

```{r}
#| label: sankey_inspect
sk <- hv_sankey(dta_san)
sk
sk$meta[c("cluster_cols", "node_levels", "n_patients", "n_k")]
xtabs(~ x + node, sk$data)
```

### Build it

The constructor finds the `C2` through `C9` columns on its own, so the default
build is a single call. It derives the node order from the data — seating each
cluster next to the parent it splits from — so the flows stay uncrossed and no
empty placeholder boxes appear, without you naming an order. We add a title and
the manuscript theme; the Set1 palette gives each cluster letter its own colour.

```{r}
#| label: fig-sankey
#| fig-cap: "Cluster-stability Sankey tracing how patient assignments split across K = 2 through K = 9"
#| fig-width: 8
#| fig-height: 5
plot(sk) +
  labs(title = "Cluster Stability: K = 2 to 9") +
  theme_hv_manuscript()
```

### Read it

Read this figure left to right, watching what happens to each band as K
increases:

- **Clean forks mean stability.** When one cluster at K splits into two at
  K + 1 and the two children stay intact afterward, that split is real. Bands
  that hold their width across several columns are the partitions you can trust.
- **Crossing and churn mean you have gone too far.** Once you see patients
  jumping between clusters that are not parent-and-child (bands crossing each
  other, narrow ribbons peeling off in both directions), the extra clusters are
  carving up noise. The last K before the churn starts is a defensible choice.
- **Node counts anchor the bands.** Each node is labelled with its size. A
  cluster that drops to a handful of patients as K grows is a sliver the
  algorithm split off, not a group you would report.

### Variations

#### Custom colour palette

Replace the default Set1 colours with a named vector. The names must match the
node labels (the cluster letters) in the data, so every letter from A through I
needs an entry.

```{r}
#| label: fig-sankey-custom-cols
#| fig-cap: "Cluster-stability Sankey with a custom named colour palette mapped to the cluster letters"
#| fig-width: 8
#| fig-height: 5
my_cols <- c(
  A = "#1f77b4", B = "#ff7f0e", C = "#2ca02c", D = "#d62728",
  E = "#9467bd", F = "#8c564b", G = "#e377c2", H = "#7f7f7f",
  I = "#bcbd22"
)
sk_custom <- hv_sankey(dta_san, node_colours = my_cols)
plot(sk_custom) +
  labs(title = "Cluster Stability: K = 2 to 9") +
  theme_hv_manuscript()
```

#### Subset of K values

Pass a shorter `cluster_cols` vector to the constructor to show only a range of
K. Once a stability plateau is clear, you rarely want the full K = 2 to 9 span
in a manuscript panel; trimming to K = 2 to 6 keeps the bands wide enough to
read in print.

```{r}
#| label: fig-sankey-subset
#| fig-cap: "Cluster-stability Sankey trimmed to K = 2 through K = 6 to keep the bands legible in a manuscript panel"
#| fig-width: 6
#| fig-height: 5
sk_sub <- hv_sankey(dta_san, cluster_cols = paste0("C", 2:6))
plot(sk_sub) +
  labs(title = "Cluster Stability: K = 2 to 6") +
  theme_hv_manuscript()
```

#### Milestone labels under the K axis

When the *number of distinct groups* changes at a particular column — most often
in a collapsed view, where several clusters merge into one — it helps to call
that out under the axis. `group_labels` takes a named vector mapping a column to
a label, and `plot()` prints it beneath that column's tick, leaving the others
bare. `flow_alpha` and `label_alpha` set the band and box transparency if a dense
panel needs a lighter touch.

```{r}
#| label: fig-sankey-milestones
#| fig-cap: "Cluster-stability Sankey with group-count milestone labels printed beneath selected K columns"
#| fig-width: 8
#| fig-height: 5
plot(
  hv_sankey(dta_san),
  group_labels = c(C2 = "2 groups", C4 = "4 groups",
                   C6 = "6 groups", C9 = "9 groups"),
  flow_alpha   = 0.5,
  label_alpha  = 0.3
) +
  labs(title = "Cluster Stability with group-count milestones") +
  theme_hv_manuscript()
```

The flow geometry is unchanged; only the x-axis gains the annotations under the
columns you name.

## Pitfalls

- **Width is count, not proportion.** Both figures size bands by raw patient
  counts. A wide band in a large cohort and a wide band in a small one mean
  different things; always show the axis (`Patients (n)`) or the node counts so
  the reader is not guessing.
- **Alluvial axis order is a choice.** The left-to-right order of `axes` is the
  reading order. Put the columns in clinical sequence (pre, procedure, post),
  not alphabetical or column order, or the flows will tell a scrambled story.
- **Colour the start, not the end.** Filling an alluvial by the *first* axis
  lets you trace where each starting group went. Filling by the last axis hides
  exactly the movement you built the figure to show.
- **A stable Sankey is not a correct K.** The stability plot tells you which K
  values the *data* supports geometrically. It does not tell you which K is
  clinically meaningful. Read it alongside the cluster profiles, not instead of
  them.
- **ggsankey is not on CRAN.** `hv_sankey()` needs the ggsankey package, which
  installs from GitHub (`remotes::install_github("davidsjoberg/ggsankey")`). If
  the Sankey chunks error on a fresh machine, that missing dependency is the
  first thing to check.
