# Scatter plots

Do two continuous measurements move together, and is their relationship straight
or curved? A scatter plot keeps every observation visible while a smoother gives
the cloud a direction.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

Before you fit anything, you want to look at the data. A continuous measurement
plotted against a time axis answers the most basic exploratory question we ask
of a longitudinal cohort: is this number drifting as the years go by? Reach for
the scatter plot whenever you have a continuous variable (ejection fraction, LV
mass, peak gradient) and a continuous reference axis (years since first surgery,
operation date) and you want to see the cloud of points and the trend through
them at the same time.

`hv_eda()` [@R-hvtiPlotR] is the exploratory workhorse. Point it at a column and
a reference axis and it picks the right display: when `y_col` is continuous it
returns a scatter plot with a LOESS smoother (locally weighted regression, a
flexible curve that follows the data without assuming a straight line) drawn
through the cloud. The same function draws bar charts for categorical columns,
but those live in their own chapter; here we stay with the single continuous
case. As with the rest of the package, `plot()` on the `hv_eda` object hands
back a bare ggplot. You finish it with colour scales, labels, and a house theme
using the usual `+`.

One detail that matters in practice: observations missing `y_col` are not
silently dropped. They are drawn as rug ticks along the x-axis, so you can see
where in time your measurement is missing rather than pretending the data are
complete.

## The data it needs

`hv_eda()` expects one row per observation with a continuous `x_col` (the
reference axis) and a continuous `y_col` (the measurement). Nothing else is
required. `sample_eda_data()` returns a mixed-type cohort of 300 patients;
here we use the continuous `op_years` (years since the patient's first surgery)
for x and `ef` (ejection fraction, the percent of blood the left ventricle
pumps with each beat) for y.

```{r}
#| label: eda_data
dta_eda <- sample_eda_data(n = 300, seed = 42)
head(dta_eda)
```

If you are not sure a column will be treated as continuous, ask the classifier
directly. `eda_classify_var()` reports `"Cont"` for continuous columns,
`"Cat_Num"` for numeric categories, and `"Cat_Char"` for character categories,
so you can confirm the plot type before you build it.

```{r}
#| label: eda_classify
sapply(dta_eda[c("op_years", "ef")], eda_classify_var)
```

Build the S3 object once so the bare, styled, and sensitivity views all use the
same prepared data.

```{r}
#| label: eda_object
eda_scatter <- hv_eda(
  dta_eda,
  x_col = "op_years",
  y_col = "ef",
  y_label = "Ejection Fraction (%)"
)
```

## Inspect it

The class selects `plot.hv_eda()`. Its metadata confirms the continuous display
and the observation count, while `rug_data` holds the missing-EF observations
that will appear as ticks.

```{r}
#| label: eda_inspect
class(eda_scatter)
eda_scatter$meta
head(eda_scatter$tables$rug_data)
```

## Build it

Start from the bare panel so you can see what the constructor produced before
any styling. Build the object with `x_col`/`y_col`; the `y_label` argument sets
the axis and legend text in place of the raw column name.

```{r}
#| label: eda_bare
p_scatter <- plot(eda_scatter)
p_scatter
```

That is the raw material: the scatter cloud, the LOESS smooth through it, and
the rug of missing-EF observations along the bottom. Now layer on the house
style, constrain the axes to a sensible range, and add a caption that tells the
reader what the x-axis ticks mean. There is no grouping aesthetic, so the panel
needs no legend.

```{r}
#| label: fig-scatter
#| fig-cap: "Ejection fraction against years from first surgery, with a LOESS smoother and a rug of observations missing EF"
p_scatter +
  scale_x_continuous(breaks = seq(0, 15, 5)) +
  scale_y_continuous(limits = c(20, 80), breaks = seq(20, 80, 20)) +
  labs(x = "Years from First Surgery Year",
       caption = "Tick marks on x-axis: observations with missing EF") +
  theme_hv_manuscript()
```

Swap in `theme_hv_poster()` when you want the larger, poster-sized text for
arm's-length reading.

## Read it

A scatter plot with a LOESS smooth gives you two things to read at once. Years
from first surgery are on the x-axis and ejection fraction is on the y-axis.
The default LOESS fit uses nearby observations to summarize the descriptive
association; it is not a patient-specific trajectory and does not show that time
since surgery caused EF to change. Look for:

- **The trend, not the points.** The smoother is the signal. Is the curve flat
  (the measurement is stable across the cohort), sloping (a drift you may need
  to adjust for), or bending (a non-linear pattern a straight line would miss)?
  In our example the LOESS line stays roughly level, which says ejection
  fraction is not drifting systematically with time since surgery.
- **The spread around the trend.** A tight cloud means the trend is well
  determined; a wide one means individual patients vary a lot and the smooth
  is an average you should not over-read.
- **The rug at the bottom.** Each tick is a patient whose EF is missing. If the
  ticks cluster in one region of time, your measurement is missing in a pattern,
  not at random, and the trend in that region rests on fewer points than the
  cloud suggests.

## Adapt it

### Tuning the LOESS smoother

`plot.hv_eda()` exposes `smooth_method` and `smooth_span` to control the
overlaid curve. `smooth_span` is the fraction of the data used in each local
fit: a smaller span follows local structure more closely (and wiggles more),
a larger span gives a flatter, more conservative trend. When you are unsure
whether a bump is real, compare two spans before you commit to a story.

```{r}
#| label: fig-scatter-span
#| fig-cap: "The same scatter with a narrower LOESS span (0.5), which follows local structure more closely than the default"
plot(eda_scatter, smooth_method = "loess", smooth_span = 0.5) +
  scale_x_continuous(breaks = seq(0, 15, 5)) +
  scale_y_continuous(limits = c(20, 80), breaks = seq(20, 80, 20)) +
  labs(x = "Years from First Surgery Year",
       y = "Ejection Fraction (%)") +
  theme_hv_manuscript()
```

## Deliver it

Move the final panel to [Finish and deliver the result](publications.qmd) for
manuscript, poster, or slide output. Keep the smoother method, span, and meaning
of the rug in the caption so the exported figure can stand on its own.

## Pitfalls

- **Reading the points instead of the trend.** With a few hundred patients the
  cloud looks busy and your eye finds patterns that are not there. The LOESS
  curve is what you report; the scatter is context.
- **Over-fitting the span.** A small `smooth_span` will track every local dip,
  including dips driven by one or two patients. If a feature disappears when you
  widen the span, it was probably noise.
- **Trusting the ends of the curve.** LOESS is least stable at the extremes of
  the x-axis, where it has data on only one side. Treat the leftmost and
  rightmost bends with the same caution you would give a sparse tail.
- **Ignoring the rug.** Missing values become rug ticks, not gaps in the trend.
  If the ticks pile up in one time window, the smooth there is built on thin
  data even though the line looks just as confident.
