# Spaghetti plots

How does each patient change across repeated measurements, and does the
cohort-level trend tell the same story? Spaghetti and trend plots answer those
two levels of the longitudinal question without pretending they are the same.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

You measured the same thing on the same patients more than once, and you want to
see how it moves. That is the case for a spaghetti plot. Each patient becomes one
line connecting their repeated measurements over time, and the tangle of lines (the
"spaghetti" the name promises) gives you the shape of the cohort: where the
values cluster, how much individual patients swing, whether the bundle drifts up
or down as the years pass. A serial echo measurement (aortic valve gradient,
valve area, regurgitation grade) is the natural candidate.

This chapter covers two related displays. The spaghetti plot, `hv_spaghetti()`,
keeps every patient's trajectory visible, so you see the spread as well as the
trend. Its companion, `hv_trends()`, throws the individual lines away and plots a
single summary per time point (a mean or median per year), which is what you want
when the question is the population-level drift and the per-patient detail is
noise. You reach for the spaghetti when individual variation is the story, and
for the trend curve when the annual summary is.

Both return a bare ggplot you finish with scales, labels, and a house theme.

## The data it needs

`hv_spaghetti()` expects long-format data: one row per measurement, with an id
column identifying the patient, a time column, and the value being tracked.
Optionally, a grouping column lets you colour trajectories by stratum.
`sample_spaghetti_data()` generates 150 patients with up to 6 observations each,
split by a named proportion vector that stands in for a sex stratification. Build
both the plain and the colour-stratified objects up front so the variants below
can reuse them.

```{r}
#| label: spaghetti_data
dta_sp <- sample_spaghetti_data(
  n_patients = 150,
  max_obs    = 6,
  groups     = c(Female = 0.45, Male = 0.55),
  seed       = 42L
)
head(dta_sp)

# Build S3 objects for reuse below
sp     <- hv_spaghetti(dta_sp)
sp_col <- hv_spaghetti(dta_sp, colour_col = "group")
```

`hv_trends()` works the other way: you hand it patient-level data and it computes
the annual summary internally, so you do not pre-aggregate. The `year_range` and
`groups` arguments set the study window and group structure.

## Inspect it

Both spaghetti objects use the `hv_spaghetti` plot method. Inspecting their
metadata confirms the patient identifier, observation count, and whether a
grouping colour was retained before either object is plotted.

```{r}
#| label: spaghetti_inspect
class(sp)
sp$meta
sp_col$meta
```

## Build it

Start from the bare spaghetti panel so you can see what the constructor produced
before any styling: one thin trajectory per patient over time, no colour, no axis
limits, no theme.

```{r}
#| label: spaghetti_bare
p_sp <- plot(sp)
p_sp
```

Now set sensible axis limits and breaks with the usual scale layers, then add a
theme. The y-axis here covers the observed range of AV mean gradient without
compressing the trajectories into the lower half of the panel.

```{r}
#| label: fig-spaghetti
#| fig-cap: "One trajectory per patient for AV mean gradient over five years of follow-up"
plot(sp) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 40, 10)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 40)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript()
```

## Read it

A spaghetti plot is read as a cloud first and individual lines second. Follow-up
years are on the x-axis and AV mean gradient is on the y-axis; each line is one
patient, so the number of observed trajectories at a time point is the effective
denominator. Compare the bundle across time, but do not treat the descriptive
paths as a treatment effect or assume that dropout is unrelated to the outcome.
Look for:

- **The shape of the bundle.** Where the lines are dense is where most patients
  sit; where they fan out is where patients diverge. A bundle that drifts upward
  across the x-axis is a cohort whose measurement is rising over time, even if no
  single line makes that obvious.
- **The crossing lines.** A patient whose trajectory cuts across the bundle, from
  the bottom to the top or back, is a patient whose value changed a lot. A few of
  these are normal; a whole sheaf of steep lines means the measurement is volatile
  and a single summary will hide more than it shows.
- **Where the lines stop.** Trajectories that end early are patients lost to
  follow-up. If the lines that drop out tend to sit high or low in the bundle,
  your later time points are a biased sample of the early cohort, and the apparent
  trend there may be attrition rather than change.

## Adapt it

### Spaghetti, with a LOESS overlay

When the bundle is too dense to read by eye, pass `add_smooth = TRUE` to overlay
a LOESS trend line per group (locally weighted regression, a flexible curve that
follows the data without assuming a straight line). The individual trajectories
stay as neutral context while the coloured smooths carry the group comparison.

```{r}
#| label: fig-spaghetti-smooth
#| fig-cap: "Stratified trajectories with a per-group LOESS overlay carrying the trend while the individual lines fade to context"
plot(sp_col, add_smooth = TRUE) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  guides(fill = "none") +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 40, 10)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 40)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

The smoother is fitted separately within each sex stratum. It summarizes the
observed trajectories, not within-patient causal change, and it can move when
the mix of patients still under observation changes.

### Making the trajectories recede behind the trend

A plain spaghetti plot needs no legend because every line represents the same
thing. When you add a LOESS overlay, use the constructor arguments to make the
patient trajectories light and thin while the smooth carries the comparison.
The caption names both layers, while line weight keeps the trend visually
separate without adding a legend to a one-group panel.

```{r}
#| label: fig-spaghetti-emphasis
#| fig-cap: "Patient trajectories in light grey with a wider LOESS trend carrying the cohort-level pattern"
plot(
  sp,
  add_smooth  = TRUE,
  line_colour = "grey70",
  line_width  = 0.3,
  alpha       = 0.4,
  smooth_width = 1.3
) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 40, 10)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 40)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript()
```

This layer comparison is carried by line weight and the caption, not colour
alone. The LOESS curve is still a descriptive smoother and cannot distinguish
true change from selective follow-up.

### Temporal trend: a single annual series

When the per-patient detail stops being the point, `hv_trends()` collapses the
cohort to one summary per year. Subset to a single group and fit with
`group_col = NULL` for a single annual series. `hv_trends()` plots the annual mean
of the continuous `value` column, so set the x-axis to the study window and let
the y-axis span the observed range of the mean (here roughly 20 to 70). One
detail to watch, called out again in the pitfalls below: setting `limits` tighter
than the data, for instance `c(0, 10)`, silently drops every point and line and
leaves a blank panel. The `c(0, 80)` limits here comfortably contain the series.

```{r}
#| label: trends_data
dta_tr <- sample_trends_data(
  n          = 600,
  year_range = c(1968L, 2000L),
  groups     = c("Group I", "Group II", "Group III", "Group IV")
)
head(dta_tr)
```

```{r}
#| label: fig-spaghetti-trend
#| fig-cap: "A single annual mean series from hv_trends() for one group across the study window"
one_grp <- dta_tr[dta_tr$group == "Group I", ]
tr1 <- hv_trends(one_grp, group_col = NULL)
class(tr1)
tr1$meta
head(tr1$tables$summary)

plot(tr1) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(0, 80),      breaks = seq(0, 80, 20)) +
  labs(x = "Year", y = "Annual mean outcome") +
  theme_hv_manuscript()
```

Year is on the x-axis and the annual mean is on the y-axis. Each point uses the
observations available in that year, and the LOESS curve smooths those annual
means. It describes the observed series; it does not show that calendar time
caused the outcome to change.

### Temporal trend: multiple groups

When `group_col` is set, the constructor computes per-group annual means and
`plot()` draws one line per group. Pairing `scale_colour_brewer()` with
`scale_shape_manual()` gives each group both a distinct colour and a distinct
marker, so the figure stays readable when it is printed or photocopied in
greyscale.

```{r}
#| label: fig-spaghetti-trend-groups
#| fig-cap: "Annual mean series for four groups, each given a distinct colour and marker so the figure survives greyscale printing"
tr <- hv_trends(dta_tr)
tr$meta
head(tr$tables$summary)
plot(tr) +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15L, "Group II" = 19L,
               "Group III" = 17L, "Group IV" = 18L),
    name = "Group"
  ) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Annual mean outcome") +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

### Temporal trend: with a confidence ribbon

Pass `se = TRUE` to add a `geom_ribbon()` around the mean line, and `alpha` to
control its opacity. The ribbon turns a bare trend line into an honest one: it
shows where the annual mean is well determined and where it rests on too few
patients to trust.

```{r}
#| label: fig-spaghetti-trend-ribbon
#| fig-cap: "The annual mean series with a confidence ribbon showing where the estimate is well determined and where it thins"
plot(tr1, se = TRUE, alpha = 0.2) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Annual mean outcome") +
  theme_hv_manuscript()
```

## Deliver it

Take the chosen longitudinal view to
[Finish and deliver the result](publications.qmd) for manuscript, poster, or
slide output. State the follow-up denominator, summary function, smoother, and
handling of dropout in the caption; those details determine what the trend can
support.

## Pitfalls

- **Axis limits that blank the panel.** This is the one that catches everyone.
  `scale_*_continuous(limits = ...)` does not zoom, it *filters*: any point
  outside the limits is dropped before plotting. Set the limits tighter than the
  data, say `c(0, 10)` on a series that runs to 70, and the trend line silently
  vanishes, leaving an empty panel with no warning. When a panel comes back blank,
  check the limits against the data range first. Use `coord_cartesian()` when you
  want to zoom without discarding points.
- **Too many lines.** Past a few hundred patients the spaghetti turns into a solid
  block and the individual trajectories stop being legible. At that point lower
  the line `alpha`, or switch to the trend curve, which is built for exactly this.
- **Reading a trend line without its ribbon.** A bare `hv_trends()` line looks
  equally confident at every year, but the early and late years of a study often
  rest on a handful of patients. Add `se = TRUE` so the reader can see where the
  estimate is thin.
- **Mistaking attrition for change.** Both displays connect only the patients who
  were measured. If sicker patients drop out, the surviving bundle drifts toward
  the healthy end, and the trend line follows. A trend is only a trend if the same
  kind of patient is being measured throughout.
