# Spaghetti plots

How does each patient change across repeated measurements, and does the
cohort-level trend tell the same story? Spaghetti and trend plots answer those
two levels of the longitudinal question without pretending they are the same.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

You measured the same thing on the same patients more than once, and you want to
see how it moves. That is the case for a spaghetti plot. Each patient becomes one
line connecting their repeated measurements over time, and the tangle of lines (the
"spaghetti" the name promises) gives you the shape of the cohort: where the
values cluster, how much individual patients swing, whether the bundle drifts up
or down as the years pass. A serial echo measurement (aortic valve gradient,
valve area, regurgitation grade) is the natural candidate.

This chapter covers two related displays. The spaghetti plot, `hv_spaghetti()`,
keeps every patient's trajectory visible, so you see the spread as well as the
trend. Its companion, `hv_trends()`, removes the individual lines, fits a LOESS
curve to the raw observations, and overlays one annual summary point per group.
Use it when the population-level pattern matters more than tracing individual
patients.

Both return a bare ggplot you finish with scales, labels, and a house theme.

## The data it needs

`hv_spaghetti()` expects long-format data: one row per measurement, with an id
column identifying the patient, a time column, and the value being tracked.
Optionally, a grouping column lets the smoother compare strata when
`add_smooth = TRUE`. In hvtiPlotR 2.7.10 the individual trajectory layer remains
black even when `colour_col` is set; the group colours apply to the smoother.
`sample_spaghetti_data()` generates 150 patients with up to 6 observations each,
split by a named proportion vector that stands in for a sex stratification. Build
both the plain and the group-aware objects up front so the variants below can
reuse them.

```{r}
#| label: spaghetti_data
dta_sp <- sample_spaghetti_data(
  n_patients = 150,
  max_obs    = 6,
  groups     = c(Female = 0.45, Male = 0.55),
  seed       = 42L
)
head(dta_sp)

# Build S3 objects for reuse below
sp     <- hv_spaghetti(dta_sp)
sp_col <- hv_spaghetti(dta_sp, colour_col = "group")
```

`hv_trends()` works the other way: you hand it patient-level data and it computes
the annual summary internally, so you do not pre-aggregate. The `year_range` and
`groups` arguments set the study window and group structure.

## Inspect it

Both spaghetti objects use the `hv_spaghetti` plot method. Inspecting their
metadata confirms the patient identifier, observation count, and whether a
grouping column was retained for the optional smoother before either object is
plotted.

```{r}
#| label: spaghetti_inspect
class(sp)
sp$meta
sp_col$meta
```

## Build it

Start from the bare spaghetti panel so you can see what the constructor produced
before any styling: one thin trajectory per patient over time, no colour, no axis
limits, no theme.

```{r}
#| label: spaghetti_bare
p_sp <- plot(sp)
p_sp
```

Now set sensible axis limits and breaks with the usual scale layers, then add a
theme. The y-axis here covers the observed range of AV mean gradient without
compressing the trajectories into the lower half of the panel.

```{r}
#| label: fig-spaghetti
#| fig-cap: "One trajectory per patient for AV mean gradient over five years of follow-up"
plot(sp) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 40, 10)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 40)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript()
```

## Read it

A spaghetti plot is read as a cloud first and individual lines second. Follow-up
years are on the x-axis and AV mean gradient is on the y-axis; each line is one
patient, so the number of observed trajectories at a time point is the effective
denominator. Compare the bundle across time, but do not treat the descriptive
paths as a treatment effect or assume that dropout is unrelated to the outcome.
Look for:

- **The shape of the bundle.** Where the lines are dense is where most patients
  sit; where they fan out is where patients diverge. A bundle that drifts upward
  across the x-axis is a cohort whose measurement is rising over time, even if no
  single line makes that obvious.
- **The crossing lines.** A patient whose trajectory cuts across the bundle, from
  the bottom to the top or back, is a patient whose value changed a lot. A few of
  these are normal; a whole sheaf of steep lines means the measurement is volatile
  and a single summary will hide more than it shows.
- **Where the lines stop.** Trajectories that end early are patients lost to
  follow-up. If the lines that drop out tend to sit high or low in the bundle,
  your later time points are a biased sample of the early cohort, and the apparent
  trend there may be attrition rather than change.

## Adapt it

### Spaghetti, with a LOESS overlay

When the bundle is too dense to read by eye, pass `add_smooth = TRUE` to overlay
a LOESS trend line per group (locally weighted regression, a flexible curve that
follows the data without assuming a straight line). The individual trajectories
stay as neutral context while the coloured smooths carry the group comparison.

```{r}
#| label: fig-spaghetti-smooth
#| fig-cap: "Black patient trajectories with coloured group-specific LOESS smoothers carrying the comparison"
plot(sp_col, add_smooth = TRUE) +
  scale_colour_manual(
    values = c(Female = "firebrick", Male = "steelblue"),
    name   = NULL
  ) +
  guides(fill = "none") +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 40, 10)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 40)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

The smoother is fitted separately within each sex stratum. It summarizes the
observed trajectories, not within-patient causal change, and it can move when
the mix of patients still under observation changes.

### Making the trajectories recede behind the trend

A plain spaghetti plot needs no legend because every line represents the same
thing. When you add a LOESS overlay, use the constructor arguments to make the
patient trajectories light and thin while the smooth carries the comparison.
The caption names both layers, while line weight keeps the trend visually
separate without adding a legend to a one-group panel.

```{r}
#| label: fig-spaghetti-emphasis
#| fig-cap: "Patient trajectories in light grey with a wider LOESS trend carrying the cohort-level pattern"
plot(
  sp,
  add_smooth  = TRUE,
  line_colour = "grey70",
  line_width  = 0.3,
  alpha       = 0.4,
  smooth_width = 1.3
) +
  scale_x_continuous(breaks = seq(0, 5, 1)) +
  scale_y_continuous(breaks = seq(0, 40, 10)) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 40)) +
  labs(x = "Years", y = "AV Mean Gradient (mmHg)") +
  theme_hv_manuscript()
```

This layer comparison is carried by line weight and the caption, not colour
alone. The LOESS curve is still a descriptive smoother and cannot distinguish
true change from selective follow-up.

### Temporal trend: a single annual series

When the per-patient detail stops being the point, `hv_trends()` replaces the
individual trajectories with a smoother plus one summary point per year. Subset
to a single group and fit with `group_col = NULL` for a single series. The points
are annual means of the continuous `value` column, but the LOESS curve is fitted
to the raw rows in `tr1$data`, not to those means. Set the x-axis to the study
window and let the y-axis span both layers (here roughly 20 to 70). One detail to
watch, called out again in the pitfalls below: setting `limits` tighter than the
data, for instance `c(0, 10)`, silently drops every point and line and leaves a
blank panel. The `c(0, 80)` limits here comfortably contain the series.

```{r}
#| label: trends_data
dta_tr <- sample_trends_data(
  n          = 600,
  year_range = c(1968L, 2000L),
  groups     = c("Group I", "Group II", "Group III", "Group IV")
)
head(dta_tr)
```

```{r}
#| label: fig-spaghetti-trend
#| fig-cap: "Annual mean points for one group with a LOESS curve fitted to the underlying raw observations"
one_grp <- dta_tr[dta_tr$group == "Group I", ]
tr1 <- hv_trends(one_grp, group_col = NULL)
class(tr1)
tr1$meta
head(tr1$tables$summary)

plot(tr1) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  scale_y_continuous(limits = c(0, 80),      breaks = seq(0, 80, 20)) +
  labs(x = "Year", y = "Outcome") +
  theme_hv_manuscript()
```

Year is on the x-axis and the outcome is on the y-axis. Each point is the mean of
the observations available in that year. The LOESS curve instead fits all raw
observations, so years with more rows can exert more influence than years with
fewer rows. Both layers describe the observed series; neither shows that
calendar time caused the outcome to change.

### Temporal trend: multiple groups

When `group_col` is set, the constructor computes per-group annual means for the
points and `plot()` fits a raw-observation smoother within each group. Pairing
`scale_colour_brewer()` with
`scale_shape_manual()` gives each group both a distinct colour and a distinct
marker, so the figure stays readable when it is printed or photocopied in
greyscale.

```{r}
#| label: fig-spaghetti-trend-groups
#| fig-cap: "Annual mean points and raw-observation LOESS curves for four groups, with distinct colours and point shapes"
tr <- hv_trends(dta_tr)
tr$meta
head(tr$tables$summary)
plot(tr) +
  scale_colour_brewer(palette = "Set1", name = "Group") +
  scale_shape_manual(
    values = c("Group I" = 15L, "Group II" = 19L,
               "Group III" = 17L, "Group IV" = 18L),
    name = "Group"
  ) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Outcome") +
  theme_hv_manuscript() +
  theme(legend.position = "right")
```

### Temporal trend: with a model-based smoother band

Pass `se = TRUE` to ask `geom_smooth()` for its model-based uncertainty band
around the LOESS fitted to raw observations, and use `alpha` to control the
band's opacity. This is not a confidence interval for the annual mean points and
does not directly display the number of observations available in each year.
Show yearly counts separately when that denominator matters.

```{r}
#| label: fig-spaghetti-trend-ribbon
#| fig-cap: "Annual mean points with a model-based uncertainty band around the LOESS fitted to raw observations"
plot(tr1, se = TRUE, alpha = 0.2) +
  scale_x_continuous(limits = c(1968, 2000), breaks = seq(1968, 2000, 4)) +
  labs(x = "Year", y = "Outcome") +
  theme_hv_manuscript()
```

## Deliver it

Take the chosen longitudinal view to
[Finish and deliver the result](publications.qmd) for manuscript, poster, or
slide output. State the follow-up denominator, summary function, smoother, and
handling of dropout in the caption; those details determine what the trend can
support.

## Pitfalls

- **Axis limits that blank the panel.** This is the one that catches everyone.
  `scale_*_continuous(limits = ...)` does not zoom, it *filters*: any point
  outside the limits is dropped before plotting. Set the limits tighter than the
  data, say `c(0, 10)` on a series that runs to 70, and the trend line silently
  vanishes, leaving an empty panel with no warning. When a panel comes back blank,
  check the limits against the data range first. Use `coord_cartesian()` when you
  want to zoom without discarding points.
- **Too many lines.** Past a few hundred patients the spaghetti turns into a solid
  block and the individual trajectories stop being legible. At that point lower
  the line `alpha`, or switch to the trend curve, which is built for exactly this.
- **Reading the smoother band as yearly precision.** With `se = TRUE`, the band
  belongs to the raw-observation LOESS fit. It is not a confidence interval for
  each annual mean and does not reveal the yearly denominator. Inspect or report
  annual counts separately.
- **Mistaking attrition for change.** Both displays connect only the patients who
  were measured. If sicker patients drop out, the surviving bundle drifts toward
  the healthy end, and the trend line follows. A trend is only a trend if the same
  kind of patient is being measured throughout.
