# Survival plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

The Kaplan-Meier curve is the workhorse of every CORR outcomes paper. Reach for
it whenever the question is *how long until an event* and some patients have not
had the event yet by the end of follow-up: death, reoperation, readmission.
That last part is what makes survival analysis its own method: a patient who is
alive at last contact is **right-censored**, not a non-event, and the
product-limit estimator is what lets those partial observations still count.

Before you choose a panel, name the question. These displays share a time axis,
but they do not answer the same thing:

- **Survival probability, $S(t)$:** What proportion remains alive and event-free
  beyond time $t$?
- **Cumulative hazard, $H(t)$:** How much event intensity has accumulated through
  time $t$? This is dimensionless and is not a probability.
- **Hazard rate, $h(t)$:** Among patients still at risk at time $t$, how quickly
  are events occurring? Its units are events per unit of person-time.
- **Absolute survival difference, $S_2(t)-S_1(t)$:** How many percentage points
  separate two groups at a named time? We build that contrast in the
  [NNT and survival-difference chapter](nnt.qmd).
- **Number needed to treat, $NNT(t)$:** How many patients must receive group 2
  rather than group 1 to prevent one event by time $t$? The direction and time
  horizon belong beside the number.
- **At-risk denominator, $n(t)$:** How many observed, event-free patients support
  the estimate immediately before time $t$? It tells you how much information
  remains; it is not another outcome curve.

`hv_survival()` [@R-hvtiPlotR], built on the survival package [@R-survival],
estimates the product-limit survival function once and stores everything five
related panels need (the survival curve, cumulative hazard, hazard rate,
log-log check, and integrated survivorship), plus tidy summary tables in
`$tables`. If the old program used `%kaplan`, `hv_survival()` is the maintained
R workflow. Build the S3 object once, inspect its data and tables, then call
`plot()` with a `type` argument to pull out the panel you want. Each panel comes
back as a bare ggplot you finish with the usual `+`.

## The data it needs

`hv_survival()` expects long-format data: one row per patient with a follow-up
`time` and an `event` indicator (`1` = event, `0` = censored). Nothing else is
required for an unstratified curve; add a grouping column when you want curves
per arm. `sample_survival_data()` builds a simulated right-censored cohort in
exactly this shape, so the recipe runs end to end before you point it at your
own data.

```{r}
#| label: km_data
dta_km <- sample_survival_data(n = 500, seed = 42)
head(dta_km)

# Build the S3 object once; reuse it across every panel below.
km <- hv_survival(dta_km, report_times = seq(0, 20, 5))
km$meta
km$tables$risk
```

## Build it

Start from the bare panel so you can see what the constructor produced before
any styling. The default `type = "survival"` is the curve with a logit-transform
95% confidence ribbon.

```{r}
#| label: km_bare
plot(km)
```

Then layer on the house style. The pattern below is the one we use most often in
CORR: a single steelblue for a one-cohort figure, percent labels on the y-axis,
an `n =` callout in the lower-left, and the manuscript theme. Swap in
`theme_hv_poster()` when you want the larger, poster-sized text.

```{r}
#| label: fig-survival
#| fig-cap: "Kaplan-Meier overall survival for the simulated cohort, styled in the CORR house style with a logit-transform 95% confidence ribbon"
plot(km, alpha = 0.8) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)",
       title = "Overall Survival") +
  annotate("text", x = 1, y = 5, label = paste0("n = ", nrow(dta_km)),
           hjust = 0, size = 3.5) +
  theme_hv_manuscript()
```

## Read it

A Kaplan-Meier curve starts at 100% and steps downward at each event time; it
never rises. A few things to look at:

- **Censoring changes the denominator, not the curve height.** A censor time
  removes that patient from later risk sets without creating a downward step.
  This simulated cohort has administrative censoring at 20 years, so the final
  at-risk count is the group still event-free when observation ends.
- **The ribbon widens to the right.** As the at-risk population thins, each
  remaining event moves the estimate more, so the confidence band fans out. Be
  cautious reading the far-right tail, which may rest on a handful of patients.
- **Median survival** is where the curve crosses 50%. If it never crosses, the
  median is not reached within follow-up and you report a different landmark
  (e.g. 5-year survival) instead.
- **The numbers at risk** below the figure tell the reader how much data backs
  each region of the curve. Always show them.

The at-risk strip and the per-time-point report (point estimates with CIs) are
stored on the object, so you never recompute them:

```{r}
#| label: km_tables
km$tables$risk
km$tables$report
```

## Numbers at risk

The "always show them" advice has a one-call answer. `hv_atrisk()` turns the
object's risk table into an aligned panel, and `hv_atrisk_compose()` stacks it
under the curve, matching the table's x-range to the plot's so the counts sit
beneath the time points they describe. Style the curve as usual, build the table
from the same object, and compose the two.

```{r}
#| label: fig-survival-atrisk
#| fig-cap: "Survival curve with an aligned numbers-at-risk table composed beneath the time axis"
#| fig-height: 5
curve <- plot(km, alpha = 0.8) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)") +
  theme_hv_manuscript()

risk_panel <- hv_atrisk(km)
hv_atrisk_compose(curve, risk_panel)
```

The counts under the axis are the same `km$tables$risk` numbers from above, now
positioned at the curve's 0-, 5-, 10-, 15-, and 20-year ticks. A stratified
figure gives the table one row per group, labelled to match the curves.

## Variations

### Stratified by group

To compare arms, pass `group_col` and the constructor fits a separate curve per
stratum. Here we simulate a two-arm valve-type cohort with a 1.4x hazard ratio.
Look for clearly separated curves whose CI ribbons stop overlapping in the
windows where the treatment difference matters.

```{r}
#| label: fig-survival-strata
#| fig-cap: "Stratified survival curves for a two-arm valve-type cohort simulated with a 1.4x hazard ratio"
dta_km_s <- sample_survival_data(
  n             = 500,
  strata_levels = c("Type A", "Type B"),
  hazard_ratios = c(1, 1.4),
  seed          = 42
)
km_s <- hv_survival(dta_km_s, group_col = "valve_type")

plot(km_s, alpha = 0.8) +
  scale_color_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  scale_fill_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(x = "Years after Operation", y = "Freedom from Death (%)",
       title = "Survival by Valve Type") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

### Cumulative hazard

`type = "cumhaz"` gives the Nelson-Aalen cumulative hazard, a monotonically
non-decreasing estimate of accumulated event intensity. Under the usual
survival relationship, $H(t) = -\log S(t)$; it is not $1-S(t)$ and it is not a
percentage. The local slope is the instantaneous hazard rate. Same object,
different `type`.

```{r}
#| label: fig-survival-cumhaz
#| fig-cap: "Nelson-Aalen cumulative hazard, a dimensionless measure of event intensity accumulated through each follow-up time"
plot(km, type = "cumhaz") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation", y = "Cumulative Hazard H(t)",
       title = "Nelson-Aalen Cumulative Hazard") +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  theme_hv_manuscript()
```

### Hazard rate

`type = "hazard"` is the instantaneous event rate among patients still at risk.
Because follow-up is measured in years, its units are events per patient-year.
The raw point estimates are noisy, so for a figure overlay a LOESS smooth. The
smoothed curve answers *when* events occur fastest; it does not report the
probability of an event at that time. Very short event intervals can produce a
few large point estimates, so the displayed y-range stops at 1.5 events per
patient-year to keep the smooth and most points readable.

```{r}
#| label: fig-survival-hazard
#| fig-cap: "Instantaneous event rate per patient-year with a LOESS smooth; the y-axis stops at 1.5 so a few unstable interval spikes fall outside the displayed range"
plot(km, type = "hazard") +
  geom_smooth(aes(x = mid_time, y = hazard, color = strata),
              method = "loess", se = FALSE, span = 0.6) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  scale_y_continuous(breaks = seq(0, 1.5, 0.5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 1.5)) +
  labs(x = "Years after Operation", y = "Hazard rate (events/patient-year)",
       title = "Hazard Rate") +
  theme_hv_manuscript()
```

### Log-log survival (proportional-hazards check)

`type = "loglog"` plots `log(-log S(t))` against `log(time)`. Parallel lines
across strata mean the proportional-hazards assumption is reasonable. Run this
check before you quote a single hazard ratio. Fit with `group_col` so
there are strata to compare.

```{r}
#| label: fig-survival-loglog
#| fig-cap: "Log-log survival plot for checking the proportional-hazards assumption across strata"
plot(km_s, type = "loglog") +
  scale_color_manual(
    values = c("Type A" = "steelblue", "Type B" = "firebrick"),
    name   = "Valve Type"
  ) +
  labs(x = "log(Years after Operation)", y = "log(-log S(t))",
       title = "Log-Log Survival") +
  theme_hv_manuscript()
```

### Integrated life table (restricted mean survival)

`type = "life"` is the integrated survivorship: the area under the survival
curve up to time `t`, equal to the restricted mean survival time (RMST) at that
horizon. It rises linearly while `S(t)` is near 1 and bends as events
accumulate. RMST is the summary to reach for when the log-log check above shows
hazards are *not* proportional and a single hazard ratio would mislead.

```{r}
#| label: fig-survival-life
#| fig-cap: "Integrated survivorship, the restricted mean survival time accumulated up to each horizon"
plot(km, type = "life") +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Years after Operation",
       y = "Restricted Mean Survival (years)",
       title = "Integral of Survivorship") +
  theme_hv_manuscript()
```

## Pitfalls

- **Event coding.** `event` must be `1` for the event and `0` for censored. Flip
  them and you estimate freedom from *being censored*, which is meaningless. Check
  the coding before you trust the curve.
- **Reading the tail.** The right end of the curve often rests on very few
  patients; the widening ribbon is the warning. Quote a landmark estimate (with
  its at-risk count) rather than the visual endpoint.
- **Non-proportional hazards.** If the log-log curves cross, a single hazard
  ratio is the wrong summary. Report RMST from the integrated panel instead.
