# Fit additive hazard models

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
library(TemporalHazard)
library(survival)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

Postoperative risk rarely stays constant. It may be high immediately after an
operation, settle to a low background rate after recovery, then rise again as
the patient and the repair age. A Kaplan-Meier curve tells us how much survival
remains. An additive hazard model asks the companion clinical question: which
early, constant, or late process is contributing risk at each time?

Begin with the empirical curves. They are the evidence for choosing a
parametric shape, not decoration added after the model is fit. If the empirical
hazard falls sharply and then flattens, one monotone Weibull may be a useful
summary. If it shows distinct regions, fit those phases explicitly and check
the total curve against the empirical estimate.

`TemporalHazard` [@R-TemporalHazard] is the maintained R implementation of the
Blackstone, Naftel, and Turner additive hazard model. It replaces the historical
SAS/C HAZARD workflow in this recipe. We fit and inspect the R object here; the
[prediction chapter](temporal_prediction.qmd) then turns it into curves.

## The data it needs

Both examples use public data bundled with `TemporalHazard`. `cabgkul` contains
5,880 primary isolated CABG patients, with follow-up in months and a death
indicator. `avc` contains 310 patients after atrioventricular canal repair and
supports the two clinically identifiable phases used below. No patient extract
is needed to run either recipe.

```{r}
#| label: load-data
data(cabgkul, package = "TemporalHazard")
data(avc, package = "TemporalHazard")
avc <- na.omit(avc)

data.frame(
  cohort = c("CABG", "AVC repair"),
  patients = c(nrow(cabgkul), nrow(avc)),
  deaths = c(sum(cabgkul$dead), sum(avc$dead)),
  maximum_follow_up_months = c(max(cabgkul$int_dead), max(avc$int_dead))
)
```

## Look at the empirical shape first

`hzr_kaplan()` estimates survival and `hzr_nelson()` estimates cumulative
hazard from the same subject-level records. Build both objects before choosing a
distribution, then inspect the columns that will be plotted.

```{r}
#| label: empirical-objects
km <- hzr_kaplan(cabgkul$int_dead, cabgkul$dead)
na <- hzr_nelson(cabgkul$int_dead, cabgkul$dead)

head(as.data.frame(km)[c("time", "n_risk", "n_event", "survival",
                         "cl_lower", "cl_upper")])
head(as.data.frame(na)[c("time", "n_risk", "n_event", "cumhaz",
                         "cl_lower", "cl_upper")])
```

The Kaplan-Meier panel shows the surviving fraction. The Nelson-Aalen panel
turns the same follow-up into accumulated event intensity. Its local slope is
the empirical clue to the hazard shape: changing slope means the instantaneous
risk is changing with time.

```{r}
#| label: fig-temporal-fit-kaplan
#| fig-cap: "Empirical Kaplan-Meier survival for the CABG cohort, with logit-transform 95% confidence limits"
ggplot(km, aes(time, survival * 100)) +
  geom_ribbon(aes(ymin = cl_lower * 100, ymax = cl_upper * 100),
              fill = "steelblue", alpha = 0.18) +
  geom_step(colour = "steelblue", linewidth = 0.7) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Months after CABG", y = "Freedom from death") +
  theme_hv_manuscript()
```

```{r}
#| label: fig-temporal-fit-nelson
#| fig-cap: "Empirical Nelson-Aalen cumulative hazard for the CABG cohort, with lognormal 95% confidence limits"
ggplot(na, aes(time, cumhaz)) +
  geom_ribbon(aes(ymin = cl_lower, ymax = cl_upper),
              fill = "firebrick", alpha = 0.18) +
  geom_step(colour = "firebrick", linewidth = 0.7) +
  labs(x = "Months after CABG", y = "Cumulative hazard, H(t)") +
  theme_hv_manuscript()
```

## Fit a reviewed Weibull model

Now fit the simple parametric description. The reviewed starting vector has a
scale and shape value; it is an optimizer starting point, not a result to
report. The estimates belong to the fitted object returned by `hazard()`.

```{r}
#| label: fit-weibull
fit <- hazard(
  time = cabgkul$int_dead,
  status = cabgkul$dead,
  theta = c(0.01, 1),
  dist = "weibull",
  fit = TRUE
)

summary(fit)
data.frame(
  converged = fit$fit$converged,
  log_likelihood = fit$fit$objective,
  function_evaluations = unname(fit$fit$counts["function"]),
  gradient_evaluations = unname(fit$fit$counts["gradient"])
)
```

The fitted Weibull shape is below 1, so its hazard declines with time. That is
consistent with an early postoperative component, but the single curve cannot
separate an early decline from a continuing background rate. Thus we treat it
as a compact population summary and keep the empirical panels beside it as the
fit check.

## Fit the early and constant phases

The AVC data have a clear early risk period and a much lower continuing rate.
We represent those with an early CDF phase plus a constant phase. The early
shape values below are the reviewed package example; `fixed = "shapes"` holds
them fixed while the model estimates the scale of each phase. We do not invent
new starting values.

```{r}
#| label: fit-multiphase
fit_multiphase <- hazard(
  survival::Surv(int_dead, dead) ~ 1,
  data = avc,
  dist = "multiphase",
  phases = list(
    early = hzr_phase(
      "cdf", t_half = 0.5, nu = 1, m = 1,
      fixed = "shapes"
    ),
    constant = hzr_phase("constant")
  ),
  fit = TRUE,
  control = list(n_starts = 3, maxit = 500)
)

summary(fit_multiphase)
fit_multiphase$fit$phase_share
```

Inspect convergence separately from the coefficient table. A populated summary
is not enough: the optimizer must converge, and the covariance calculation must
be positive definite for the free parameters.

```{r}
#| label: inspect-multiphase
data.frame(
  converged = fit_multiphase$fit$converged,
  positive_definite = fit_multiphase$fit$pd,
  reciprocal_condition = fit_multiphase$fit$rcond,
  log_likelihood = fit_multiphase$fit$objective
)

phase_scales <- exp(coef(fit_multiphase)[
  c("early.log_mu", "constant.log_mu")
])
phase_scales
```

The early scale is much larger than the constant scale, which is the clinical
story we expected for death after AVC repair: risk is concentrated near the
operation, then a low rate remains. The two phases overlap and add; the model
does not switch from one to the other at 0.5 months. Here `t_half = 0.5` shapes
the early contribution, while the fitted phase scales say how much each shape
contributes.

## Read convergence warnings before reading the curves

A multiphase model can return numbers even when the data do not identify the
requested phases. Stop and simplify the model when any of these occur:

- `fit$fit$converged` is `FALSE`, especially after the iteration limit.
- The covariance matrix is unavailable or `fit$fit$pd` is `FALSE`.
- A free parameter has an enormous standard error, or a fitted phase scale is
  effectively zero.
- Two phases trace nearly the same shape and compete to explain the same
  events.

Those are model warnings, not requests for a larger `maxit`. First ask whether
the empirical curve contains the phase at all. Fixing reviewed shape parameters
reduces the number of quantities the data must identify; adding an unsupported
late phase does the opposite.
