# Predict from a hazard model

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
library(TemporalHazard)
library(survival)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

An additive hazard fit becomes clinically useful when it tells us how risk
changes after the operation. Survival probability is the fraction of comparable
patients expected to remain alive beyond a named time. Cumulative hazard is the
event intensity accumulated through that time. Instantaneous hazard is the
conditional event rate among patients still at risk, with units of deaths per
patient-month here. It is not the observed number of deaths. The interval with
the most deaths also depends on how many patients remain at risk and how wide
the interval is, so it need not coincide with the peak hazard. In a multiphase
model, the hazard curve can also show which early or constant process is
contributing that rate.

This chapter refits every model it uses. You can copy it into a clean R session
without first running the [fitting chapter](temporal_fit.qmd). The working
pattern is the same throughout: fit the object, inspect it, predict on an
explicit time grid, then plot and read the result.

## Refit the models

The first fit is the reviewed intercept-only Weibull model for the public CABG
cohort. Follow-up is recorded in months, so every rate we derive from its time
grid is per patient-month.

```{r}
#| label: fit-models
data(cabgkul, package = "TemporalHazard")

fit <- hazard(
  time = cabgkul$int_dead,
  status = cabgkul$dead,
  theta = c(0.01, 1),
  dist = "weibull",
  fit = TRUE
)
summary(fit)

grid <- data.frame(
  time = seq(0.5, max(cabgkul$int_dead), length.out = 120)
)
```

## Predict survival and cumulative hazard

With `se.fit = TRUE`, current `predict.hazard()` returns the estimate,
delta-method standard error, and confidence limits. Survival limits stay
between 0 and 1; cumulative-hazard limits stay positive. Both predictions refer
to the same fitted curve on the same grid.

```{r}
#| label: predict-weibull
survival <- predict(
  fit, newdata = grid, type = "survival",
  se.fit = TRUE, level = 0.95
)
cumhaz <- predict(
  fit, newdata = grid, type = "cumulative_hazard",
  se.fit = TRUE, level = 0.95
)

pred <- data.frame(
  time = grid$time,
  survival,
  cumhazard = cumhaz$fit,
  cumhazard_lower = cumhaz$lower,
  cumhazard_upper = cumhaz$upper
)
head(pred)

# Survival and cumulative hazard must describe the same fitted model.
max(abs(pred$fit - exp(-pred$cumhazard)))
```

The last value should be zero to numerical precision because
$S(t)=\exp\{-H(t)\}$. If it is not, first check that both predictions came from
the same fit, profile, and time grid. That catches the common mistake of joining
predictions made on different rows.

### Survival against the empirical curve

The smooth curve is a model result, so we judge it against the Kaplan-Meier
estimate rather than letting it stand alone. The ribbon is uncertainty in the
parametric curve; the step line is the empirical reference.

```{r}
#| label: fig-temporal-prediction-survival
#| fig-cap: "Weibull survival with a 95% delta-method confidence band, overlaid on empirical Kaplan-Meier survival for the CABG cohort"
km <- hzr_kaplan(cabgkul$int_dead, cabgkul$dead)

ggplot() +
  geom_step(
    data = km,
    aes(time, survival * 100, colour = "Kaplan-Meier"),
    linewidth = 0.6
  ) +
  geom_ribbon(
    data = pred,
    aes(time, ymin = lower * 100, ymax = upper * 100),
    fill = "steelblue", alpha = 0.18
  ) +
  geom_line(
    data = pred,
    aes(time, fit * 100, colour = "Weibull"),
    linewidth = 0.9
  ) +
  scale_colour_manual(
    values = c("Kaplan-Meier" = "grey35", "Weibull" = "steelblue"),
    name = NULL
  ) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Months after CABG", y = "Freedom from death") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

The Weibull follows the broad decline but smooths over the changing early and
late behavior in the empirical curve. A narrow band does not rescue a
systematic shape mismatch; it only says the fitted parameters are estimated
precisely under that Weibull model.

### Accumulated event intensity

The Nelson-Aalen curve provides the corresponding empirical check for
cumulative hazard. $H(t)$ is dimensionless. Its slope, because time is in
months, has units of events per patient-month.

```{r}
#| label: fig-temporal-prediction-cumhaz
#| fig-cap: "Weibull cumulative hazard with a 95% delta-method confidence band, overlaid on the empirical Nelson-Aalen estimate"
na <- hzr_nelson(cabgkul$int_dead, cabgkul$dead)

ggplot() +
  geom_step(
    data = na,
    aes(time, cumhaz, colour = "Nelson-Aalen"),
    linewidth = 0.6
  ) +
  geom_ribbon(
    data = pred,
    aes(time, ymin = cumhazard_lower, ymax = cumhazard_upper),
    fill = "firebrick", alpha = 0.16
  ) +
  geom_line(
    data = pred,
    aes(time, cumhazard, colour = "Weibull"),
    linewidth = 0.9
  ) +
  scale_colour_manual(
    values = c("Nelson-Aalen" = "grey35", "Weibull" = "firebrick"),
    name = NULL
  ) +
  labs(x = "Months after CABG", y = "Cumulative hazard, H(t)") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

## Predict an additive hazard directly

Now refit the reviewed AVC model. It has an early CDF phase and a constant
phase, both on a monthly time scale. Current `predict.hazard()` supports direct
instantaneous-hazard prediction for this multiphase model. We use that API;
there is no need to approximate a derivative from cumulative-hazard values.

```{r}
#| label: fit-multiphase
data(avc, package = "TemporalHazard")
avc <- na.omit(avc)

fit_multiphase <- hazard(
  survival::Surv(int_dead, dead) ~ 1,
  data = avc,
  dist = "multiphase",
  phases = list(
    early = hzr_phase(
      "cdf", t_half = 0.5, nu = 1, m = 1,
      fixed = "shapes"
    ),
    constant = hzr_phase("constant")
  ),
  fit = TRUE,
  control = list(n_starts = 3, maxit = 500)
)
summary(fit_multiphase)

phase_grid <- data.frame(
  time = exp(seq(log(0.01), log(max(avc$int_dead)), length.out = 240))
)

total_hazard <- predict(
  fit_multiphase, newdata = phase_grid, type = "hazard",
  se.fit = TRUE, level = 0.95
)
total_cumhaz <- predict(
  fit_multiphase, newdata = phase_grid, type = "cumulative_hazard",
  se.fit = TRUE, level = 0.95
)
head(total_hazard)
```

The instantaneous hazard and its confidence limits have units of deaths per
patient-month. The cumulative hazard is dimensionless. Keep those labels
separate even though the two quantities come from the same fit.

## Rebuild and verify the phase contributions

`hzr_phase_hazard()` and `hzr_phase_cumhaz()` return unit-scale temporal
shapes. Multiply each by its fitted scale, `exp(log_mu)`, to recover that
phase's contribution. The helper signatures mirror the phase specification:
the early CDF uses `t_half`, `nu`, and `m`; the constant phase needs only its
type.

```{r}
#| label: phase-contributions
theta <- coef(fit_multiphase)
early_scale <- exp(theta["early.log_mu"])
constant_scale <- exp(theta["constant.log_mu"])

early_hazard <- early_scale * hzr_phase_hazard(
  phase_grid$time, t_half = 0.5, nu = 1, m = 1, type = "cdf"
)
constant_hazard <- constant_scale * hzr_phase_hazard(
  phase_grid$time, type = "constant"
)

early_cumhaz <- early_scale * hzr_phase_cumhaz(
  phase_grid$time, t_half = 0.5, nu = 1, m = 1, type = "cdf"
)
constant_cumhaz <- constant_scale * hzr_phase_cumhaz(
  phase_grid$time, type = "constant"
)

additivity_check <- data.frame(
  quantity = c("Instantaneous hazard", "Cumulative hazard"),
  maximum_absolute_difference = c(
    max(abs(total_hazard$fit - early_hazard - constant_hazard)),
    max(abs(total_cumhaz$fit - early_cumhaz - constant_cumhaz))
  )
)
additivity_check

stopifnot(
  additivity_check$maximum_absolute_difference[1] < 1e-12,
  additivity_check$maximum_absolute_difference[2] < 1e-12
)
```

Both differences should be at floating-point noise. If they are not, check the
phase type, fixed shape values, time units, and fitted scale before plotting.
The total must equal the phase sum at every time.

```{r}
#| label: fig-temporal-prediction-phase-hazard
#| fig-cap: "Fitted AVC hazard per patient-month: the total with its 95% confidence band and the additive early and constant contributions"
hazard_df <- rbind(
  data.frame(time = phase_grid$time, hazard = total_hazard$fit,
             component = "Total"),
  data.frame(time = phase_grid$time, hazard = early_hazard,
             component = "Early"),
  data.frame(time = phase_grid$time, hazard = constant_hazard,
             component = "Constant")
)
hazard_df$component <- factor(
  hazard_df$component, levels = c("Total", "Early", "Constant")
)

ggplot(hazard_df, aes(time, hazard, colour = component,
                      linetype = component)) +
  geom_ribbon(
    data = data.frame(phase_grid, total_hazard),
    aes(time, ymin = lower, ymax = upper),
    inherit.aes = FALSE, fill = "grey35", alpha = 0.14
  ) +
  geom_line(aes(linewidth = component)) +
  scale_colour_manual(
    values = c("Total" = "grey20", "Early" = "firebrick",
               "Constant" = "steelblue"),
    name = NULL
  ) +
  scale_linetype_manual(
    values = c("Total" = "solid", "Early" = "dashed",
               "Constant" = "dashed"),
    name = NULL
  ) +
  scale_linewidth_manual(
    values = c("Total" = 1, "Early" = 0.75, "Constant" = 0.75),
    guide = "none"
  ) +
  scale_x_log10(
    breaks = c(0.01, 0.1, 0.5, 1, 5, 20, 100),
    labels = c("0.01", "0.1", "0.5", "1", "5", "20", "100")
  ) +
  scale_y_log10(
    breaks = c(0.0005, 0.001, 0.005, 0.01, 0.05, 0.1, 0.5),
    labels = c("0.0005", "0.001", "0.005", "0.01", "0.05", "0.1", "0.5")
  ) +
  labs(x = "Months after AVC repair (log scale)",
       y = "Deaths per patient-month (log scale)") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

The log scales keep the first postoperative days and the low constant rate
visible on one panel. The early contribution dominates immediately after repair
and then falls. The constant contribution is small and flat. Their sum is the
solid total curve. If the phase order is clinically implausible, do not relabel
it into a better story. Return to the empirical curve and the phase
specification.

```{r}
#| label: fig-temporal-prediction-phase-cumhaz
#| fig-cap: "Fitted AVC cumulative hazard: the total with its 95% confidence band and the additive early and constant contributions"
cumhaz_df <- rbind(
  data.frame(time = phase_grid$time, cumhaz = total_cumhaz$fit,
             component = "Total"),
  data.frame(time = phase_grid$time, cumhaz = early_cumhaz,
             component = "Early"),
  data.frame(time = phase_grid$time, cumhaz = constant_cumhaz,
             component = "Constant")
)
cumhaz_df$component <- factor(
  cumhaz_df$component, levels = c("Total", "Early", "Constant")
)

ggplot(cumhaz_df, aes(time, cumhaz, colour = component,
                      linetype = component)) +
  geom_ribbon(
    data = data.frame(phase_grid, total_cumhaz),
    aes(time, ymin = lower, ymax = upper),
    inherit.aes = FALSE, fill = "grey35", alpha = 0.14
  ) +
  geom_line(aes(linewidth = component)) +
  scale_colour_manual(
    values = c("Total" = "grey20", "Early" = "firebrick",
               "Constant" = "steelblue"),
    name = NULL
  ) +
  scale_linetype_manual(
    values = c("Total" = "solid", "Early" = "dashed",
               "Constant" = "dashed"),
    name = NULL
  ) +
  scale_linewidth_manual(
    values = c("Total" = 1, "Early" = 0.75, "Constant" = 0.75),
    guide = "none"
  ) +
  labs(x = "Months after AVC repair", y = "Cumulative hazard, H(t)") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

The early cumulative contribution rises quickly and then approaches a plateau.
The constant contribution keeps accumulating linearly, even though its
instantaneous rate is low. Thus a phase can look small on the hazard panel and
still matter over long follow-up. Read the two panels together.
