# Move a legacy HAZARD analysis into R

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
library(TemporalHazard)
```

## When to use it

The destination is a maintained R analysis. `TemporalHazard`
[@R-TemporalHazard] can read the structure of a historical `PROC HAZARD` job,
translate the parts it recognizes, and import a legacy `OUTHAZ=` artifact for
a bounded comparison. Once we have checked the mapping, the R code, fitted
object, diagnostics, and predictions become the source we keep. A SAS license
or live SAS job is not part of the new workflow.

Think of the old program as a set of labeled boxes. `TIME`, `EVENT`, `PARMS`,
and the phase statements tell us what each box contained. The migration job is
to unpack those boxes once into named R arguments, inspect the result, and then
work in the current API.

The example below is synthetic and contains no patient data.

## Translate a short historical input

`hzr_translate_sas()` reads a file path, so we write the small text block to a
session temporary file. No legacy software is called. The parser reads the
statements and returns an `hzr_sas_job` containing the R expressions it would
place in a translated Quarto document.

```{r}
#| label: translate-historical-input
sas_text <- c(
  "PROC HAZARD DATA=AVC CONSERVE CONDITION=14;",
  "  TIME INT_DEAD;",
  "  EVENT DEAD;",
  "  PARMS MUE=0.20 NU=1;",
  "RUN;"
)

sas_path <- tempfile(fileext = ".sas")
writeLines(sas_text, sas_path)
job <- hzr_translate_sas(sas_path)

cat(paste(sas_text, collapse = "\n"))
data.frame(
  tokens_seen = job$coverage$tokens_seen,
  tokens_mapped = job$coverage$tokens_mapped,
  untranslated_constructs = nrow(job$untranslated),
  emitted_expressions = length(job$calls)
)
```

Coverage reports what the parser recognized, not whether a model has been
validated. For this bounded input all six tokens are mapped and no construct
is left untranslated. Now inspect the emitted fit call itself.

```{r}
#| label: inspect-translated-call
cat(paste(deparse(job$calls$fit, width.cutoff = 80), collapse = "\n"))
```

The generated call carries `TIME` to `time`, derives a 0/1 status from the
historical event count, preserves event counts as weights, turns the early
parameters into `hzr_phase("cdf", ...)`, and carries `CONSERVE` and
`CONDITION=14` into `control`. The returned job also contains a data guard and
status expression. Translation does not recreate a SAS `DATA` step, so you
must bind the reviewed R data frame before running the generated fit.

::: {.callout-warning}
## Read the generated code before you run it

The translator deliberately refuses some constructs, including a `SELECTION`
statement and censoring combinations that cannot be expressed by one current R
argument. An unresolved external fit or prediction grid can also stop the
generated document. Never replace that stop with guessed code. Rebuild the
unsupported step directly in R and document the choice.
:::

## Map the old names once

`hzr_argument_mapping()` exposes the package's formal crosswalk. Keep planned
rows visible during migration: the status column distinguishes a maintained
mapping from a broader legacy concept that still needs review.

```{r}
#| label: inspect-argument-map
argument_map <- hzr_argument_mapping(include_planned = TRUE)

argument_map[
  argument_map$sas_statement %in% c("TIME", "EVENT", "PARMS", "DIST", "G1", "G2"),
  c("sas_statement", "legacy_input", "r_parameter",
    "transform_rule", "implementation_status")
]
```

The general `PARMS` row remains marked as planned even though the translator
can parse the small early-phase subset above. Treat that as a limit on the
claim: the successful example is evidence for those statements, not blanket
support for every historical parameter combination. For a new analysis, write
the reviewed `hazard()` and `hzr_phase()` call directly rather than carrying
legacy names forward.

## Import an old `OUTHAZ=` artifact for a bounded check

The package ships a public synthetic fixture with the row layout of an
`OUTHAZ=` dataset. `hzr_read_outhaz()` reads that artifact into estimates,
free/fixed status, a covariance block for free parameters, and model-structure
flags. The same function can read a real `.sas7bdat` file when `haven` is
installed, but the fixture keeps this recipe public and reproducible.

```{r}
#| label: import-outhaz-fixture
outhaz_path <- system.file(
  "extdata", "outhaz-fixture.rds",
  package = "TemporalHazard"
)
legacy_fit <- hzr_read_outhaz(outhaz_path)

data.frame(
  fixture = basename(outhaz_path),
  parameter_estimates = length(legacy_fit$estimates),
  status_values = length(legacy_fit$status),
  free_parameters = nrow(legacy_fit$vcov),
  covariance_columns = ncol(legacy_fit$vcov),
  structure_flags = length(legacy_fit$flags)
)

data.frame(
  parameter = names(legacy_fit$estimates),
  estimate = unname(legacy_fit$estimates),
  free = unname(legacy_fit$status == 1L)
)
```

This import can verify that parameter names, stored estimates, fixed/free
status, covariance dimensions, and phase flags were carried into R. It cannot
recover the original analysis rows, the original log-likelihood, deletion
tallies, event-time coding, or every data transformation. It is not a refit and
cannot establish that a new model is correct merely because selected numbers
match.

Use the artifact once as a migration checksum. Then refit from the reviewed R
analysis data, run the diagnostics in [Check a hazard
model](temporal_diagnostics.qmd), save the R model object and code under version
control, and retire the legacy artifact from the active workflow.

## What the maintained record should contain

Before calling the migration complete, keep these pieces together:

- the reviewed R data-building code and a public or synthetic test example;
- the explicit `hazard()` model, phase definitions, starting values, and time
  units;
- any translator warnings and every manual resolution;
- the comparison quantities chosen in advance, with tolerances;
- the fitted R object, diagnostics, predictions, and session/package versions.

That record answers the question the old file cannot: not only whether a few
stored estimates matched, but exactly how the maintained analysis was built and
how we will rerun it next time.
