# Check a hazard model

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 150)
library(TemporalHazard)
library(survival)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

A smooth curve is not evidence of a faithful model. Once the optimizer has
converged, we still have five decisions to make. Does the fitted shape follow
the empirical survival? Is a continuous covariate sensible on the scale we
gave it? Does predicted risk agree with observed experience across the cohort?
Would variable selection or resampling tell a different story? And have we
defined competing outcomes with the right risk set?

These checks form a sequence, not a collection of stamps to put on a model.
Global misfit sends us back to the hazard shape. Curvature in a grouped
functional-form check sends us back to the covariate transformation. Poor
decile agreement sends us back to the linear predictor or baseline hazard.
Unstable selection or bootstrap estimates tell us to simplify the model or
collect more information. None of them turns an observational association into
a causal effect.

Everything here uses public data bundled with `TemporalHazard`
[@R-TemporalHazard]. The chapter is independently runnable and requires no
patient extract.

## Refit the reviewed model

The AVC example follows 305 complete records after atrioventricular canal
repair. Follow-up is in months. We refit the reviewed Weibull model so every
diagnostic below refers to one known object and one analysis cohort.

```{r}
#| label: fit-reviewed-model
data(avc, package = "TemporalHazard")
avc <- na.omit(avc)

fit <- hazard(
  survival::Surv(int_dead, dead) ~ age + status + mal + com_iv,
  data = avc,
  dist = "weibull",
  theta = c(mu = 0.20, nu = 1.0, rep(0, 4)),
  fit = TRUE,
  control = list(maxit = 500)
)

data.frame(
  records = nrow(avc),
  deaths = sum(avc$dead),
  maximum_follow_up_months = max(avc$int_dead),
  converged = fit$fit$converged,
  log_likelihood = fit$fit$objective
)
```

Stop here if `converged` is not `TRUE`. A diagnostic cannot repair a fit that
never reached its optimum.

## Does the overall shape fit?

`hzr_gof()` places the fitted Weibull survival beside the Kaplan-Meier
estimate and accumulates observed and expected events over follow-up. With
covariates in the model, the parametric curve is evaluated at the mean design
values. The Kaplan-Meier curve is marginal over the cohort. Thus this is a
first-pass shape and conservation check, not patient-specific calibration.

```{r}
#| label: inspect-gof
gof <- hzr_gof(fit)
gof_summary <- attr(gof, "summary")

data.frame(
  records = gof_summary$n,
  observed_events = gof_summary$total_observed,
  expected_events_at_mean_profile = gof_summary$total_expected,
  final_residual_expected_minus_observed = gof_summary$final_residual
)
```

```{r}
#| label: fig-temporal-diagnostics-gof
#| fig-cap: "Kaplan-Meier survival for the AVC cohort compared with Weibull survival evaluated at the mean covariate profile"
ggplot() +
  geom_step(
    data = gof,
    aes(time, km_surv * 100, colour = "Kaplan-Meier, cohort",
        linetype = "Kaplan-Meier, cohort"),
    linewidth = 0.75
  ) +
  geom_line(
    data = gof,
    aes(time, par_surv * 100, colour = "Weibull, mean profile",
        linetype = "Weibull, mean profile"),
    linewidth = 0.75
  ) +
  scale_colour_manual(
    values = c("Kaplan-Meier, cohort" = "grey35",
               "Weibull, mean profile" = "steelblue"),
    name = NULL
  ) +
  scale_linetype_manual(
    values = c("Kaplan-Meier, cohort" = "solid",
               "Weibull, mean profile" = "dashed"),
    name = NULL
  ) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(x = "Months after AVC repair", y = "Freedom from death") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

The curves differ, and the summary reports fewer expected events at the mean
covariate profile than were observed in the cohort. Those quantities have
different estimands: one is conditional on a single reference profile, while
the other is marginal over the cohort. The discrepancy therefore demonstrates
an estimand mismatch, not underprediction or a misspecified hazard shape. Before
concluding that the model is miscalibrated, standardize predictions over the
cohort or use a subject-specific check such as the decile observed-to-expected
comparison below.

## Is age sensible on the logit scale?

The name `hzr_calibrate()` comes from the historical grouped-screening
workflow, but its current contract is narrower than a fitted survival-model
calibration. It bins a continuous `x`, counts a binary `event` within each
group, and transforms that observed proportion. It does not use event time or
censoring. Here it asks whether eventual death is roughly linear in age on the
logit scale, which is a functional-form screen for the next model fit.

```{r}
#| label: inspect-age-groups
cal <- hzr_calibrate(
  x = avc$age,
  event = avc$dead,
  groups = 10,
  link = "logit"
)
cal
```

```{r}
#| label: fig-temporal-diagnostics-calibration
#| fig-cap: "Observed binary death proportion by age group, shown on the logit scale as a functional-form check"
ggplot(cal, aes(mean, link_value)) +
  geom_point(size = 2.4, colour = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.6) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = "grey35", linetype = "dashed", linewidth = 0.65) +
  labs(x = "Mean age at repair within group (months)",
       y = "Logit of observed death proportion") +
  theme_hv_manuscript()
```

Strong curvature would argue for a transformation, spline, or clinically
defined age categories before refitting. A straight pattern would support a
linear age term, but it would not prove that the full survival model is
calibrated.

## Does risk-group agreement hold?

`hzr_deciles()` asks a different question. It ranks all 305 patients by
predicted survival at 120 months and divides them into ten nearly equal groups.
The horizon defines the ranking only. Within each group, observed events use
each patient's full follow-up and expected events sum predicted cumulative
hazard at each patient's own follow-up time. No patient is excluded for having
less than 120 months of observation.

```{r}
#| label: inspect-risk-deciles
dec <- hzr_deciles(fit, time = 120, groups = 10)
dec_overall <- attr(dec, "overall")

data.frame(
  risk_horizon_months = dec_overall$time,
  included = dec_overall$n_included,
  excluded = dec_overall$n_excluded,
  observed_events = dec_overall$total_events,
  expected_events = dec_overall$total_expected,
  chi_square = dec_overall$chi_sq,
  degrees_freedom = dec_overall$df,
  p_value = dec_overall$p_value
)
dec
```

```{r}
#| label: fig-temporal-diagnostics-deciles
#| fig-cap: "Full-follow-up observed and expected event rates within groups ranked by predicted 120-month risk"
dec_long <- rbind(
  data.frame(group = dec$group, rate = dec$observed_rate * 100,
             series = "Observed"),
  data.frame(group = dec$group, rate = dec$expected_rate * 100,
             series = "Expected")
)

ggplot(dec_long, aes(group, rate, fill = series)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.7) +
  scale_fill_manual(values = c("Observed" = "grey55",
                               "Expected" = "steelblue"), name = NULL) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Predicted-risk group (1 = lowest)",
       y = "Full-follow-up events per 100 patients") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

The omnibus result is compatible with agreement across these groups, but a
non-significant test is not proof of calibration. Read the group sizes and the
observed-versus-expected pattern with the statistic. Sparse events and the
choice of groups both limit what this check can see.

## Does automated selection tell a stable story?

For the selection demonstration, we start from the reviewed early-plus-
constant model. The `scope` is phase-specific, so `age` can enter the early
phase without being forced into the constant phase. We request two-way
selection with the refit-based Wald criterion and keep the optimizer bounded.

```{r}
#| label: stepwise-selection
base_mp <- hazard(
  survival::Surv(int_dead, dead) ~ 1,
  data = avc,
  dist = "multiphase",
  phases = list(
    early = hzr_phase(
      "cdf", t_half = 0.5, nu = 1, m = 1,
      fixed = "shapes"
    ),
    constant = hzr_phase("constant")
  ),
  fit = TRUE,
  control = list(n_starts = 3, maxit = 500)
)

selection_warnings <- character()
selected <- withCallingHandlers(
  hzr_stepwise(
    base_mp,
    scope = list(
      early = ~ age + status + mal + com_iv,
      constant = ~ age + status + mal + com_iv
    ),
    data = avc,
    direction = "both",
    criterion = "wald",
    trace = FALSE,
    control = list(n_starts = 2, maxit = 500)
  ),
  warning = function(w) {
    selection_warnings <<- c(selection_warnings, conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)

trace <- stepwise_trace(selected)
cat(paste(trace, collapse = "\n"))

data.frame(
  final_converged = selected$fit$converged,
  final_positive_definite = selected$fit$pd,
  final_reciprocal_condition = selected$fit$rcond,
  accepted_steps = nrow(selected$steps),
  intermediate_ill_conditioned = sum(grepl("ill-conditioned", selection_warnings)),
  intermediate_not_positive_definite = sum(grepl("not positive-definite", selection_warnings)),
  intermediate_nonpositive_variance = sum(grepl("Non-positive variance", selection_warnings)),
  intermediate_unidentified_phase = sum(grepl("contributes at most", selection_warnings))
)
```

The trace names the criterion and records each accepted addition or removal.
This run adds five phase-specific terms. The final model converges with a
positive-definite covariance matrix, but intermediate candidate fits produce
ill-conditioned Hessians, non-positive variances, and one unidentified phase.
That instability is part of the selection result. The entry and stay thresholds
are a search rule, not a correction for repeated testing, and the selected path
is conditional on this sample. Preserve clinically required covariates with
`force_in`, examine the fit after every accepted step, and do not describe a
selected coefficient as a causal effect.

## How much do the estimates move under resampling?

`hzr_bootstrap()` resamples the analysis rows, refits the same Weibull formula,
and summarizes the fitted parameters among successful replicates. The seed
makes this teaching run reproducible. Thirty replicates are enough to exercise
the real refit path in a book render; they are not enough for study inference.

```{r}
#| label: bootstrap-stability
boot <- hzr_bootstrap(fit, n_boot = 30, seed = 20260828)

data.frame(
  requested_replicates = 30,
  successful_replicates = boot$n_success,
  failed_replicates = boot$n_failed,
  parameter_rows = nrow(boot$replicates)
)
boot$summary
```

```{r}
#| label: fig-temporal-diagnostics-bootstrap
#| fig-cap: "Teaching-run percentile intervals for the four Weibull covariate coefficients across 30 bootstrap refits"
boot_beta <- subset(
  boot$summary,
  parameter %in% c("age", "status", "mal", "com_iv")
)
boot_beta$parameter <- factor(
  boot_beta$parameter,
  levels = boot_beta$parameter[order(boot_beta$mean)]
)
boot_beta$parameter <- factor(
  boot_beta$parameter,
  levels = c("age", "mal", "status", "com_iv"),
  labels = c("Age (per month)", "Malalignment", "NYHA status (per class)",
             "Interventricular communication")
)

ggplot(boot_beta, aes(mean, parameter)) +
  geom_vline(xintercept = 0, colour = "grey65", linetype = "dashed") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper),
                 height = 0.16, colour = "steelblue") +
  geom_point(size = 2.4, colour = "steelblue") +
  labs(x = "Bootstrap mean and percentile interval",
       y = "Covariate coefficient") +
  theme_hv_manuscript()
```

The `mal` interval crosses zero in this small run, while the other displayed
intervals do not. That is a stability signal, not a final confidence interval.
For a study analysis, prespecify a substantially larger resample, review the
failure count, and report the resampling method with the model specification.

## Are competing events using the right denominator?

Competing-risk incidence is built from a shared risk set. At each event time,
`n_risk` is the number still free of every modeled event just before the events
at that time. A cause-specific cumulative-incidence curve adds the current
event-free survival multiplied by that cause's event proportion. The curves
therefore share one probability budget: event-free survival plus every cause's
cumulative incidence equals 1.

The public `omc` extract contains death/last-contact follow-up and indicators
that one or more thromboembolic events were recorded, but it does not contain
the thromboembolism date. We use the required three-state construction to show
the estimator mechanics. Do not interpret the resulting type 1 curve as a
clinical time-to-thromboembolism estimate. For that analysis, supply the time
to first thromboembolism or death and code the first observed event.

```{r}
#| label: competing-risk-object
data(omc, package = "TemporalHazard")

event <- with(
  omc,
  ifelse(
    te1 == 1 | te2 == 1 | te3 == 1,
    1L,
    ifelse(dead == 1, 2L, 0L)
  )
)
cr <- hzr_competing_risks(omc$int_dead, event)

data.frame(
  records = nrow(omc),
  censored = sum(event == 0),
  any_thromboembolism_recorded = sum(event == 1),
  death_without_recorded_thromboembolism = sum(event == 2),
  event_times = nrow(cr),
  final_probability_sum = tail(cr$surv + cr$incid_1 + cr$incid_2, 1)
)
head(as.data.frame(cr)[
  c("time", "n_risk", "n_event_1", "n_event_2", "n_censor",
    "surv", "incid_1", "incid_2")
])
```

```{r}
#| label: fig-temporal-diagnostics-competing-risks
#| fig-cap: "Estimator demonstration using OMC follow-up classification; the public extract does not contain the thromboembolism time required for clinical incidence"
cif <- rbind(
  data.frame(time = cr$time, incidence = cr$incid_1 * 100,
             cause = "Any thromboembolism recorded"),
  data.frame(time = cr$time, incidence = cr$incid_2 * 100,
             cause = "Death without recorded thromboembolism")
)

ggplot(cif, aes(time, incidence, colour = cause, linetype = cause)) +
  geom_step(linewidth = 0.75) +
  scale_colour_manual(
    values = c("Any thromboembolism recorded" = "steelblue",
               "Death without recorded thromboembolism" = "firebrick"),
    name = NULL
  ) +
  scale_linetype_manual(
    values = c("Any thromboembolism recorded" = "solid",
               "Death without recorded thromboembolism" = "dashed"),
    name = NULL
  ) +
  labs(x = "Months to death or last contact",
       y = "Cumulative incidence in constructed example (%)") +
  theme_hv_manuscript() +
  theme(legend.position = "top")
```

If the probability sum is not 1 to numerical precision, the event coding or
risk-set construction needs attention. In a real competing-risk analysis,
also verify that every row carries the time to the first event of any modeled
cause, not a later administrative time.
