# Themes for consistency

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

hvtiPlotR [@R-hvtiPlotR] ships output-target themes via `theme_hv_*()`. Each sets a complete
non-data formatting baseline; the constructor and `plot()` add none of it. Pick
the theme that matches where the figure lands.

| Theme | Target |
|-------|--------|
| `theme_hv_manuscript()` | Journal PDF, black-on-white |
| `theme_hv_poster()` | Conference poster, larger text |
| `theme_hv_ppt_light()` | PowerPoint on white/light background |
| `theme_hv_ppt_dark()` | PowerPoint on dark/navy background |

## When to use it

A theme is everything on a plot that is not the data: text sizes, line weights,
background, grid. Two figures of the same cohort can need very different
versions of that chrome depending on where they end up. A journal page is read
from a foot away in a quiet office; a poster is read from across a hall; a slide
is read for fifteen seconds in a dim room. The font and tick weights that work
in one fail in the others, and the right time to deal with that is by swapping a
single theme call rather than hand-tuning a dozen `theme()` elements per figure.

The choice follows the destination, not your taste. Use `theme_hv_manuscript()`
for anything headed to a journal: small text, white background, minimal chrome,
tuned for a figure reduced to a column width. Use `theme_hv_poster()` for foam
board and, often, for a projector in a meeting room, where the larger tick
labels survive the distance. Use `theme_hv_ppt_light()` for slides on a white or
light-grey background (the Cleveland Clinic template, most default Office decks)
and `theme_hv_ppt_dark()` for navy or dark-gradient decks where the convention
is white-on-dark. Match the theme to the background it will sit on and the
distance it will be read from, and the rest takes care of itself.

One behaviour is shared by all four and is easy to read as a bug the first time
you meet it. Each sets `legend.position = "none"`, so every figure below comes
back without a key even though `decorate()` gives both scales a name. That is the
house convention rather than an oversight: a CORR figure names its groups on the
panel, where the reader's eye already is, instead of sending it out to a margin
box and back once per series. @sec-annotation-series has that recipe, including
why the label takes the theme's ink rather than the colour of the series it
names. When a grouping genuinely cannot be named on the panel, the
[legends](legends.qmd) chapter covers putting a key back, and `hv_legend_inside()`
does it without surrendering panel width.

We render the same base trends plot under each theme, then show targeted
overrides. The shared decorations are factored out once.

```{r}
#| label: base
dta_trends <- sample_trends_data(n = 600, seed = 42)
p_base     <- plot(hv_trends(dta_trends))

decorate <- function(p) {
  p +
    scale_colour_brewer(palette = "Set1", name = "Group") +
    scale_shape_manual(
      values = c("Group I" = 15, "Group II" = 19,
                 "Group III" = 17, "Group IV" = 18),
      name = "Group"
    ) +
    labs(x = "Surgery Year", y = "Outcome")
}
```

## Manuscript

Default for journal submissions: white background, black text, minimal chrome.
Font sizes tuned for letter-size PDFs reduced to a column width.

```{r}
#| label: fig-themes
#| fig-cap: "Grouped trends plot under theme_hv_manuscript(), the journal default with white background and minimal chrome"
decorate(p_base) + theme_hv_manuscript()
```

## Poster

Larger axis text and tick weights, panel grid removed: reads at arm's length
on a foam board, and holds up on a projector.

```{r}
#| label: fig-themes-poster
#| fig-cap: "The same plot under theme_hv_poster(), with larger axis text and heavier tick weights for arm's-length reading"
decorate(p_base) + theme_hv_poster()
```

## Light PowerPoint

Dark text and lines for white or light-grey slide backgrounds: the Cleveland
Clinic standard template and most default Office decks.

```{r}
#| label: fig-themes-ppt-light
#| fig-cap: "The same plot under theme_hv_ppt_light(), with dark text and lines for white or light-grey slide backgrounds"
decorate(p_base) + theme_hv_ppt_light()
```

## Dark PowerPoint

White-on-dark for navy or dark-blue gradient decks. The theme sets a
transparent plot background so the slide shows through; the `plot.background`
override below simulates that navy in this rendered book and would be omitted
when saving to an actual `.pptx`.

```{r}
#| label: fig-themes-ppt-dark
#| fig-cap: "The same plot under theme_hv_ppt_dark(), with white-on-dark styling for navy or dark-gradient decks"
decorate(p_base) +
  theme_hv_ppt_dark() +
  theme(plot.background = element_rect(fill = "navy", colour = "navy"))
```

## Theme overrides

Each theme is a baseline. Layer additional `theme()` calls after it to adjust
individual elements without touching the rest. The same overrides can also be
passed directly through the theme's `...`, e.g.
`theme_hv_poster(legend.position = "bottom")`.

### Axis text size

Resize tick labels (`axis.text`) and axis titles (`axis.title`) independently,
useful when scaling a figure down for a two-column layout.

```{r}
#| label: axis_text
decorate(p_base) +
  theme_hv_poster() +
  theme(
    axis.text  = element_text(size = 10),
    axis.title = element_text(size = 12)
  )
```

### Removing minor grid lines

Minor grid lines add density without precision. Blank them while keeping the
major grid.

```{r}
#| label: grid
decorate(p_base) +
  theme_hv_poster() +
  theme(panel.grid.minor = element_blank())
```

### Rotating x-axis labels

For time-point or long category labels. `angle = 45` with `hjust = 1, vjust = 1`
keeps the rotated label right-aligned to its tick.

```{r}
#| label: rotate
decorate(p_base) +
  theme_hv_poster() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

### Plot title and subtitle

Titles are stripped from the base themes (rarely used in journal figures) but
can be added back via `labs()` and styled in `theme()`.

```{r}
#| label: title
dta_km <- sample_survival_data(n = 500, seed = 42)
km     <- hv_survival(dta_km)

plot(km) +
  scale_color_manual(values = c(All = "steelblue"), guide = "none") +
  scale_fill_manual(values  = c(All = "steelblue"), guide = "none") +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  coord_cartesian(xlim = c(0, 20), ylim = c(0, 100)) +
  labs(
    title    = "Overall Survival",
    subtitle = paste0("n = ", nrow(dta_km), " patients"),
    x        = "Years after Operation",
    y        = "Freedom from Death (%)"
  ) +
  theme_hv_poster() +
  theme(
    plot.title    = element_text(size = 14, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 11, colour = "grey40", hjust = 0)
  )
```

### Expanding plot margins

Add breathing room around the panel, for a figure placed directly on a poster,
or when labels clip a tight device edge. Arguments follow top / right / bottom /
left order.

```{r}
#| label: margin
decorate(p_base) +
  theme_hv_poster() +
  theme(plot.margin = margin(t = 10, r = 20, b = 10, l = 20, unit = "pt"))
```

## Pitfalls

- **Poster text on a manuscript figure.** The most common mismatch is reusing a
  poster-themed figure in a paper. The tick labels and line weights that read
  cleanly from across a hall look heavy and oversized on a journal page, and a
  reviewer notices. When the same analysis travels from a conference to a
  manuscript, re-render with `theme_hv_manuscript()` rather than shrinking the
  poster version, so the text scales correctly for its new size.
- **A vanished legend is the theme working, not failing.** All four
  `theme_hv_*()` functions set `legend.position = "none"`. If you add a house
  theme and your key disappears, nothing has broken; the convention is to name
  groups on the panel instead. Restore a key deliberately when the grouping
  cannot be named that way: pass `legend.position` through the theme's `...`, or
  apply `hv_legend_inside()` after the theme so its position wins.
- **Dark theme on a white slide.** `theme_hv_ppt_dark()` sets a transparent
  background and white text on the assumption that a dark slide sits behind it.
  Drop that figure onto a white slide and the white text vanishes. The reverse
  fails just as badly: a light-theme figure on a navy slide gives you a glaring
  white box in the middle of the deck. Pick the PowerPoint theme to match the
  slide background, and check the figure against the actual master rather than
  the white render in this book.
