# Venn diagrams and UpSet plots

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(hvtiPlotR)
library(ggplot2)
```

## When to use it

Reach for a set-membership figure whenever a patient can belong to more than one
group at once and you want to show how those groups overlap. The classic
cardiac-surgery case is concomitant procedure: a single operation might be an
aortic valve replacement, or a valve replacement *plus* a CABG, or all three of
a valve replacement, a CABG, and an aortic repair. The question is not "how many
had each procedure" (a bar chart answers that) but "which combinations actually
occur, and how often."

The Venn diagram is the familiar answer, and for two or three sets it is the
right one: overlapping circles, the eye reads the intersections directly. At
four sets, switch to an UpSet plot. The number of possible regions has already
outgrown what the eye can compare reliably in overlapping shapes.

That is where the UpSet plot takes over. Instead of drawing every region as a
patch of a circle, it lists each observed combination as a column and draws a
bar for how many patients fall in it, with a matrix of filled and empty dots
below showing which sets that column represents. It reads like a bar chart with
a key, and it scales cleanly to seven sets or more.

## Venn diagrams

For two or three overlapping groups, a Venn diagram is still the clearest
display: the reader sees the intersections without a legend. A fourth set is
the decision point: switch to the UpSet plot below, which is built to carry
larger intersections.

`hv_venn()`, from hvtiPlotR [@R-hvtiPlotR], builds one from the same kind of
set-membership data the UpSet plot uses: pass the data frame and the columns to
treat as sets, then `plot()`. We pick three procedures that can co-occur. A Venn
is coordinate-free, so we do **not** add a house theme. `plot.hv_venn()` styles
the diagram itself, and `theme_hv_manuscript()` would only paste spurious x/y
axes back onto it.

```{r}
#| label: venn_inspect
venn_dta <- sample_upset_data(n = 400, seed = 42)
venn <- hv_venn(
  venn_dta,
  sets = c("AV_Replacement", "MV_Replacement", "CABG")
)
venn
venn$tables$regions
```

### Build it

```{r}
#| label: venn
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Three-set Venn of co-occurring procedures, each region labelled with its patient count and percent"
plot(venn, set_name_size = 5)
```

Each region carries its patient count and percentage. The empty
replacement-on-replacement overlap (`0`) is the clinically expected finding.
Patients rarely have both valves replaced in one operation, while CABG overlaps
both. With four or more sets, use the UpSet plot.

## UpSet plots

`hv_upset()`, from hvtiPlotR [@R-hvtiPlotR], builds an UpSet diagram via
`ggupset::scale_x_upset()` to visualise surgical procedure co-occurrences or
any set-membership data. Use it for four or more sets, when the question is
which exact combinations occur and how large they are.

`plot.hv_upset()` returns a **ggplot** when `set_size = FALSE`, or a
**patchwork composite** of an intersection-bar plot plus a set-size sidebar
when `set_size = TRUE` (the default). For the bare ggplot path, themes apply
via `+`; for the patchwork path, use patchwork's `&` operator to theme every
sub-panel.

### The data it needs

`sample_upset_data()` returns a binary indicator matrix: one column per
procedure, one row per patient, with 1 indicating the procedure was performed.
Pass the column names to `intersect` to define the set-membership axes. Run
`colSums()` first to confirm the marginal counts; those totals are exactly what
the set-size sidebar reports, so seeing them now makes the finished figure easy
to sanity-check.

```{r}
#| label: upset_data
sets <- c("AV_Replacement", "AV_Repair", "MV_Replacement", "MV_Repair",
          "TV_Repair", "Aorta", "CABG")

upset_dta <- sample_upset_data(n = 400, seed = 42)
head(upset_dta)
colSums(upset_dta)

hu <- hv_upset(upset_dta, intersect = sets)
```

### Inspect it

Inspect the returned object before plotting. Its printed summary and
`set_counts` are the marginal totals; the table built from `.Procedures` is the
exact-intersection audit. These are different denominators. A patient with both
an aortic valve replacement and CABG contributes once to each marginal total,
but once to the single `AV_Replacement + CABG` intersection.

```{r}
#| label: upset_inspect
hu
hu$tables$set_counts

intersection_counts <- sort(
  table(vapply(hu$data$.Procedures, paste, collapse = " + ",
               FUN.VALUE = character(1))),
  decreasing = TRUE
)
head(intersection_counts, 10)
```

### Build it

The default plot shows intersection bars (the ten most frequent combinations)
with a set-size sidebar on the right. Because `set_size = TRUE` is the default,
`plot()` returns a patchwork composite, so theme it with `&` to reach both
sub-panels at once.

```{r}
#| label: fig-upset
#| fig-cap: "The ten most frequent procedure combinations as intersection bars, with a set-size sidebar of marginal totals"
#| fig-width: 9
#| fig-height: 5
plot(hu) &
  theme_hv_manuscript()
```

## Read it

An UpSet plot is read from the top down, then from left to right. Look for:

- **The bars, in order.** The columns are sorted by frequency, so the tallest
  bar on the left is the single most common combination of procedures in the
  cohort. The dot matrix beneath each bar tells you which combination: a filled
  dot means that set is present, an empty dot means it is absent, and a vertical
  line joins the filled dots in a multi-set combination.
- **Singletons versus combinations.** A bar over a single filled dot is patients
  who had only that one procedure. A bar over two or more joined dots is the
  concomitant cases. The balance between the two is often the point of the
  figure: are most aortic valve replacements isolated, or do they travel with a
  CABG?
- **The set-size sidebar.** The horizontal bars on the right are the marginal
  totals, the same numbers `colSums()` printed above. A set with a large margin
  but few tall intersection bars is one that spreads thinly across many rare
  combinations.
- **What is missing.** UpSet shows only the combinations that actually occur,
  and only the top ten by default. A combination you expected to see and do not
  is genuinely absent (or rare enough to fall below the cutoff), which is itself
  worth noticing.

## Variations

### Custom intersection bar colour

The intersection bars are a standard `geom_bar()`, so you can recolour them with
the `bar_fill` argument. Setting `set_size = FALSE` drops the sidebar and returns
just the intersection-bar ggplot, which you then finish with the usual `+`.

```{r}
#| label: fig-upset-fill
#| fig-cap: "The intersection bars alone, recoloured and with the set-size sidebar dropped"
#| fig-width: 9
#| fig-height: 5
plot(hu, bar_fill = "steelblue", set_size = FALSE) +
  ggplot2::labs(y = "Patients (n)") +
  theme_hv_manuscript()
```

### Colour bars by era

To split each intersection by a grouping column, pass `fill_col`. The bars are
then stacked by that column's levels, which lets you ask whether the mix of
procedure combinations shifted over time. Combine it with `scale_fill_manual()`
to assign the colours explicitly.

```{r}
#| label: fig-upset-era
#| fig-cap: "Each intersection bar split and stacked by era, showing whether the mix of procedure combinations shifted over time"
#| fig-width: 9
#| fig-height: 5
upset_dta$era <- ifelse(seq_len(nrow(upset_dta)) <= 200, "Early", "Recent")
hu_era <- hv_upset(upset_dta, intersect = sets)

plot(hu_era, fill_col = "era", set_size = FALSE) +
  ggplot2::scale_fill_manual(
    values = c("Early" = "grey60", "Recent" = "steelblue"),
    name   = "Era"
  ) +
  ggplot2::labs(y = "Patients (n)") +
  theme_hv_manuscript()
```

## Pitfalls

- **Theming the wrong way for the wrong object.** When `set_size = TRUE` (the
  default) the plot is a patchwork composite, and `+ theme_hv_manuscript()` styles
  only one sub-panel, leaving the sidebar mismatched. Use `&` for the composite
  and `+` only when you have set `set_size = FALSE`.
- **Reading the top-ten cutoff as the whole story.** The default shows the ten
  most frequent combinations. The long tail of rarer combinations is real and
  may carry the unusual cases you care about; if so, raise the cutoff rather than
  assuming the figure is exhaustive.
- **Over-stuffing the Venn.** Two or three sets belong in a Venn. Four or more
  belong in an UpSet plot, not another ellipse.
- **Confusing a margin with a combination.** The sidebar totals count every
  patient who had a procedure, regardless of what else they had; the intersection
  bars count exact combinations. A set can have a large margin and still appear in
  no tall intersection bar because its patients scatter across many combinations.
