# varPro partial dependence

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 5, dpi = 150)
library(varPro)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
library(randomForestSRC)
library(survival)
```

## When to use it

The varPro chapter ranks and selects variables; this one asks the next question
*how* a selected variable moves the prediction. `gg_partial_varpro()` runs
`varPro::partialpro()` on a fitted `varpro` object and tidies the result into
marginal-effect curves: one panel per variable, the predicted response across
that variable's range with the rest of the design held as observed. Reach for it
once selection has narrowed the field and you want to read the shape and
direction of each survivor's effect, not just its rank.

The y-axis is the part worth knowing in advance, because it depends on the
outcome family and you can choose it. A `scale` argument sets what the curve
plots: response units for regression, a probability or log-odds for
classification, and survival probability, restricted mean survival time, or
ensemble mortality for a survival fit. That choice is the difference between a
figure a clinician reads at a glance and one only the modeller understands.

## The data it needs

`gg_partial_varpro()` reads a fitted `varpro` object, so we build one of each
family up front, reusing the datasets from the random-forest chapters: `housing`
for regression, `breast` for classification, and `pbc` for survival.

```{r}
#| label: fits
set.seed(42)
data(housing, package = "randomForestSRC")
data(breast,  package = "randomForestSRC")
data(pbc,     package = "randomForestSRC")

o_reg  <- varpro(SalePrice ~ .,            data = housing, ntree = 100)
o_cls  <- varpro(status ~ .,               data = breast,  ntree = 100)
o_surv <- varpro(Surv(days, status) ~ .,   data = pbc,     ntree = 100)
c(regression = o_reg$family, classification = o_cls$family, survival = o_surv$family)
```

Across all three the call is the same; only the `scale` (and, for survival, an
optional `time`) changes. With wide data we pass `nvars` so `partialpro` only
computes the top-ranked handful — it is the expensive step, and on a survival
fit it is expensive enough to matter.

## Regression: the response scale

On a regression fit the y-axis is the response in its own units. Here that is
`SalePrice`, so each curve reads as dollars of sale price across the predictor.

```{r}
#| label: fig-varpro_partial
#| fig-cap: "Partial-dependence curves for the top six predictors of the housing regression fit, plotted on the response scale in dollars of sale price"
plot(gg_partial_varpro(object = o_reg, nvars = 6, method = "rnd")) & theme_hv_manuscript()
```

`plot()` returns a `patchwork` panel, so we theme it with `&`. A rising curve is
a variable that pushes the prediction up over its range; a flat one is a variable
the forest barely leans on once the others are fixed.

## Classification: probability or log-odds

A classification fit defaults to **probability** — `P(Y = the target class)`,
bounded 0 to 1 and the scale a clinician expects.

```{r}
#| label: fig-varpro_partial-prob
#| fig-cap: "Partial dependence for the breast classification fit on the probability scale, bounded zero to one"
plot(gg_partial_varpro(object = o_cls, nvars = 6, scale = "prob", method = "rnd")) &
  theme_hv_manuscript()
```

Setting `scale = "logodds"` plots the same effect on the log-odds scale, which is
linear in the model's own terms and is the only scale on which the `causal`
(virtual-twins) contrast is shown. Use probability to communicate and log-odds to
diagnose.

```{r}
#| label: fig-varpro_partial-logodds
#| fig-cap: "The same classification partial dependence on the log-odds scale, linear in the model's own terms and the only scale showing the causal contrast"
plot(gg_partial_varpro(object = o_cls, nvars = 6, scale = "logodds", method = "rnd")) &
  theme_hv_manuscript()
```

## Survival: probability, RMST, or mortality

A survival fit gives you three readings of the same partial effect, and the
choice is the whole point of the `scale` argument.

`scale = "surv"` (the default) plots **survival probability** `S(tau | x)`,
bounded 0 to 1 at a horizon `tau`. Leave `time` out and `tau` resolves to the
median follow-up, printed on the axis label (here `t = 1788`), so you get a
units-safe horizon without guessing.

```{r}
#| label: fig-varpro_partial-surv
#| fig-cap: "Partial dependence for the pbc survival fit on the survival-probability scale at the median follow-up horizon"
plot(gg_partial_varpro(object = o_surv, nvars = 4, scale = "surv", method = "rnd")) &
  theme_hv_manuscript()
```

`scale = "rmst"` plots **restricted mean survival time**: the expected event-free
time within the first `tau` days, read in the model's own time units and
higher-is-better. It answers "how much time does this variable buy?" rather than
"what is the probability at `tau`."

```{r}
#| label: fig-varpro_partial-rmst
#| fig-cap: "Partial dependence for the pbc survival fit on the restricted-mean-survival-time scale, read as event-free time within the horizon"
plot(gg_partial_varpro(object = o_surv, nvars = 4, scale = "rmst", method = "rnd")) &
  theme_hv_manuscript()
```

The third option, `scale = "mortality"`, is the unbounded ensemble-mortality
score — higher is worse, the opposite direction from the two bounded scales above.
It is the right choice only when you are comparing against other random-forest
mortality outputs; for a figure a reviewer reads, prefer `surv` or `rmst`.

## Pitfalls

- **`partialpro` is the expensive step.** It recomputes for every variable you
  ask for and, on a survival fit, for the whole time grid. Always cap it with
  `nvars`; under Quarto's `freeze` the result is cached after the first render.
- **The y-axis depends on `scale`, and the default depends on the family.**
  Regression is response units, classification defaults to probability, survival
  defaults to survival probability. State the scale in the caption so a reader
  knows whether higher is better (`surv`, `rmst`) or worse (`mortality`).
- **`tau` must be in the fit's time units.** For `surv`/`rmst`, a `time` beyond
  the largest event time truncates to the full restricted mean; omit `time` to
  let it default to the median follow-up.
- **The `causal` contrast only shows on `logodds`.** On the bounded probability
  and survival scales it is hidden by design, so do not expect the virtual-twins
  estimate there.
- **Set `method` to `"rnd"` on every `gg_partial_varpro()` call.** Its `...`
  forwards straight to `varPro::partialpro()`, which grows its own isolation
  forest for the partial-dependence draws; leave `method` unset and that forest
  defaults to `unsupv`, handing `randomForestSRC` a zero-length `yvar.wt`
  pointer. Nothing breaks in practice — the pointer is formed, never
  dereferenced — but the real fix belongs upstream. Set it anyway: every call
  in this chapter does, because these are the chunks readers copy.
