# Variable priority with varPro

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 150)
library(varPro)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
```

## When to use it

Permutation importance (the VIMP chapter) answers "how much does the forest lean
on this variable?" by breaking the variable and watching error rise. `varPro`
[@R-varPro] asks a different question: "how much do the forest's own splitting
rules *depend* on this variable?" Rather than permuting predictors, it scores
each one by how much a tree's release rules rely on it, and calls the result
variable *priority*. Reach for it when you want a selection that comes with a
built-in cutoff (so the method tells you which variables to keep, not just how to
rank them), when you want the direction and magnitude of a local effect, or when
you want to flag unusual observations the same forest can isolate.

So varPro sits alongside VIMP as a statement about the model, while
[SHAP](rf_shap.qmd) speaks about a patient and
[dependence](rf_dependence.qmd) speaks about a predictor. The
[SHAP chapter](rf_shap.qmd#when-to-use-it) lays those four views out side by
side if you are deciding which one a figure needs.

`ggRandomForests` [@R-ggRandomForests] provides a family of `gg_*` extractors
that pull these scores out of a fitted `varpro` object and render them as tidy
ggplot2 graphics. This chapter works through the main extractors on a regression
fit and a classification fit. As elsewhere, each `plot()` hands back a bare
ggplot you finish with a house theme.

## The data it needs

The extractors read from a fitted `varpro` object, so the input is whatever
`varpro()` itself needs: a formula and a data frame. The outcome type decides
the family, and that matters here because several of the extractors are
family-specific. We start with a regression fit on the `housing` data shipped
with `randomForestSRC` [@R-randomForestSRC]: 2,930 homes and 80 predictors,
modelling sale price (`SalePrice`). This is the same kind of dataset the
random-forest chapters use, and the extra width is the point: variable priority
earns its keep when there are far more candidates than you can eyeball.

```{r}
#| label: fig-varpro
#| fig-cap: "Variable priority for the top 20 predictors of the housing regression fit, with the dashed cutoff line separating selected from unselected variables"
#| fig-height: 5.5
set.seed(42)
data(housing, package = "randomForestSRC")
o <- varpro(SalePrice ~ ., data = housing, ntree = 100)
plot(gg_varpro(o, nvar = 20)) + theme_hv_manuscript()
```

`gg_varpro()` returns an honest 15/85 boxplot for each variable. With 80
predictors we pass `nvar = 20` to show the top-ranked; the box spans the
importance values obtained when the variable is released across the rule
ensemble, and the dashed reference line at the `cutoff` (default 0.79 of the top
score on the local-standardized scale) separates selected from unselected
variables. Variables whose boxes sit to the right of the cutoff are retained.

## Build it

The plot in @fig-varpro is the core figure. Two variants make it more informative.
Setting `faithful = TRUE` shows the per-tree importance draws as jittered points
behind the boxplot, exposing the spread that the summary hides:

```{r}
#| label: fig-varpro-faithful
#| fig-cap: "The same priority boxplots with per-tree importance draws shown as jittered points behind each box, exposing the spread the summary hides"
#| fig-height: 5.5
plot(gg_varpro(o, faithful = TRUE, nvar = 20)) + theme_hv_manuscript()
```

Setting `nvar` lower restricts the display further, handy when a wide problem
like this one would otherwise crowd the panel and you only want the leaders:

```{r}
#| label: fig-varpro-nvar
#| fig-cap: "Variable priority restricted to the top eight predictors, the leaders-only view for a crowded wide problem"
plot(gg_varpro(o, nvar = 8)) + theme_hv_manuscript()
```

## Read it

The 15/85 boxplot is the figure to learn to read, and the cutoff line is what
makes it a *selection* rather than just a ranking.

- **The cutoff line decides membership.** Boxes to the right of the dashed line
  are the variables varPro keeps; boxes to the left are set aside. Unlike a VIMP
  bar chart, where you eyeball the break yourself, the method draws the line for
  you. You can move it with the `cutoff` argument when you want to be stricter or
  looser.
- **Box width is stability.** A tight box means the variable scored consistently
  across the rule ensemble; a wide box straddling the cutoff is a variable whose
  selection you should not bet the paper on. Turn on `faithful = TRUE` to see the
  raw draws behind the summary.
- **Priority is not the same as importance.** A variable can carry high priority
  (the rules depend on it) without the large error jump that permutation VIMP
  rewards, and vice versa. Read the two methods as complementary views, not as
  two attempts at the same number.

## Variations

### Beta-refined importance

`gg_beta_varpro()` is regression-only. Within each release region it fits a
lasso and reports the refined slope (β) coefficients instead of a VIMP-style
importance, so you get not just *whether* a variable matters but the direction
and magnitude of its local linear effect.

```{r}
#| label: fig-varpro-beta
#| fig-cap: "Beta-refined importance for the housing regression fit, reporting lasso slope coefficients that carry direction and magnitude rather than VIMP-style scores"
#| fig-height: 6
plot(gg_beta_varpro(o)) + theme_hv_manuscript()
```

The per-rule lasso refinement (`beta.varpro`) is the expensive step; under
Quarto's `freeze` the result is cached after the first render. Because these are
lasso-refined β coefficients rather than importance scores, they can be negative.

`gg_beta_uvarpro()` is the same idea for a fit with no outcome at all.
`varPro::uvarpro()` grows the forest without a response (the unsupervised
section later in this chapter has the full treatment), and the lasso inside each
region asks how well the *other* variables explain the released one. Average
those absolute coefficients and you get one importance per variable, on the same
lasso scale but answering a question about the design rather than about
`SalePrice`. We fit it on the numeric `housing` predictors and reuse that fit for
the dependency graph in @fig-varpro-udependent.

```{r}
#| label: fig-varpro-beta-uvarpro
#| fig-cap: "Beta-refined importance for the unsupervised housing fit, the outcome-free counterpart to the regression beta plot above"
#| fig-height: 6
set.seed(42)
num_cols <- names(housing)[vapply(housing, is.numeric, logical(1))]
uv <- uvarpro(housing[, setdiff(num_cols, "SalePrice")], ntree = 100)
plot(gg_beta_uvarpro(uv)) + theme_hv_manuscript()
```

The two are told apart by the object you hand them. `gg_beta_varpro()` needs a
regression `varpro` fit and reports a slope on the outcome; `gg_beta_uvarpro()`
needs a `uvarpro` fit and reports how much the rest of the design explains each
variable. Bars at or above the dashed cutoff are flagged as selected, and that
cutoff sits at `mean(beta_mean)` unless you set `cutoff` yourself.

### Individual variable priority

`gg_ivarpro()` computes *individual* (observation-level) variable priority,
decomposing the forest importance down to each case. On a regression fit with
weak per-observation signal the decomposition can collapse to nothing (it
returns zero rows), so we demonstrate it on a classification forest, which also
lets us show the `which_class` argument. We fit `status ~ .` on the `breast`
data (also from `randomForestSRC`): 198 patients, 32 predictors, and a two-level
recurrence outcome (`N` = no recurrence, `R` = recurrence).

```{r}
#| label: fig-varpro-ivarpro
#| fig-cap: "Individual variable priority for the breast classification fit, decomposing forest importance down to each observation"
#| fig-height: 5.5
set.seed(42)
data(breast, package = "randomForestSRC")
oc <- varpro(status ~ ., data = breast, ntree = 100)
plot(gg_ivarpro(oc)) + theme_hv_manuscript()
```

Each panel summarises how strongly individual observations rely on a given
variable. You can focus on a single outcome level with `which_class`, here the
recurrence class:

```{r}
#| label: fig-varpro-ivarpro-class
#| fig-cap: "Individual variable priority focused on the recurrence class of the breast fit via the which_class argument"
#| fig-height: 5.5
plot(gg_ivarpro(oc, which_class = "R")) + theme_hv_manuscript()
```

### Conditional priority by class

The classification fit also supports `conditional = TRUE` in `gg_varpro()`,
which facets the priority bars by outcome class so you can read which predictors
separate which classes. This option needs a classification forest; it errors on
the regression fit.

```{r}
#| label: fig-varpro-conditional
#| fig-cap: "Conditional variable priority faceted by outcome class, showing which predictors separate which classes in the breast fit"
#| fig-height: 5.5
plot(gg_varpro(oc, conditional = TRUE, nvar = 15)) + theme_hv_manuscript()
```

### Anomaly / isolation scores

`gg_isopro()` visualises in-sample anomaly (isolation) scores. These come from
an `isopro` object, and there is no `.varpro` method that builds one for you: you
construct it first with `varPro::isopro()` on the fitted forest, then pass that
to `gg_isopro()`.

```{r}
#| label: fig-varpro-isopro
#| fig-cap: "In-sample anomaly scores from an isolation forest, where higher scores flag observations the forest separates with few splits"
iso <- isopro(object = o, ntree = 100)
plot(gg_isopro(iso)) + theme_hv_manuscript()
```

Higher isolation scores flag observations the forest separates with few splits,
candidate outliers. Pass `newdata` to `gg_isopro()` to score held-out
observations against the fitted isolation forest.

### Partial dependence

The extractors so far rank and refine variables; partial dependence asks *how* a
selected variable moves the prediction. `gg_partial_varpro()` runs
`varPro::partialpro()` on the fitted forest and draws a marginal-effect curve per
variable, with a `scale` argument that sets the y-axis to suit the outcome
family. Because that treatment spans regression, classification, and survival, it
gets its own chapter; see [varPro partial dependence](varpro_partial.qmd).
(`gg_partialpro()` is the superseded name for this extractor; it is now a thin
alias for `gg_partial_varpro()`, so reach for the new name.)

## Unsupervised variable dependence

The beta plot in @fig-varpro-beta-uvarpro was a first taste of an *unsupervised*
fit. `varPro::uvarpro()` grows a forest with no response to predict, then scores
how strongly each variable depends on the others. `gg_udependent()` reads those
cross-variable scores and draws them as a dependency graph, where nodes are
variables, edges the dependencies between them, so you can see which predictors
travel together before any modelling. Reach for it when you want to understand
the structure among the predictors themselves, for example to spot a cluster of
collinear measurements.

```{r}
#| label: fig-varpro-udependent
#| fig-cap: "Unsupervised dependency graph among the numeric housing predictors, where thick edges mark strongly interdependent variables"
#| fig-height: 5.5
plot(gg_udependent(uv)) + theme_hv_manuscript()
```

This is the same `uv` fit, built on the numeric `housing` predictors with the
`SalePrice` outcome and the factor columns dropped, because the dependency graph
comes from a numeric distance and cannot square a mixed-type design. A thick edge
between two variables means the forest finds them strongly interdependent; an
isolated node is a variable that carries information the others do not.

### Signal-variable detection

The graph shows you the structure; `gg_sdependent()` gives you the ranked table
behind it. Both read the same lasso-coefficient matrix, but where
`gg_udependent()` draws nodes and edges, `gg_sdependent()` returns one row per
candidate variable carrying a signal score, that variable's degree in the graph,
and a flag for whether it landed in `sdependent()`'s detected signal set. Reach
for it when you need to *name* the variables rather than point at a picture.

```{r}
#| label: fig-varpro-sdependent
#| fig-cap: "Signal-variable detection for the same unsupervised housing fit, ranking each variable by its dependency signal score with the detected signal set highlighted"
#| fig-height: 5.5
plot(gg_sdependent(uv)) + theme_hv_manuscript()
```

Both `gg_udependent()` and `gg_sdependent()` take a `labels` argument, the same
named character vector used in the [dependence chapter](rf_dependence.qmd), so a
graph built on raw column names can be relabelled for a manuscript without
renaming anything in the data.

Highlighted variables are the detected signal set, and the caption reports how
many of them cleared the threshold. Four arguments carry over from
`gg_udependent()`: `threshold`, `q.signal`, `directed`, and `min.degree`.
`threshold` (default 0.25) only builds the dependency graph, deciding which
entries of the lasso-coefficient matrix count as an edge, so raising it shrinks
the signal set solely by isolating nodes, which then get dropped before scoring
ever runs. The argument that actually sizes the signal set is `q.signal` (default
0.75): a variable is flagged as signal when its degree in that graph clears
`min.degree` (`NULL` by default, which falls back to 1 for a directed graph or 2
for an undirected one) *and* its score clears the `q.signal` quantile of every
score in the fit. Raise `q.signal` for a shorter, more conservative list; lower it
for a longer one. `directed` (default `TRUE`) keeps the two directions of a pair
apart, because "j explains i" and "i explains j" are separate lasso coefficients
and need not agree; set it `FALSE` when you only want to know that two variables
are dependent, not which way the dependency reads. The `degree` column rides
along in the returned data frame even though the plot does not draw it, so a
variable with a middling score but a high degree is worth a second look.

## Pitfalls

- **`gg_isopro()` needs an `isopro` object first.** There is no `.varpro` method,
  so calling `gg_isopro()` on the `varpro` object directly fails. Build
  `iso <- isopro(object = o, ...)` and then plot `iso`.
- **`gg_beta_varpro()` is regression-only.** The lasso β refinement is defined
  for a regression fit; do not expect it off a classification `varpro` object.
- **`conditional = TRUE` needs a classification forest.** `gg_varpro(conditional
  = TRUE)` facets priority by class, so it errors on a regression fit like the
  `housing` model. Use it on a classification fit such as the `breast` one.
- **`gg_ivarpro()` needs real signal.** Individual priority decomposes importance
  to each observation, and on a weak problem that decomposition can return zero
  rows and an empty plot. It works on a strong-signal fit like the `breast`
  classification forest, which is why the individual-priority examples use it.
- **`gg_udependent()` needs all-numeric data.** The unsupervised dependency graph
  is built from a numeric distance, so a mixed-type design (factors plus
  numerics, like the full `housing` frame) fails with "Adjacency matrices must be
  square." Pass only the numeric columns, as the example does.
- **Wide data needs `nvar`/`nvars`.** With 80 predictors the bare `gg_varpro()`
  and `gg_partial_varpro()` displays are unreadable walls; cap them with `nvar`
  (the boxplot) and `nvars` (which variables `partialpro` even computes, the
  expensive step).
- **`o$xvar.names` is not a list of columns you can subset with.** A `varpro`
  fit reports the features its rules actually use, and it one-hot encodes every
  factor it keeps, so a `Neighborhood` column comes back as
  `NeighborhoodNoRidge`, `NeighborhoodNridgHt`, and `NeighborhoodStoneBr`. Hand
  those to `housing[, ...]` and you get an undefined-columns error, which is a
  confusing way to find out that the name you asked for was never a column.
  `varpro_feature_names()` maps them back.

For the housing fit above, `o$xvar.names` runs 30 feature names against the
81-column frame we handed `varpro()`, 14 of them encoded level names, and
`varpro_feature_names()` recovers the 26 original columns:

```{r}
#| label: varpro-feature-names
ncol(housing)                                # what you handed it
length(o$xvar.names)                         # what the fit reports
setdiff(o$xvar.names, names(housing))        # encoded level names
varpro_feature_names(o$xvar.names, housing)  # mapped back to real columns
```

That gap between what a fit reports and what your data frame actually holds is
the same seam that runs through every extractor in this chapter: the score, the
graph, and the ranked table all read straight from the fitted object, and
`varpro_feature_names()` is the one step that translates its column names back
into ones you can hand to `housing[, ...]`.
