# varPro partial dependence

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 5, dpi = 150)
library(varPro)
library(ggRandomForests)
library(ggplot2)
library(hvtiPlotR)
library(randomForestSRC)
library(survival)
```

## When to use it

The varPro chapter ranks and selects variables; this one asks the next question
*how* a selected variable moves the prediction. `gg_partial_varpro()` runs
`varPro::partialpro()` on a fitted `varpro` object and tidies the result into
marginal-effect curves: one panel per variable, the predicted response across
that variable's range with the rest of the design held as observed. Reach for it
once selection has narrowed the field and you want to read the shape and
direction of each survivor's effect, not just its rank.

The y-axis is the part worth knowing in advance, because it depends on the
outcome family and you can choose it. A `scale` argument sets what the curve
plots: response units for regression, a probability or log-odds for
classification, and survival probability, restricted mean survival time, or
ensemble mortality for a survival fit. That choice is the difference between a
figure a clinician reads at a glance and one only the modeller understands.

The default is `scale = "auto"`, which reads the family off the fit: a
classification forest resolves to probability, a survival forest to survival
probability, and a regression forest to a generic scale. Generic still plots the
response in its own units, but it labels the axis "Partial Effect" rather than
naming the response, so a regression figure headed for a manuscript needs its
own `labs(y = ...)`. Auto needs the fitted `object` to read a family at all.
Hand it precomputed partial-dependence data and no fit and it warns, then falls
back to that same generic axis.

## The data it needs

`gg_partial_varpro()` reads a fitted `varpro` object, so we build one of each
family up front, reusing the datasets from the random-forest chapters: `housing`
for regression, `breast` for classification, and `pbc` for survival.

```{r}
#| label: fits
set.seed(42)
data(housing, package = "randomForestSRC")
data(breast,  package = "randomForestSRC")
data(pbc,     package = "randomForestSRC")

o_reg  <- varpro(SalePrice ~ .,            data = housing, ntree = 100)
o_cls  <- varpro(status ~ .,               data = breast,  ntree = 100)
o_surv <- varpro(Surv(days, status) ~ .,   data = pbc,     ntree = 100)
c(regression = o_reg$family, classification = o_cls$family, survival = o_surv$family)
```

Across all three the call is the same; only the `scale` (and, for survival, an
optional `time`) changes. With wide data we pass `nvars` so `partialpro` only
computes the top-ranked handful, because it is the expensive step, and on a
survival fit it is expensive enough to matter.

`nvars` selects by importance, not by list position. The variables are reordered
by their rank in `varPro::get.topvars()` first, and only then does `nvars` keep
the top few, so `nvars = 6` returns the six varPro ranked highest rather than the
six that happened to come back first. The panels arrive sorted the same way, by
importance rather than alphabetically, so the top-left panel is the strongest
variable and you can read the grid in rank order.

## Regression: the response scale

On a regression fit the y-axis is the response in its own units. Here that is
`SalePrice`, so each curve reads as dollars of sale price across the predictor.

```{r}
#| label: fig-varpro_partial
#| fig-cap: "Partial-dependence curves for the top six predictors of the housing regression fit, plotted on the response scale in dollars of sale price"
plot(gg_partial_varpro(object = o_reg, nvars = 6, method = "rnd")) & theme_hv_manuscript()
```

`plot()` returns a `patchwork` panel, so we theme it with `&`. A rising curve is
a variable that pushes the prediction up over its range; a flat one is a variable
the forest barely leans on once the others are fixed. The panels arrive in varPro
importance order, so `Gr.Liv.Area` leading the grid is a statement about rank and
not about the alphabet.

Note the y-axis label. The values are dollars, but a regression fit resolves to
the generic scale and prints "Partial Effect", which is the one place this
chapter's figures need a manual `labs(y = "Sale price (USD)")` before they go
into a manuscript.

## Classification: probability or log-odds

A classification fit defaults to **probability**, `P(Y = the target class)`,
bounded 0 to 1 and the scale a clinician expects.

```{r}
#| label: fig-varpro_partial-prob
#| fig-cap: "Partial dependence for the breast classification fit on the probability scale, bounded zero to one"
plot(gg_partial_varpro(object = o_cls, nvars = 6, scale = "prob", method = "rnd")) &
  theme_hv_manuscript()
```

Setting `scale = "logodds"` plots the same effect on the log-odds scale, which is
linear in the model's own terms and is the only scale on which the `causal`
(virtual-twins) contrast is shown. Use probability to communicate and log-odds to
diagnose.

```{r}
#| label: fig-varpro_partial-logodds
#| fig-cap: "The same classification partial dependence on the log-odds scale, linear in the model's own terms and the only scale showing the causal contrast"
plot(gg_partial_varpro(object = o_cls, nvars = 6, scale = "logodds", method = "rnd")) &
  theme_hv_manuscript()
```

## Survival: probability, RMST, or mortality

A survival fit gives you three readings of the same partial effect, and the
choice is the whole point of the `scale` argument.

`scale = "surv"` (the default) plots **survival probability** `S(tau | x)`,
bounded 0 to 1 at a horizon `tau`. Leave `time` out and `tau` resolves to the
median follow-up, printed on the axis label (here `t = 1788`), so you get a
units-safe horizon without guessing.

```{r}
#| label: fig-varpro_partial-surv
#| fig-cap: "Partial dependence for the pbc survival fit on the survival-probability scale at the median follow-up horizon"
plot(gg_partial_varpro(object = o_surv, nvars = 4, scale = "surv", method = "rnd")) &
  theme_hv_manuscript()
```

`scale = "rmst"` plots **restricted mean survival time**: the expected event-free
time within the first `tau` days, read in the model's own time units and
higher-is-better. It answers "how much time does this variable buy?" rather than
"what is the probability at `tau`."

```{r}
#| label: fig-varpro_partial-rmst
#| fig-cap: "Partial dependence for the pbc survival fit on the restricted-mean-survival-time scale, read as event-free time within the horizon"
plot(gg_partial_varpro(object = o_surv, nvars = 4, scale = "rmst", method = "rnd")) &
  theme_hv_manuscript()
```

The third option, `scale = "mortality"`, is the unbounded ensemble-mortality
score, where higher is worse, the opposite direction from the two bounded scales above.
It is the right choice only when you are comparing against other random-forest
mortality outputs; for a figure a reviewer reads, prefer `surv` or `rmst`.

## Pitfalls

- **`partialpro` is the expensive step.** It recomputes for every variable you
  ask for and, on a survival fit, for the whole time grid. Always cap it with
  `nvars`; under Quarto's `freeze` the result is cached after the first render.
- **The y-axis depends on `scale`, and the default reads the family off the
  fit.** `scale = "auto"` gives probability for classification and survival
  probability for survival, both of which name themselves on the axis. A
  regression fit resolves to the generic scale instead: the values are still
  response units, but the axis says "Partial Effect", so add your own
  `labs(y = ...)` before anyone reads it as unitless. Without a fitted `object`
  auto cannot resolve at all; it warns and uses the same generic axis. State the
  scale in the caption either way, so a reader knows whether higher is better
  (`surv`, `rmst`) or worse (`mortality`).
- **`nvars` ranks before it slices.** It keeps the top n by varPro importance,
  not the first n in the list it was handed, and the returned panels sort by
  importance rather than alphabetically. Raising `nvars` therefore extends the
  grid downward in rank; it does not reshuffle the panels you already had.
- **`tau` must be in the fit's time units.** For `surv`/`rmst`, a `time` beyond
  the largest event time truncates to the full restricted mean; omit `time` to
  let it default to the median follow-up.
- **The `causal` contrast only shows on `logodds`.** On the bounded probability
  and survival scales it is hidden by design, so do not expect the virtual-twins
  estimate there.
- **Set `method` to `"rnd"` on every `gg_partial_varpro()` call.** Its `...`
  forwards straight to `varPro::partialpro()`, which grows its own isolation
  forest for the partial-dependence draws; leave `method` unset and that forest
  defaults to `unsupv`, handing `randomForestSRC` a zero-length `yvar.wt`
  pointer. Nothing breaks in practice, since the pointer is formed and never
  dereferenced, but the real fix belongs upstream. Set it anyway: every call
  in this chapter does, because these are the chunks readers copy.
